# (PART) Foundations {-}

# Why causal inference? {#why}

## Questions epidemiology cannot answer with association alone

Epidemiologic studies are increasingly used to investigate the safety
and effectiveness of medical products and procedures as they are used in
routine care, in unselected patient populations. This book builds a
narrative around variations on two such questions: 1) Does starting a
Cox-2 selective NSAID result in a lower risk of peptic ulcer disease
than starting a nonselective NSAID? and 2) Does starting a statin reduce
the risk of cardiovascular events compared with no treatment in a broad
routine-care population? Both are causal questions, in that they ask
what would happen to a group of patients if a particular treatment
decision changed, and both are the kind of question epidemiologists are
asked to answer using data collected for other purposes, such as
electronic health records, administrative claims, and disease
registries. Because such data were not collected as part of a designed
study, many variables the researcher might wish to have remain
unrecorded, and the meaning of the variables that are available is often
unclear.

One of the principal problems of such studies is confounding — a
systematic difference between the group of patients exposed to an
intervention and a chosen comparator group. Comparing patients who took
a treatment with patients who did not yields an estimate of an
association between treatment and outcome. It answers a question of the
form "How did the risk of peptic ulcer disease differ between patients
who initiated a Cox-2 selective versus a nonselective NSAID?" — a
question about a difference in outcomes between two groups of patients
as they actually existed, not a causal question. Confounding arises when
the factors that influence physician treatment decisions and patient
medication use are also independent determinants of health outcomes.

Those factors are rarely simple. The use of a medical intervention is
determined by system-, physician-, and patient-level variables that
often interact in complex and poorly understood ways. A physician's
treatment decision may be based on an evaluation of the patient's health
status and prognosis, the physician's past experience with the
medication, or an assessment of the patient's ability and willingness to
take the medication as prescribed. Patients may initiate and remain
adherent to a new therapy because of their health beliefs, their
preferences about treatment, their trust in the physician, the
out-of-pocket cost of the medication, or their perceptions of their
disease risk and the benefits of treatment. Initiation and adherence may
also depend on a patient's physical and cognitive abilities. Each of
these variables may directly affect health outcomes or be related to
them through indirect pathways. From this process, several recognizable
patterns of bias result; they recur often enough in pharmacoepidemiology
to have earned names.

A common and often intractable form is **confounding by indication**,
sometimes called channeling. It arises from good medical practice:
physicians tend to prescribe medications to, and perform procedures on,
the patients who are most likely to benefit. Because it is often
difficult to measure medical indications and underlying disease severity
and prognosis, confounding by indication often makes medications appear
to cause the outcomes they are meant to prevent. Statins, lipid-lowering
drugs that reduce the risk of cardiovascular events, tend to be
prescribed to patients perceived to be at increased cardiovascular risk,
and incomplete control of those risk factors can make statins appear to
cause rather than prevent cardiovascular events. The same logic runs
through the first of our two questions: Cox-2 selective NSAIDs were
developed, and are often chosen, specifically to reduce gastrointestinal
risk in patients thought to need that protection, so Cox-2 initiators
can appear to fare worse with respect to peptic ulcer disease not
because the drug fails to protect them but because they were a
higher-risk group at baseline. If the magnitude of that confounding is
large enough, the apparent effect of a genuinely protective drug can be
reversed.

A second pattern is the **healthy user effect**. Patients who initiate a
preventive medication may be more likely than other patients to engage
in other healthy, prevention-oriented behaviors: to seek out preventive
healthcare services, exercise regularly, moderate their alcohol
consumption, and avoid unsafe and unhealthy activities. Incomplete
adjustment for such behaviors can make use of a preventive medication
spuriously associated with reduced risk of a wide range of adverse
health outcomes. Patients who adhere to a treatment once started may
likewise be more likely to engage in other healthy behaviors — the
closely related **healthy adherer effect**. Strong evidence for it comes
from a meta-analysis of randomized trials in which adherence to
*placebo* was associated with reduced mortality. This is clearly not an
effect of the placebo; it reflects the characteristics of patients who
take a medication as prescribed.

Related phenomena operate at the other end of the health spectrum.
Patients who are functionally impaired — who have difficulty performing
daily activities of living — may be less able to visit a physician or
pharmacy, and therefore less likely to fill prescriptions and receive
preventive healthcare services, a difference that can exaggerate the
apparent benefit of prescription medications, vaccines, and screening
tests. Similarly, patients who are perceived by a physician to be close
to death may be less likely to be prescribed preventive medications, and
patients may themselves discontinue preventive medications when their
health deteriorates. These mechanisms arise through different processes
but have the same effect: decreased exposure to preventive medications
among the patients at highest risk of death, which can produce a
substantially decreased mortality risk among users of statins and other
preventive drugs compared with apparently similar nonusers.

A final pattern worth naming here, because it shapes how this book's
cohorts are constructed, is **prevalent-user bias**: comparing people
already on a treatment to people never treated mixes in everyone who
started the drug, tolerated it, and stayed on it long enough to be
captured in the data, while excluding the patients who stopped early
because of side effects or lack of benefit. The comparisons in this
book are built around new users — patients captured at the moment they
start a drug, as `ns` and `sta` are — specifically to avoid folding that
kind of survivorship into the treatment groups before the analysis
begins.

These concerns are not hypothetical. Observational studies of
postmenopausal hormone therapy in the 1990s reported that use was
associated with substantially lower risk of coronary heart disease;
when the Women's Health Initiative randomized women to hormone therapy
or placebo, the trial found an increased risk. The discrepancy did not
arise from careless analysis. Rather, women who used hormone therapy
were healthier, more health-conscious, and more closely monitored than
women who did not — the healthy user effect operating at scale — and
that difference was not fully removed by adjustment for the variables
the studies had measured.

If a body of careful nonexperimental research can get the sign of an
effect wrong, then a large or richly detailed data set is not by itself
sufficient. Indeed, the property of being a confounder is a question of
causality, not of statistics: it is not identifiable, or testable, in
the data themselves, so no variable selection procedure or statistical
model can by itself guarantee an unbiased estimate, even in a very large
study. What is needed is a framework, grounded in subject-matter
knowledge, that says when an association can be given a causal
interpretation and what to do when it cannot. The rest of the book
describes such a framework and how the analysis can be implemented in R.

## A first look at the data

```{r setup-ch01, include=FALSE}
source("R/setup.R")
```

We will spend the whole book with the two cohorts introduced properly in
the next chapter, although it is worth looking at the simplest possible
comparison in one of them now, before any adjustment, to see what an
unadjusted answer to a causal question looks like.
In `ns`, the cohort of new NSAID users, patients either initiated a
nonselective NSAID or a Cox-2 selective inhibitor, and the outcome of
interest is incident peptic ulcer disease (PUD). The crude risk of PUD in
each group follows from a single `group_by()` and `summarise()`:

```{r crude-risk}
risk_tab <- ns |>
  group_by(cox2_initiation) |>
  summarise(n = n(), risk = mean(pud))

risk_tab |>
  knitr::kable(
    col.names = c("Initiated drug", "N", "Risk of incident PUD"),
    digits = c(0, 0, 3)
  )
```

Among the `r risk_tab$n[risk_tab$cox2_initiation == "No"]` patients who
initiated a nonselective NSAID,
`r sprintf("%.1f%%", 100 * risk_tab$risk[risk_tab$cox2_initiation == "No"])`
went on to develop incident PUD; among the
`r risk_tab$n[risk_tab$cox2_initiation == "Yes"]` patients who initiated a
Cox-2 selective drug, the figure was
`r sprintf("%.1f%%", 100 * risk_tab$risk[risk_tab$cox2_initiation == "Yes"])`.

Is that comparison an estimate of the causal effect of Cox-2 selectivity
on peptic ulcer risk? It is not, and there is reason to think it is far
from one. Nothing in the code above accounts for why a given patient
started one drug rather than the other. This is exactly the setting for
the confounding by indication described above: a patient with a history
of arthritis, prior GI bleeding, or concurrent corticosteroid or
aspirin use is more likely to be steered toward the Cox-2 selective
option. If Cox-2 initiators are systematically older or sicker in
ways connected to ulcer risk, as we will confirm with a proper Table 1 in
the next chapter, then the two rows of this table are not comparing like
with like, and the crude risk difference above combines whatever
causal effect the drug has with the baseline differences between the
groups. Separating those two components, that is, asking what the
comparison would look like if the two groups had been alike at baseline,
is the problem the rest of the book addresses.

## What this book covers

The book is organized in four parts. Part I, Foundations, which this
chapter opens, builds the conceptual vocabulary before any adjustment
method is introduced: potential outcomes, the randomized trial as the
benchmark against which a causal analysis is implicitly compared, causal
diagrams for reasoning about which variables require adjustment, and the
elements of study design that turn a causal question into an analyzable
data set. Part II, Core Methods, covers the two families of tools most
observational analyses reach for first: outcome regression and
standardization, and propensity scores, culminating in inverse
probability of treatment weighting. Part III, Time, Censoring, and
Survival, brings in what happens when the outcome does not occur at a
single fixed moment but can happen, or fail to happen, at any point over
follow-up: survival analysis, censoring, inverse probability of
censoring weights, and the clone-censor-weight approach to comparing
dynamic treatment strategies, all worked out on the statin cohort,
`sta`. Part IV, Beyond Exchangeability, deals with the harder case in
which adjustment for measured covariates is insufficient because of
unmeasured confounding, and introduces instrumental variables
and sensitivity analysis as ways of asking what an unmeasured confounder
would have to look like to explain away a finding.

Threaded through all four parts is a toolkit that you will build
yourself, one function at a time. By the end of the book you will have
written, tested, and reused `standardize()`
for outcome-model-based standardization, `iptw_estimator()` for
propensity-score weighting, `ipcw_estimator()` for censoring weights,
`ccw_estimator()` for the clone-censor-weight design, and `iv_estimator()`
for instrumental variable analysis. Each is introduced once, in the
chapter where its logic is developed, and then reused in a later
exercise on the other cohort, with a different set of covariates, or
with a different specification. Writing the functions yourself,
rather than only calling a package's implementation, is meant to make
the assumptions each method depends on explicit.

## How to read this book

The chapters are meant to be read with R running alongside them. Every
chapter is a self-contained R Markdown file that sources `R/setup.R` and
rebuilds what it needs from there, so you can open any chapter, run it
chunk by chunk, and confirm that the numbers in the text are the
numbers the code produces. The two risks in the table above, for
example, are computed from `ns` each time this chapter is rendered
rather than typed in. When a new function is introduced, read the code
closely enough to know what each argument does, then use it on data
slightly different from the example in the text before moving to the
exercises at the end of each chapter. The exercises themselves mostly do
not ask you to write new methods; they ask you to point a function you
already wrote at a different variable, a different cohort, or a
different specification, and to think about whether the answer it gives
still makes sense. That habit of applying a tool you understand to an
unfamiliar problem is closer to what causal inference looks like in
practice than anything else the book can offer.
