# (PART) Foundations {-}

# Why causal inference? {#why}

## Three kinds of questions

Epidemiologic analyses of healthcare data — electronic health records,
administrative claims, disease registries — can serve several distinct
scientific purposes, and it is worth separating them at the outset,
because the methods in this book apply to only one of them.

The first purpose is **description**. Descriptive questions ask what is
happening in a population: How many patients in North Carolina Medicaid
have asthma? Which counties have the highest rates? How many patients
will develop asthma this year? Answering such questions well requires
careful attention to how conditions are defined and to who is captured
in the data, but it does not require comparing groups of patients or
reasoning about what would happen under a different course of action.

The second purpose is **prediction**. Predictive questions include
classification problems — does this radiological image show a patient
with lung damage from asthma? — and risk prediction problems — which
patients are at highest risk of being admitted for an asthma
exacerbation? A model built for prediction is judged by how accurately
it predicts, and its inputs need no causal interpretation: a variable
can be an excellent predictor of an outcome without causing it. This is
why the common phrase "risk factors for admission" is ambiguous — it may
describe variables that merely flag high-risk patients, or variables
that, if changed, would change the risk.

The third purpose is **causal**. What is the effect of treatment with a
biologic, versus standard of care, on the one-year risk of admission for
an asthma exacerbation among newly diagnosed patients? Which patient
subgroups are most likely to benefit? What is the effect of reducing
copayments on adherence to guideline-recommended asthma treatments? What
is the effect of a spike in fine particulate air pollution on emergency
department visits for asthma? Each of these asks what would happen to a
population if something — a treatment, a policy, an exposure — were
different from what it was.

This book is about the third kind of question. It builds a narrative
around variations on two of them: 1) Does starting a Cox-2 selective
NSAID result in a lower risk of peptic ulcer disease than starting a
nonselective NSAID? and 2) Does starting a statin reduce the risk of
cardiovascular events compared with no treatment in a broad routine-care
population? Both ask what would happen to a group of patients if a
particular treatment decision changed, and both are the kind of question
epidemiologists are asked to answer using data collected for other
purposes.

## Why not just run a trial?

The natural first response to a causal question about a treatment is to
propose a randomized trial. Randomization ensures that the groups being
compared are, on average, alike at baseline, so a difference in their
outcomes can be attributed to the treatment; Chapter \@ref(rct) develops
this argument carefully, because the trial is the benchmark against
which every analysis in this book is implicitly judged. If a trial could
be run for every causal question in medicine, there would be little need
for the rest of this book. Several forces ensure that it cannot.

First, trials are expensive and slow. The ALLHAT trial randomized 33,357
patients with hypertension and at least one cardiovascular risk factor
to one of three classes of antihypertensive drugs and followed them for
between three and eight years. The trial established that thiazide
diuretics were as good as or superior to the newer, more expensive
alternatives for every outcome studied — a finding that still shapes
hypertension guidelines — but it took eight years to complete and cost
roughly $120 million. Only a small fraction of the clinically important
comparisons in medicine will ever command resources on that scale.

Second, it is often unclear who would pay. Head-to-head comparative
trials are commercially risky for manufacturers: in the PROVE-IT trial,
the sponsor's statin proved inferior to a competitor's more intensive
regimen, a result that reshaped practice at the sponsor's expense. A
manufacturer weighing whether to fund a comparison against a rival
product must accept the possibility of financing the demonstration that
its own drug is worse.

Third, the efficacy a trial estimates may differ from the effectiveness
a treatment achieves in routine care. Denosumab, an injectable biologic
for osteoporosis, reduced fracture risk in its pivotal placebo-controlled
trial; whether it prevents fractures in real-world use depends on
matters the trial protocol controlled — whether patients return for
their injections on schedule, and how the medication is offered and
covered by insurers. Nonexperimental studies can provide evidence about
real-world effectiveness that a trial, by design, cannot. The second of
this book's two questions — whether starting a statin reduces
cardiovascular risk in a broad routine-care population — is a question
of exactly this kind.

Fourth, trials often exclude the patients in whom a treatment will
actually be used: the elderly, patients with chronic comorbidities such
as end-stage renal disease or liver disease, racial and ethnic
minorities, patients with limited access to the healthcare system,
pregnant women, and children. For these groups, evidence about a
treatment's effects must frequently come from data generated by routine
care.

Finally, a trial may be unethical or impractical. Antipsychotic
medications, approved to treat schizophrenia, came to be widely used
off-label to manage behavioral disturbances in elderly patients with
dementia. Placebo-controlled trials of the newer atypical agents were
conducted in this population, and a meta-analysis of those trials by the
FDA found an increased risk of death among patients randomized to the
atypical drugs. The finding created a clinical dilemma: should
physicians switch patients to the older, conventional antipsychotics?
Those agents have many known side effects and a poor safety profile of
their own, and a head-to-head trial in frail elderly patients — with
death as the anticipated outcome — will never be done. The question
could only be answered by analyzing existing data. We will return to how
it was answered at the end of this chapter.

When a trial cannot be run, the causal question does not go away; it
falls to nonexperimental studies, and with them comes the central
problem this book is organized around.

## The central problem: confounding

A nonexperimental study must answer its question with data generated by
routine care — electronic health records, administrative claims, disease
registries. Because such data were not collected as part of a designed
study, many variables the researcher might wish to have remain
unrecorded, and the meaning of the variables that are available is often
unclear.

One of the principal problems of such studies is confounding — a
systematic difference between the group of patients exposed to an
intervention and a chosen comparator group. Comparing patients who took
a treatment with patients who did not yields an estimate of an
association between treatment and outcome. It answers a question of the
form "How did the risk of peptic ulcer disease differ between patients
who initiated a Cox-2 selective versus a nonselective NSAID?" — a
question about a difference in outcomes between two groups of patients
as they actually existed, not a causal question. Confounding arises when
the factors that influence physician treatment decisions and patient
medication use are also independent determinants of health outcomes.

Those factors are rarely simple. The use of a medical intervention is
determined by system-, physician-, and patient-level variables that
often interact in complex and poorly understood ways. A physician's
treatment decision may be based on an evaluation of the patient's health
status and prognosis, the physician's past experience with the
medication, or an assessment of the patient's ability and willingness to
take the medication as prescribed. Patients may initiate and remain
adherent to a new therapy because of their health beliefs, their
preferences about treatment, their trust in the physician, the
out-of-pocket cost of the medication, or their perceptions of their
disease risk and the benefits of treatment. Initiation and adherence may
also depend on a patient's physical and cognitive abilities. Each of
these variables may directly affect health outcomes or be related to
them through indirect pathways. From this process, several recognizable
patterns of bias result; they recur often enough in pharmacoepidemiology
to have earned names.

A common and often intractable form is **confounding by indication**,
sometimes called channeling. It arises from good medical practice:
physicians tend to prescribe medications to, and perform procedures on,
the patients who are most likely to benefit. Because it is often
difficult to measure medical indications and underlying disease severity
and prognosis, confounding by indication often makes medications appear
to cause the outcomes they are meant to prevent. Statins, lipid-lowering
drugs that reduce the risk of cardiovascular events, tend to be
prescribed to patients perceived to be at increased cardiovascular risk,
and incomplete control of those risk factors can make statins appear to
cause rather than prevent cardiovascular events. The same logic runs
through the first of our two questions: Cox-2 selective NSAIDs were
developed, and are often chosen, specifically to reduce gastrointestinal
risk in patients thought to need that protection, so Cox-2 initiators
can appear to fare worse with respect to peptic ulcer disease not
because the drug fails to protect them but because they were a
higher-risk group at baseline. If the magnitude of that confounding is
large enough, the apparent effect of a genuinely protective drug can be
reversed.

A second pattern is the **healthy user effect**. Patients who initiate a
preventive medication may be more likely than other patients to engage
in other healthy, prevention-oriented behaviors: to seek out preventive
healthcare services, exercise regularly, moderate their alcohol
consumption, and avoid unsafe and unhealthy activities. Incomplete
adjustment for such behaviors can make use of a preventive medication
spuriously associated with reduced risk of a wide range of adverse
health outcomes. Patients who adhere to a treatment once started may
likewise be more likely to engage in other healthy behaviors — the
closely related **healthy adherer effect**. Strong evidence for it comes
from a meta-analysis of randomized trials in which adherence to
*placebo* was associated with reduced mortality. This is clearly not an
effect of the placebo; it reflects the characteristics of patients who
take a medication as prescribed.

Related phenomena operate at the other end of the health spectrum.
Patients who are functionally impaired — who have difficulty performing
daily activities of living — may be less able to visit a physician or
pharmacy, and therefore less likely to fill prescriptions and receive
preventive healthcare services, a difference that can exaggerate the
apparent benefit of prescription medications, vaccines, and screening
tests. Similarly, patients who are perceived by a physician to be close
to death may be less likely to be prescribed preventive medications, and
patients may themselves discontinue preventive medications when their
health deteriorates. These mechanisms arise through different processes
but have the same effect: decreased exposure to preventive medications
among the patients at highest risk of death, which can produce a
substantially decreased mortality risk among users of statins and other
preventive drugs compared with apparently similar nonusers.

A final pattern worth naming here, because it shapes how this book's
cohorts are constructed, is **prevalent-user bias**: comparing people
already on a treatment to people never treated mixes in everyone who
started the drug, tolerated it, and stayed on it long enough to be
captured in the data, while excluding the patients who stopped early
because of side effects or lack of benefit. The comparisons in this
book are built around new users — patients captured at the moment they
start a drug, as `ns` and `sta` are — specifically to avoid folding that
kind of survivorship into the treatment groups before the analysis
begins.

These concerns are not hypothetical. Observational studies of
postmenopausal hormone therapy in the 1990s reported that use was
associated with substantially lower risk of coronary heart disease;
when the Women's Health Initiative randomized women to hormone therapy
or placebo, the trial found an increased risk. The discrepancy did not
arise from careless analysis. Rather, women who used hormone therapy
were healthier, more health-conscious, and more closely monitored than
women who did not — the healthy user effect operating at scale — and
that difference was not fully removed by adjustment for the variables
the studies had measured.

If a body of careful nonexperimental research can get the sign of an
effect wrong, then a large or richly detailed data set is not by itself
sufficient. Indeed, the property of being a confounder is a question of
causality, not of statistics: it is not identifiable, or testable, in
the data themselves, so no variable selection procedure or statistical
model can by itself guarantee an unbiased estimate, even in a very large
study. What is needed is a framework, grounded in subject-matter
knowledge, that says when an association can be given a causal
interpretation and what to do when it cannot. The rest of the book
describes such a framework and how the analysis can be implemented in R.

## Causal inference, old and new

Learning about causes from nonexperimental data is not a new ambition in
epidemiology. In 1854, John Snow traced a cholera outbreak in London to
a single water pump on Broad Street; the company supplying the pump drew
its water downstream from a sewage outflow, and the outbreak subsided
after the pump handle was removed. The investigation provided early
support for the germ theory of disease — that cholera was caused by
something in the water — against the then-prevailing miasma theory,
which attributed it to bad air.

What is new is the formality. Modern causal inference restates questions
like Snow's as questions about well-defined interventions: What would
have happened had the pump handle not been removed? How many cases would
have been prevented had it been removed a day earlier? These are
counterfactual quantities — statements about outcomes under conditions
that did not occur — and Chapter \@ref(potential-outcomes) introduces
the notation and assumptions that make them precise. As a formal
discipline, causal inference is young: it took shape over roughly the
past three decades, developing in parallel in statistics, epidemiology,
economics, and computer science. Its methods are now widely used and
increasingly accepted by drug regulatory agencies, studies built on them
appear regularly in leading medical journals, and the 2021 Nobel
Memorial Prize in Economic Sciences was awarded for work on drawing
causal conclusions from observational data.

The antipsychotic dilemma described earlier in this chapter was, in
fact, answered this way. Using Medicare claims data, investigators
compared elderly patients who newly started a conventional antipsychotic
with those who newly started an atypical agent — roughly 23,000 new
users in all — using both conventional adjustment methods and an
instrumental variable analysis that exploited variation in physicians'
prescribing preferences, a design developed in Chapter \@ref(iv). The
primary finding was an approximately 37% higher risk of death within 180
days among new users of the conventional agents. No single
nonexperimental study settles such a question, but the result argued
against switching patients to the older drugs, and no trial has emerged
to contradict it.

What made that study credible is worth noticing, because it previews
much of this book. The comparison was between two active treatments with
similar indications, so the patients being compared were far more alike
at baseline than treated patients and untreated patients would have
been, limiting confounding by indication. The outcome window was short,
leaving less room for changes in treatment and health status to
accumulate over follow-up. Both the exposure and the outcome could be
measured well in the available claims data. And the cohorts were
restricted to new users, avoiding the prevalent-user bias described
above. This combination — the new-user, active-comparator cohort with a
well-measured, short-term outcome — is something of a sweet spot for
pharmacoepidemiology, and it is the basis of most pharmacoepidemiologic
investigations, including the `ns` cohort at the center of this book.
The later parts of the book extend beyond that sweet spot, to questions
involving sustained treatment strategies, censoring, and unmeasured
confounding, where more is asked of both the data and the methods.

## A first look at the data

```{r setup-ch01, include=FALSE}
source("R/setup.R")
```

We will spend the whole book with the two cohorts introduced properly in
the next chapter, although it is worth looking at the simplest possible
comparison in one of them now, before any adjustment, to see what an
unadjusted answer to a causal question looks like.
In `ns`, the cohort of new NSAID users, patients either initiated a
nonselective NSAID or a Cox-2 selective inhibitor, and the outcome of
interest is incident peptic ulcer disease (PUD). The crude risk of PUD in
each group follows from a single `group_by()` and `summarise()`:

```{r crude-risk}
risk_tab <- ns |>
  group_by(cox2_initiation) |>
  summarise(n = n(), risk = mean(pud))

risk_tab |>
  knitr::kable(
    col.names = c("Initiated drug", "N", "Risk of incident PUD"),
    digits = c(0, 0, 3)
  )
```

Among the `r risk_tab$n[risk_tab$cox2_initiation == "No"]` patients who
initiated a nonselective NSAID,
`r sprintf("%.1f%%", 100 * risk_tab$risk[risk_tab$cox2_initiation == "No"])`
went on to develop incident PUD; among the
`r risk_tab$n[risk_tab$cox2_initiation == "Yes"]` patients who initiated a
Cox-2 selective drug, the figure was
`r sprintf("%.1f%%", 100 * risk_tab$risk[risk_tab$cox2_initiation == "Yes"])`.

Is that comparison an estimate of the causal effect of Cox-2 selectivity
on peptic ulcer risk? It is not, and there is reason to think it is far
from one. Nothing in the code above accounts for why a given patient
started one drug rather than the other. This is exactly the setting for
the confounding by indication described above: a patient with a history
of arthritis, prior GI bleeding, or concurrent corticosteroid or
aspirin use is more likely to be steered toward the Cox-2 selective
option. If Cox-2 initiators are systematically older or sicker in
ways connected to ulcer risk, as we will confirm with a proper Table 1 in
the next chapter, then the two rows of this table are not comparing like
with like, and the crude risk difference above combines whatever
causal effect the drug has with the baseline differences between the
groups. Separating those two components, that is, asking what the
comparison would look like if the two groups had been alike at baseline,
is the problem the rest of the book addresses.

## What this book covers

The book is organized in four parts. Part I, Foundations, which this
chapter opens, builds the conceptual vocabulary before any adjustment
method is introduced: potential outcomes, the randomized trial as the
benchmark against which a causal analysis is implicitly compared, causal
diagrams for reasoning about which variables require adjustment, and the
elements of study design that turn a causal question into an analyzable
data set. Part II, Core Methods, covers the two families of tools most
observational analyses reach for first: outcome regression and
standardization, and propensity scores, culminating in inverse
probability of treatment weighting. Part III, Time, Censoring, and
Survival, brings in what happens when the outcome does not occur at a
single fixed moment but can happen, or fail to happen, at any point over
follow-up: survival analysis, censoring, inverse probability of
censoring weights, and the clone-censor-weight approach to comparing
dynamic treatment strategies, all worked out on the statin cohort,
`sta`. Part IV, Beyond Exchangeability, deals with the harder case in
which adjustment for measured covariates is insufficient because of
unmeasured confounding, and introduces instrumental variables
and sensitivity analysis as ways of asking what an unmeasured confounder
would have to look like to explain away a finding.

Threaded through all four parts is a toolkit that you will build
yourself, one function at a time. By the end of the book you will have
written, tested, and reused `standardize()`
for outcome-model-based standardization, `iptw_estimator()` for
propensity-score weighting, `ipcw_estimator()` for censoring weights,
`ccw_estimator()` for the clone-censor-weight design, and `iv_estimator()`
for instrumental variable analysis. Each is introduced once, in the
chapter where its logic is developed, and then reused in a later
exercise on the other cohort, with a different set of covariates, or
with a different specification. Writing the functions yourself,
rather than only calling a package's implementation, is meant to make
the assumptions each method depends on explicit.

## How to read this book

The chapters are meant to be read with R running alongside them. Every
chapter is a self-contained R Markdown file that sources `R/setup.R` and
rebuilds what it needs from there, so you can open any chapter, run it
chunk by chunk, and confirm that the numbers in the text are the
numbers the code produces. The two risks in the table above, for
example, are computed from `ns` each time this chapter is rendered
rather than typed in. When a new function is introduced, read the code
closely enough to know what each argument does, then use it on data
slightly different from the example in the text before moving to the
exercises at the end of each chapter. The exercises themselves mostly do
not ask you to write new methods; they ask you to point a function you
already wrote at a different variable, a different cohort, or a
different specification, and to think about whether the answer it gives
still makes sense. That habit of applying a tool you understand to an
unfamiliar problem is closer to what causal inference looks like in
practice than anything else the book can offer.
