# (PART) Foundations {-}

# Why causal inference? {#why}

## Questions epidemiology cannot answer with association alone

This book keeps returning to two clinical questions, in different guises,
from Chapter 2 all the way to the end. Do Cox-2 selective NSAIDs cause
less peptic ulcer disease than nonselective NSAIDs? Do statins reduce the
risk of a cardiovascular event when they are used the way clinicians
actually use them, in routine care rather than in a trial? Both are
causal questions — they ask what would happen to a group of patients if
a particular treatment decision changed — and both are the kind of
question epidemiologists are asked to answer using data no one designed
for the purpose: medical records, claims, surveys, registries.

The trouble is that the most natural thing to do with such data — compare
patients who took a treatment to patients who did not — answers a
different, easier question: whether patients who took the treatment did
better or worse. That is an association. "Patients who took statins had
fewer cardiovascular events" is a statement about a difference in
outcomes between two groups of patients as they actually existed.
"Statin initiation causes fewer cardiovascular events" is a statement
about what would have happened to the same patients under a treatment
decision they did not actually make. The two statements can point in the
same direction, but nothing about them guarantees it, because the
patients who took the treatment and the patients who did not usually
differ in the first place, for reasons connected to their risk of the
outcome.

Two patterns of difference come up often enough in pharmacoepidemiology
to have names. The first is the **healthy-user effect**: people who
reliably take a preventive medication as prescribed also tend to
exercise more, smoke less, keep other appointments, and generally do the
many small things that predict good health, so a drug looks protective
partly because the kind of person who takes it was already going to do
well. Statins are a textbook case — some of the early apparent benefit of
statin therapy in observational data reflected the fact that adherent,
health-conscious patients are also less likely to have a heart attack for
reasons that have nothing to do with cholesterol. The second is
**confounding by indication**, sometimes called channeling: physicians
do not assign treatments at random, they assign them based on their
judgment of which patient needs them, so the treated group is
systematically different in exactly the dimension the treatment is meant
to address. A Cox-2 selective NSAID is often chosen, deliberately, for a
patient a physician already suspects is at higher gastrointestinal risk
— which means Cox-2 initiators can look worse for peptic ulcer disease
not because the drug fails to protect them, but because they were a
higher-risk group to start with. Get the direction of that confounding
wrong and you can even reverse the apparent effect of a genuinely
protective drug. A third pattern worth naming here because it will recur
throughout the book is prevalent-user bias: comparing people already on
a treatment to people never treated mixes in everyone who started the
drug, tolerated it, and stayed on it long enough to be captured in the
data, quietly dropping the patients who stopped early because of side
effects or because the drug was not working. The comparisons in this
book are built around new users — patients captured at the moment they
start a drug, as `ns` and `sta` are — specifically to avoid folding that
kind of survivorship into the treatment groups before the analysis even
begins.

These are not hypothetical worries. Observational studies of
postmenopausal hormone therapy in the 1990s found consistent,
substantial reductions in coronary heart disease among users; when the
Women's Health Initiative randomized women to hormone therapy or
placebo, the trial found increased risk, not decreased risk. The gap
was not sloppy analysis — it was healthier, more health-conscious, more
closely monitored women being more likely to use hormone therapy in the
first place, a difference no amount of simple adjustment fully closed.
If a body of careful observational research can get the sign of an
effect wrong, no amount of enthusiasm for a data set substitutes for a
framework that tells you when association does and does not equal
causation, and what to do about the gap. That framework is what the
rest of this book builds.

## A first look at the data

```{r setup-ch01, include=FALSE}
source("R/setup.R")
```

We will spend the whole book with the two cohorts introduced properly in
the next chapter, but it is worth looking at the simplest possible
comparison in one of them right now, before any of the machinery, just to
see what an unadjusted answer to a causal question actually looks like.
In `ns`, the cohort of new NSAID users, patients either initiated a
nonselective NSAID or a Cox-2 selective inhibitor, and the outcome of
interest is incident peptic ulcer disease (PUD). The crude risk of PUD in
each group is one `group_by()` and one `summarise()` away:

```{r crude-risk}
risk_tab <- ns |>
  group_by(cox2_initiation) |>
  summarise(n = n(), risk = mean(pud))

risk_tab |>
  knitr::kable(
    col.names = c("Initiated drug", "N", "Risk of incident PUD"),
    digits = c(0, 0, 3)
  )
```

Among the `r risk_tab$n[risk_tab$cox2_initiation == "No"]` patients who
initiated a nonselective NSAID,
`r sprintf("%.1f%%", 100 * risk_tab$risk[risk_tab$cox2_initiation == "No"])`
went on to develop incident PUD; among the
`r risk_tab$n[risk_tab$cox2_initiation == "Yes"]` patients who initiated a
Cox-2 selective drug, the figure was
`r sprintf("%.1f%%", 100 * risk_tab$risk[risk_tab$cox2_initiation == "Yes"])`.

Is that comparison an estimate of the causal effect of Cox-2 selectivity
on peptic ulcer risk? Not yet, and probably not even close. Nothing in
the code above knows anything about why a given patient started one drug
rather than the other. Cox-2 selective inhibitors were developed, and
are often chosen, specifically to reduce gastrointestinal risk in
patients thought to need that protection — a patient with a history of
arthritis, prior GI bleeding, or concurrent corticosteroid or aspirin use
is more likely to be steered toward the Cox-2 selective option in the
first place. If Cox-2 initiators are systematically older or sicker in
ways connected to ulcer risk, as we will confirm with a proper Table 1 in
the next chapter, then the two rows of this table are not comparing like
with like, and the crude risk difference above is a mix of whatever
causal effect the drug has and however different the two groups already
were. Untangling that mix — figuring out what the comparison would look
like if the two groups had been alike to begin with — is the problem this
book exists to solve.

## What this book covers

The book is organized in four parts. Part I, Foundations, which this
chapter opens, builds the conceptual vocabulary — potential outcomes, the
randomized trial as the benchmark a causal analysis is implicitly
compared against, causal diagrams for reasoning about which variables
need adjustment, and the elements of study design that turn a causal
question into an analyzable data set — before any adjustment method is
introduced. Part II, Core Methods, covers the two families of tools most
observational analyses reach for first: outcome regression and
standardization, and propensity scores, culminating in inverse
probability of treatment weighting. Part III, Time, Censoring, and
Survival, brings in what happens when the outcome does not occur at a
single fixed moment but can happen, or fail to happen, at any point over
follow-up — survival analysis, censoring, inverse probability of
censoring weights, and the clone-censor-weight approach to comparing
dynamic treatment strategies, all worked out on the statin cohort,
`sta`. Part IV, Beyond Exchangeability, deals with the harder case where
no amount of adjustment for measured covariates will do, because an
unmeasured confounder is at work, and introduces instrumental variables
and sensitivity analysis as ways of asking what an unmeasured confounder
would have to look like to explain away a finding.

Threaded through all four parts is a toolkit you will build yourself, one
function at a time, rather than read about in the abstract. By the end
of the book you will have written, tested, and reused `standardize()`
for outcome-model-based standardization, `iptw_estimator()` for
propensity-score weighting, `ipcw_estimator()` for censoring weights,
`ccw_estimator()` for the clone-censor-weight design, and `iv_estimator()`
for instrumental variable analysis. Each is introduced once, in the
chapter where its logic is developed, and then reused — on the other
cohort, with a different set of covariates, or with a different
specification — in a later exercise. Writing the functions yourself,
rather than only calling a package's version, is meant to make the
assumptions each method depends on concrete rather than hidden inside
someone else's code.

## How to read this book

There is no substitute for running the code. Every chapter is a
self-contained R Markdown file that sources `R/setup.R` and rebuilds
everything it needs from there, so you can open any chapter, run it chunk
by chunk, and see for yourself that the numbers in the text are the
numbers the code produces — including the two risks in the table above,
which are not typed in but computed live from `ns` every time this
chapter is rendered. When a new function is introduced, read the code
closely enough to know what each argument does, then use it on data
slightly different from the example in the text before moving to the
exercises at the end of each chapter. The exercises themselves mostly do
not ask you to write new methods; they ask you to point a function you
already wrote at a different variable, a different cohort, or a
different specification, and to think about whether the answer it gives
still makes sense. That habit — reusing a tool you understand on a new
problem, rather than reaching for an unfamiliar one — is closer to what
causal inference looks like in practice than anything else this book can
teach you.
