# (PART) Foundations {-}

# Why causal inference? {#why}

## Questions epidemiology cannot answer with association alone

This book keeps returning to two clinical questions, in different guises,
from Chapter 2 onward. Do Cox-2 selective NSAIDs cause
less peptic ulcer disease than nonselective NSAIDs? Do statins reduce the
risk of a cardiovascular event when they are used the way clinicians
actually use them, in routine care rather than in a trial? Both are
causal questions, in that they ask what would happen to a group of
patients if a particular treatment decision changed, and both are the
kind of question epidemiologists are asked to answer using data
collected for other purposes: medical records, claims, surveys,
registries.

The trouble is that the most natural thing to do with such data, namely
compare patients who took a treatment to patients who did not, answers a
different and easier question: whether patients who took the treatment
did better or worse. Such a comparison yields an association. "Patients
who took statins had
fewer cardiovascular events" is a statement about a difference in
outcomes between two groups of patients as they actually existed.
"Statin initiation causes fewer cardiovascular events" is a statement
about what would have happened to the same patients under a treatment
decision they did not actually make. The two statements can point in the
same direction, but nothing about them guarantees it, because the
patients who took the treatment and the patients who did not usually
differ in the first place, for reasons connected to their risk of the
outcome.

Two patterns of difference come up often enough in pharmacoepidemiology
to have names. The first is the **healthy-user effect**: people who
reliably take a preventive medication as prescribed also tend to
exercise more, smoke less, keep other appointments, and generally do the
many small things that predict good health, so a drug looks protective
partly because the kind of person who takes it was already going to do
well. Statins are a familiar example: some of the early apparent benefit
of statin therapy in observational data reflected the fact that adherent,
health-conscious patients are also less likely to have a heart attack for
reasons unrelated to cholesterol. The second is
**confounding by indication**, sometimes called channeling: physicians
assign treatments according to their judgment of which patient needs
them, so the treated group differs
systematically in exactly the dimension the treatment is meant
to address. A Cox-2 selective NSAID is often chosen, deliberately, for a
patient a physician already suspects is at higher gastrointestinal risk,
so that Cox-2 initiators can appear to fare worse with respect to peptic
ulcer disease not because the drug fails to protect them but because
they were a higher-risk group at baseline. If the direction of that
confounding is misjudged, the apparent effect of a genuinely protective
drug can be reversed. A third pattern worth naming here because it will
recur throughout the book is prevalent-user bias: comparing people
already on a treatment to people never treated mixes in everyone who
started the drug, tolerated it, and stayed on it long enough to be
captured in the data, while excluding the patients who stopped early
because of side effects or lack of benefit. The comparisons in this
book are built around new users — patients captured at the moment they
start a drug, as `ns` and `sta` are — specifically to avoid folding that
kind of survivorship into the treatment groups before the analysis
begins.

These concerns are not hypothetical. Observational studies of
postmenopausal hormone therapy in the 1990s reported that use was
associated with substantially lower risk of coronary heart disease;
when the Women's Health Initiative randomized women to hormone therapy
or placebo, the trial found an increased risk. The discrepancy did not
arise from careless analysis. Rather, women who used hormone therapy
were healthier, more health-conscious, and more closely monitored than
women who did not, a difference that simple adjustment did not fully
remove. If a body of careful nonexperimental research can get the sign
of an effect wrong, then a large or richly detailed data set is not by
itself sufficient; what is needed is a framework that says when an
association can be given a causal interpretation, and what to do when
it cannot. The rest of the book develops such a framework.

## A first look at the data

```{r setup-ch01, include=FALSE}
source("R/setup.R")
```

We will spend the whole book with the two cohorts introduced properly in
the next chapter, although it is worth looking at the simplest possible
comparison in one of them now, before any adjustment, to see what an
unadjusted answer to a causal question looks like.
In `ns`, the cohort of new NSAID users, patients either initiated a
nonselective NSAID or a Cox-2 selective inhibitor, and the outcome of
interest is incident peptic ulcer disease (PUD). The crude risk of PUD in
each group follows from a single `group_by()` and `summarise()`:

```{r crude-risk}
risk_tab <- ns |>
  group_by(cox2_initiation) |>
  summarise(n = n(), risk = mean(pud))

risk_tab |>
  knitr::kable(
    col.names = c("Initiated drug", "N", "Risk of incident PUD"),
    digits = c(0, 0, 3)
  )
```

Among the `r risk_tab$n[risk_tab$cox2_initiation == "No"]` patients who
initiated a nonselective NSAID,
`r sprintf("%.1f%%", 100 * risk_tab$risk[risk_tab$cox2_initiation == "No"])`
went on to develop incident PUD; among the
`r risk_tab$n[risk_tab$cox2_initiation == "Yes"]` patients who initiated a
Cox-2 selective drug, the figure was
`r sprintf("%.1f%%", 100 * risk_tab$risk[risk_tab$cox2_initiation == "Yes"])`.

Is that comparison an estimate of the causal effect of Cox-2 selectivity
on peptic ulcer risk? It is not, and there is reason to think it is far
from one. Nothing in the code above accounts for why a given patient
started one drug
rather than the other. Cox-2 selective inhibitors were developed, and
are often chosen, specifically to reduce gastrointestinal risk in
patients thought to need that protection. For example, a patient with a
history of arthritis, prior GI bleeding, or concurrent corticosteroid or
aspirin use is more likely to be steered toward the Cox-2 selective
option. If Cox-2 initiators are systematically older or sicker in
ways connected to ulcer risk, as we will confirm with a proper Table 1 in
the next chapter, then the two rows of this table are not comparing like
with like, and the crude risk difference above combines whatever
causal effect the drug has with the baseline differences between the
groups. Separating those two components, that is, asking what the
comparison would look like if the two groups had been alike at baseline,
is the problem the rest of the book addresses.

## What this book covers

The book is organized in four parts. Part I, Foundations, which this
chapter opens, builds the conceptual vocabulary before any adjustment
method is introduced: potential outcomes, the randomized trial as the
benchmark against which a causal analysis is implicitly compared, causal
diagrams for reasoning about which variables require adjustment, and the
elements of study design that turn a causal question into an analyzable
data set. Part II, Core Methods, covers the two families of tools most
observational analyses reach for first: outcome regression and
standardization, and propensity scores, culminating in inverse
probability of treatment weighting. Part III, Time, Censoring, and
Survival, brings in what happens when the outcome does not occur at a
single fixed moment but can happen, or fail to happen, at any point over
follow-up: survival analysis, censoring, inverse probability of
censoring weights, and the clone-censor-weight approach to comparing
dynamic treatment strategies, all worked out on the statin cohort,
`sta`. Part IV, Beyond Exchangeability, deals with the harder case in
which adjustment for measured covariates is insufficient because of
unmeasured confounding, and introduces instrumental variables
and sensitivity analysis as ways of asking what an unmeasured confounder
would have to look like to explain away a finding.

Threaded through all four parts is a toolkit that you will build
yourself, one function at a time. By the end of the book you will have
written, tested, and reused `standardize()`
for outcome-model-based standardization, `iptw_estimator()` for
propensity-score weighting, `ipcw_estimator()` for censoring weights,
`ccw_estimator()` for the clone-censor-weight design, and `iv_estimator()`
for instrumental variable analysis. Each is introduced once, in the
chapter where its logic is developed, and then reused in a later
exercise on the other cohort, with a different set of covariates, or
with a different specification. Writing the functions yourself,
rather than only calling a package's implementation, is meant to make
the assumptions each method depends on explicit.

## How to read this book

The chapters are meant to be read with R running alongside them. Every
chapter is a self-contained R Markdown file that sources `R/setup.R` and
rebuilds what it needs from there, so you can open any chapter, run it
chunk by chunk, and confirm that the numbers in the text are the
numbers the code produces. The two risks in the table above, for
example, are computed from `ns` each time this chapter is rendered
rather than typed in. When a new function is introduced, read the code
closely enough to know what each argument does, then use it on data
slightly different from the example in the text before moving to the
exercises at the end of each chapter. The exercises themselves mostly do
not ask you to write new methods; they ask you to point a function you
already wrote at a different variable, a different cohort, or a
different specification, and to think about whether the answer it gives
still makes sense. That habit of applying a tool you understand to an
unfamiliar problem is closer to what causal inference looks like in
practice than anything else the book can offer.
