# Getting started with R and the NAMCS data

```{r setup-ch02, include=FALSE}
source("R/setup.R")
```

Every idea in this book is attached to code, and the code is R. If you
have never used R before, this chapter is for you: it introduces just
enough of the tidyverse to read and write the data manipulations in the
chapters that follow, walks through the two example cohorts you will be
living with for the rest of the semester, and ends by building your first
reusable function — a "Table 1" generator you will reach for in every
observational study you ever run, in this course and after it. If you
already know R and the tidyverse well, skim this chapter for the two data
sets and read the Table 1 section, since `table1_continuous()` and
`table1_categorical()` are used again in later chapters.

## R, RStudio, and the tidyverse

R is the language; RStudio is the editor most people use to write and run
it. Appendix A walks through installing both, along with the packages
this book relies on. Once they are installed, you will type or paste code
into a script or, as here, into a code chunk inside an R Markdown
document, and run it a line or a chunk at a time.

The basic vocabulary of R is small. An **object** is a name that holds a
value — a number, a piece of text, or, most often in this book, a
rectangular data set with rows and columns (a **data frame**, or in the
tidyverse a **tibble**). A **function** takes some input, does something
to it, and returns an output; `mean(x)` takes a vector `x` and returns
its average. A **package** is a bundle of functions someone else wrote
and shared, loaded with `library()`. The tidyverse is one such bundle —
really a collection of packages, including `dplyr` for manipulating data
frames and `ggplot2` for plotting — built around a shared design that
tries to make code read like a description of what you want done rather
than a set of instructions for how a computer should do it.

The setup chunk at the top of this chapter already did the loading for
us: `source("R/setup.R")` runs a script that attaches the tidyverse,
attaches the `NAMCS` package that supplies our example data, and builds
the two analysis cohorts, `ns` and `sta`, that we will use throughout the
book. Every chapter starts this way, so that it can be read and rendered
on its own without depending on anything that came before it.

The one piece of tidyverse syntax you should get comfortable with
immediately is the pipe, `|>`. It takes whatever is on its left and
passes it as the first argument to whatever is on its right, so that
`x |> f()` means the same thing as `f(x)`. This lets you chain several
operations together and read them roughly left to right, top to bottom,
as a sequence of steps: start with the data, then filter it, then
summarize it. Four `dplyr` verbs will carry most of the weight in this
book: `filter()` keeps rows that satisfy a condition, `select()` keeps
particular columns, `mutate()` adds or changes a column, and
`group_by()` paired with `summarise()` computes a statistic separately
within each level of a grouping variable. A fifth, `count()`, is a
shortcut for counting rows within groups.

A few examples, all using the NSAID cohort `ns`:

```{r verbs-demo}
ns |> filter(age >= 65) |> nrow()

ns |> select(cox2_initiation, age, sex) |> head(3)

ns |> mutate(age65 = as.integer(age >= 65)) |> count(age65)

ns |>
  group_by(cox2_initiation) |>
  summarise(n = n(), mean_age = mean(age))

ns |> count(sex, cox2_initiation)
```

The first line counts how many patients in `ns` are 65 or older. The
second pulls out three columns and prints the first few rows. The third
uses `mutate()` to add a new column, `age65`, that recodes age as a 0/1
indicator for being 65 or older, and then counts how many patients land
in each category — the same recode-and-count pattern behind `cox2_init`
and `pud`, the two convenience columns `R/setup.R` added for us. The
fourth groups patients by which drug class they initiated and reports
the number of patients and mean age in each group — a first, very small
taste of the comparisons we will spend the rest of the book refining. The
fifth cross-tabulates sex against treatment group. None of this is
causal inference yet; it is just description, but description is where
every analysis starts, and it is worth being fluent in it before adding
any causal machinery on top.

## The NAMCS teaching cohorts

The data behind every example in this book come from the National
Ambulatory Medical Care Survey (NAMCS), an annual survey of visits to
office-based physicians in the United States. We use five years of
NAMCS, 2005 through 2009, assembled into two teaching cohorts by the
`NAMCS` R package. Both cohorts are built from real survey visits, but
the outcomes — peptic ulcer disease in one cohort, cardiovascular events
and survival times in the other — are simulated, so that we know there
is a true underlying data-generating process while still working with
realistic covariate structure. This is common in causal inference
teaching materials: it lets us illustrate methods with plausible clinical
questions without the ambiguity of relying on outcomes we cannot
validate.

The first cohort, `ns`, follows new users of NSAIDs — patients starting
either a nonselective NSAID or a Cox-2 selective inhibitor — and asks
whether the choice of drug class affects the risk of incident peptic
ulcer disease. `glimpse()` gives a quick sense of what is in it:

```{r glimpse-ns}
glimpse(ns)
```

There are `r nrow(ns)` patients in `ns`, of whom
`r sum(ns$cox2_initiation == "Yes")` initiated a Cox-2 selective
inhibitor. Alongside the treatment indicator `cox2_initiation` and the
outcome `incident_pud`, the cohort carries baseline covariates: age, sex,
race, region, calendar year, a set of chronic conditions, and
comedications like corticosteroids, aspirin, and gastroprotective drugs.
`R/setup.R` already added two convenience columns, `cox2_init` and `pud`,
which are the same information as `cox2_initiation` and `incident_pud`
recoded as 0/1 integers — most of the model-fitting functions we write
later expect a numeric 0/1 treatment or outcome rather than a factor.

The second cohort, `sta`, follows patients who did or did not initiate a
statin, and asks whether statin initiation reduces the risk of a
cardiovascular event. Because a cardiovascular event can happen — or
fail to happen — at any point over years of follow-up, this cohort looks
a little different:

```{r glimpse-sta}
glimpse(sta)
```

There are `r nrow(sta)` patients in `sta`, and `r sum(sta$statin == 1)`
of them initiated a statin. Instead of a single yes/no outcome measured
at a fixed time, `sta` carries several time variables — `cv_time`,
`death_time`, `loss_fu_time` — recording, in years, when each event
would have occurred had it occurred at all (a patient who never has the
event has a missing `cv_time`). `R/setup.R` combines these into a single
`event_time`, the time to whichever of a cardiovascular event, death,
loss to follow-up, or the administrative end of the study (`admin_end`,
set to `r admin_end` years) happened first, and a `cv_event` indicator for whether
that first thing was actually the cardiovascular event. We will not need
any of that machinery until Chapter 10, on survival outcomes and
censoring; for now it is enough to know that `sta` exists, that it is
bigger and slower-moving than `ns`, and that it is where the book turns
once we start worrying about follow-up time.

Appendix B collects a full codebook for both cohorts, generated directly
from the data rather than typed out by hand, so it will always match
what is actually in the package. If you are ever unsure what a variable
means, that is the place to check.

## Plotting with ggplot2

`ggplot2` builds a plot by mapping columns of data to visual properties
— an x-axis, a y-axis, a color — and then adding layers on top that
say how to draw it: as points, bars, lines, or histograms. Every
`ggplot2` call in this book starts the same way, with `ggplot(data,
aes(...))` naming the data and the mapping, followed by one or more
`geom_*()` layers.

A natural first question about `ns` is whether Cox-2 initiators and
nonselective NSAID initiators look similar in age. A histogram, faceted
by treatment group, answers this directly:

```{r age-hist, fig.cap = "Age distribution by initiated drug class."}
ggplot(ns, aes(age)) +
  geom_histogram(binwidth = 5, boundary = 0) +
  facet_wrap(~cox2_initiation) +
  labs(x = "Age (years)", y = "Number of patients",
       title = "Age distribution by initiated drug class")
```

The two distributions are not identical, and that alone is worth
sitting with for a moment: it means age could confound any simple
comparison of peptic ulcer risk between the groups, since age is
plausibly related to both which drug a physician chooses to prescribe
and the baseline risk of peptic ulcer disease.

A second question is whether the two groups differ in their burden of
relevant comorbidities and comedications. A bar chart of prevalence, one
bar per condition per group, is a compact way to see several
comparisons at once:

```{r comorbid-bars, fig.cap = "Prevalence of selected comorbidities and comedications by initiated drug class."}
ns |>
  select(cox2_initiation, arthritis, diabetes, hypertension,
         corticosteroid_use) |>
  pivot_longer(-cox2_initiation, names_to = "condition") |>
  group_by(cox2_initiation, condition) |>
  summarise(prevalence = mean(value == "Yes"), .groups = "drop") |>
  ggplot(aes(condition, prevalence, fill = cox2_initiation)) +
  geom_col(position = "dodge") +
  labs(x = NULL, y = "Prevalence", fill = "Initiated Cox-2?") +
  coord_flip()
```

Arthritis, in particular, is more common among Cox-2 initiators than
among nonselective NSAID initiators, which makes clinical sense — Cox-2
selective drugs are often chosen specifically for patients thought to be
at higher gastrointestinal risk — and is exactly the kind of pattern that
will occupy us once we start asking how to adjust for confounding.

## Your first reusable functions: Table 1

Every observational paper you will ever read, and every one you will
ever write, opens with a table comparing the characteristics of the
groups being compared — universally, if a little unimaginatively, called
"Table 1." It is descriptive, not causal, but it is also indispensable:
it is where a reader first sees whether the groups look similar,
and it is often where a reader first suspects they don't. You will make
some version of this table for every study you ever do, so rather than
rebuild it from scratch each time, we are going to write it once, as a
function, and reuse it for the rest of this book.

A function in R is defined with `function(...)`, naming its arguments,
followed by a body that computes something from them and returns it (by
default, the value of the last expression evaluated). Writing your own
functions is the single habit that will save you the most time as an R
user, because it turns a block of code you would otherwise copy, paste,
and slightly mis-edit into something you write once, test once, and
trust afterward.

A Table 1 needs to summarize two kinds of variables differently:
continuous variables as a mean and standard deviation, and categorical
variables as a count and percentage, both computed separately within
each level of a grouping variable (here, treatment group). We will write
one function for each.

For continuous variables, the recipe is: reshape the requested columns
into a long format with one row per patient per variable, group by the
grouping variable and the variable name, compute `"mean (sd)"` as a
formatted string, and reshape back so that each treatment group is its
own column.

```{r table1-continuous-def}
table1_continuous <- function(data, vars, by) {
  data |>
    select(all_of(c(by, vars))) |>
    pivot_longer(all_of(vars), names_to = "variable") |>
    group_by(.data[[by]], variable) |>
    summarise(
      stat = sprintf("%.1f (%.1f)",
                     mean(value, na.rm = TRUE), sd(value, na.rm = TRUE)),
      .groups = "drop"
    ) |>
    pivot_wider(names_from = all_of(by), values_from = stat)
}
```

A couple of details are worth pointing out because they recur throughout
this book. `all_of(vars)` tells `select()` and `pivot_longer()` that
`vars` is a character vector of column names supplied by the caller,
rather than a single column to look up literally — this is what lets the
function work for any set of variables, not just the ones we happen to
try first. `.data[[by]]` does the analogous thing for `group_by()`,
letting the name of the grouping column be a string passed in as an
argument rather than hard-coded. Together these are what make the
function *reusable*: nothing about age or treatment group is written
into the body of `table1_continuous()` itself.

Categorical variables follow the same shape, but count levels instead of
averaging values:

```{r table1-categorical-def}
table1_categorical <- function(data, vars, by) {
  data |>
    select(all_of(c(by, vars))) |>
    mutate(across(all_of(vars), as.character)) |>
    pivot_longer(all_of(vars), names_to = "variable", values_to = "level") |>
    count(.data[[by]], variable, level) |>
    group_by(.data[[by]], variable) |>
    mutate(stat = sprintf("%d (%.1f%%)", n, 100 * n / sum(n))) |>
    ungroup() |>
    select(-n) |>
    pivot_wider(names_from = all_of(by), values_from = stat)
}
```

Both functions are saved in `R/table1.R` so that later chapters can pull
them in with a single `source()` call rather than redefining them. Let's
use them on the NSAID cohort, comparing Cox-2 initiators to nonselective
NSAID initiators on age and on five categorical covariates, and format
the results with `knitr::kable()` for a readable table:

```{r table1}
source("R/table1.R")
table1_continuous(ns, vars = "age", by = "cox2_initiation") |> knitr::kable()
table1_categorical(
  ns,
  vars = c("sex", "arthritis", "aspirin_use", "ppi_use", "corticosteroid_use"),
  by = "cox2_initiation"
) |> knitr::kable()
```

Look at the two tables side by side. Cox-2 initiators are, on average,
older than nonselective NSAID initiators, and a substantially larger
share of them have a history of arthritis. Neither table tells us
whether Cox-2 selectivity affects peptic ulcer risk — that question
needs the rest of this book — but both are exactly the kind of evidence
an epidemiologist reaches for first, and both came from two functions
you now understand well enough to have written yourself.

## Exercises

1. Build both a continuous and a categorical Table 1 for the statin
   cohort, comparing statin initiators to non-initiators
   (`statin_use`) on `age` and on `diabetes`, `hypertension`, and
   `tobacco_use`. Use `table1_continuous()` and `table1_categorical()`
   exactly as they were used above, just pointing them at `sta` and a
   different `by` variable and variable list.
2. The age-by-group histogram in the *Plotting with ggplot2* section was
   built for the NSAID cohort. Adapt that code to plot the distribution
   of `age` by `statin_use` in the `sta` cohort instead. Do the two
   statin groups look as similar in age as the two NSAID groups did?
3. Using `count()`, tabulate the number of visits per calendar `year`
   separately for statin initiators and non-initiators in `sta`. Is the
   number of statin initiators roughly constant across 2005–2009, or
   does it trend up or down?
