# Potential outcomes and causal effects

```{r setup-ch03, include=FALSE}
source("R/setup.R")
```

Chapter 1 asked whether Cox-2 selective NSAIDs cause less peptic ulcer
disease than nonselective NSAIDs, and then showed that the obvious way of
answering the question — comparing crude risks between the two groups —
does not actually answer it, because Cox-2 initiators and nonselective
initiators differ in ways connected to ulcer risk before either drug is
ever taken. This chapter builds the vocabulary that makes precise what
went wrong, and what a right answer would even mean. The framework is
called **potential outcomes**, and once you have it, "causal effect" stops
being a vague intuition and becomes something you can write down,
simulate, and eventually estimate.

## Two futures for every patient

Picture a single patient in the `ns` cohort — call her Patient 47 — who
shows up at her physician's office with joint pain that warrants an
NSAID. Her physician has a choice: prescribe a nonselective NSAID, or
prescribe a Cox-2 selective inhibitor. Whichever one is chosen, Patient
47 either does or does not go on to develop peptic ulcer disease over
the following year. But the question "would she have developed PUD if
she had been treated differently" has a sensible answer even though it
did not happen. If she was in fact given the Cox-2 drug and stayed
ulcer-free, we can still ask, coherently, what would have happened to
her, with everything else about her held fixed, if she had instead
received the nonselective drug.

Potential outcomes give that question a name and a symbol. Let $A$
denote the treatment Patient 47 receives, coded $A = 1$ for the Cox-2
selective drug and $A = 0$ for the nonselective drug, and let $Y$ denote
whether she develops incident PUD. Then $Y^1$ is the outcome she would
have if she were treated with $A = 1$, and $Y^0$ is the outcome she
would have if she were treated with $A = 0$. These are **potential**
outcomes — one for each treatment value she could in principle receive —
and both exist, in the sense of being well-defined quantities, before we
know which treatment she actually gets. Once treatment is assigned, one
of them becomes the **observed** outcome and the other becomes
counterfactual: contrary to the fact of what happened. If Patient 47
actually receives the Cox-2 drug, we observe $Y^1$; $Y^0$ is what
her chart would show in the world where her physician wrote the other
prescription.

A word on notation, because it varies more than the idea does. We write
these potential outcomes with superscripts, but many authors — and the
PHS 921 lectures and exams — write $Y(1)$ and $Y(0)$, and others write
$Y_a$ for the outcome under a generic treatment value $a$. All three
notations denote the same objects, so it is worth learning to move
between them freely rather than attaching the concept to any one of
them.

With both potential outcomes in hand for a single patient, her
**individual causal effect** of Cox-2 selectivity on PUD is simply the
contrast between them, $Y^1 - Y^0$. If $Y^1 = 0$ and $Y^0 = 1$, the
Cox-2 drug prevented her ulcer; if both are 0, or both are 1, the choice
of drug made no difference to her outcome, even though the drugs may
differ enormously for other patients. This is a clean, well-defined
quantity — and it is also, for any real patient, unobservable. We only
ever see one of Patient 47's two potential outcomes, because she can
only actually take one drug. This is what Holland (1986) called the
**fundamental problem of causal inference**: for any individual, at
most one potential outcome is ever realized, and the individual causal
effect can never be computed from data on that individual alone, no
matter how much we know about her or how long we follow her. Every method in this book is,
at bottom, a strategy for getting around this problem — not by
recovering the missing potential outcome for any one patient, but by
using groups of patients to estimate causal effects that are well
defined on average, even though they are unknowable one person at a
time.

## A god's-eye view: simulation

The fundamental problem of causal inference is a genuine, permanent
limitation of real data: no data set on real patients ever contains
both of anyone's potential outcomes. But nothing stops us from
simulating a toy population in which we, as the people writing the
code, get to decide both potential outcomes for every patient, and then
watch what happens when only one becomes observed. This is a god's-eye
view of exactly the causal question Chapter 1 raised, and it lets us
see confounding happen mechanically, in code, rather than only argue
about it in prose.

Imagine 100,000 patients with a severity indicator, `severity`, that
plays the role of an unmeasured driver of both treatment and outcome —
capturing whatever led some patients toward the Cox-2 drug and also
independently raised their ulcer risk. For each patient we generate
*both* potential outcomes, `y0` (the outcome under nonselective NSAID)
and `y1` (the outcome under Cox-2 selective drug), with the Cox-2
potential outcome built to carry a lower baseline risk — the drug is
genuinely protective in this toy world. We then assign actual treatment
`a` so that more severe patients are more likely to receive the Cox-2
drug, exactly the channeling Chapter 1 described, and finally construct
the single outcome `y` actually recorded: the potential outcome
matching the treatment each patient actually got.

```{r sim-po}
set.seed(921)
truth <- tibble(
  severity = rbinom(100000, 1, 0.3),                       # confounder
  y0 = rbinom(100000, 1, 0.05 + 0.15 * severity),          # risk untreated
  y1 = rbinom(100000, 1, 0.03 + 0.15 * severity),          # risk treated
  a  = rbinom(100000, 1, 0.2 + 0.5 * severity),            # treated more if severe
  y  = if_else(a == 1, y1, y0)                             # consistency
)
truth |> summarise(
  ate   = mean(y1) - mean(y0),
  att   = mean(y1[a == 1]) - mean(y0[a == 1]),
  naive = mean(y[a == 1]) - mean(y[a == 0])
)
```

```{r sim-po-values, include=FALSE}
sim_summary <- truth |> summarise(
  ate   = mean(y1) - mean(y0),
  att   = mean(y1[a == 1]) - mean(y0[a == 1]),
  naive = mean(y[a == 1]) - mean(y[a == 0])
)
```

Because we generated `y0` and `y1` ourselves, we can compute the true
average causal effect directly from columns no real data set would ever
have both of at once: `mean(y1) - mean(y0)` is
`r sprintf("%.3f", sim_summary$ate)`, a negative number, confirming that
treatment is, on average, beneficial in this toy population — it lowers
the risk of the outcome by about
`r sprintf("%.1f", -100 * sim_summary$ate)` percentage points. But the
last line of the summary, `naive`, is the only contrast we could ever
compute from a real observational data set — the mean of the *observed*
outcome `y` among the treated minus the mean of `y` among the untreated
— and it comes out to `r sprintf("%.3f", sim_summary$naive)`, positive
and on the wrong side of zero. The naive comparison says Cox-2
selectivity looks harmful; the truth we built into the simulation is
that it is helpful. The gap between `r sprintf("%.3f", sim_summary$naive)`
and `r sprintf("%.3f", sim_summary$ate)` is confounding by `severity`,
laid completely bare because in this one artificial world we happen to
know the whole truth: severe patients were both more likely to be
treated *and* more likely to have a bad outcome regardless of treatment,
so the treated group looks worse than it would have if severity had
been balanced across the two groups. Real data never hands you `y0` and
`y1` side by side like this — but the machinery in the rest of the
chapter exists precisely to close that gap without needing to.

## Average causal effects and estimands

The individual causal effect $Y^1 - Y^0$ cannot be observed, but its
*average* over a population can be, once we have the right tools —
which is why causal inference works with population-level summaries
called **estimands**. The one that shows up most often in this book is
the **average treatment effect (ATE)**,
$$\text{ATE} = E[Y^1] - E[Y^0],$$
the average, over everyone in the population of interest, of what would
happen if everyone were treated minus what would happen if no one were.
In the simulation above, `mean(y1) - mean(y0)` computed over all 100,000
patients *is* the ATE, because we happened to have both potential
outcomes for every one of them.

A second estimand, the **average treatment effect in the treated
(ATT)**,
$$\text{ATT} = E[Y^1 \mid A = 1] - E[Y^0 \mid A = 1],$$
restricts that same contrast to the patients who were actually treated:
it asks what treatment did for the people who received it, not for the
population as a whole. In the simulation, that is
`mean(y1[a == 1]) - mean(y0[a == 1])`, and in our output it comes out
very close to, but not exactly equal to, the ATE. That small remaining
gap is Monte Carlo noise, not confounding: in this particular
simulation the treatment effect `y1 - y0` is built to be the same
constant regardless of `severity`, so the *true* ATT equals the *true*
ATE exactly, and the two only look different because both are estimated
from a single finite sample of 100,000 patients rather than computed on
the population itself. Rerunning with a different seed would shift the
gap around, and often shrink it, without ever making it disappear
completely.

The reason the two estimands can genuinely diverge, rather than merely
look different by chance, is when the *effect itself* varies with
whatever determines who gets treated. Which estimand you want then
depends on the question. If a health system is deciding whether to
switch every future NSAID prescription for arthritis patients to the
Cox-2 selective drug, the ATE is the relevant quantity — it answers
"what would happen if we treated everyone." If a clinician is instead
trying to understand the effect of the prescriptions she has already
written, on the patients who actually received them, the ATT is closer
to what she wants — it answers "was this decision, for the patients who
got it, a good one." The two estimands coincide only when treatment
effects do not vary with whatever determines who gets treated, exactly
the case built into this simulation; Exercise 3 asks you to break that
by making the effect depend on `severity`, so that the ATE and ATT
genuinely part ways.

Any of these estimands can be reported on more than one scale. With $Y$
binary, as PUD is, the most direct scale is the **risk difference**,
$E[Y^1] - E[Y^0]$, interpretable as "how many fewer (or more) cases per
100 patients treated." The **risk ratio**, $E[Y^1] / E[Y^0]$, expresses
the same contrast multiplicatively — a risk ratio of 0.7 means treatment
reduces risk by 30%, regardless of baseline risk, which often makes it
more portable across populations than a difference. The **odds ratio**,
$$\frac{E[Y^1]/(1-E[Y^1])}{E[Y^0]/(1-E[Y^0])},$$
is the scale most familiar from logistic regression and case-control
studies, but it approximates the risk ratio only when the outcome is
rare, and can otherwise diverge from it substantially. None of
consistency, exchangeability, or positivity — the identification
conditions we turn to next — depend on which scale you eventually
report; the choice of scale is a separate decision, made after
identification, about how to communicate a causal contrast you have
already secured.

## The identification conditions

Estimands like the ATE are defined in terms of potential outcomes we
never fully observe. **Identification** is the step of showing that,
under stated assumptions, an estimand written in potential outcomes is
equal to some quantity we actually could compute from observed data.
Three conditions do this work, and it is worth attaching each one to
the NSAID cohort so that they read as claims about real patients rather
than abstract axioms.

**Consistency** says that a patient's observed outcome equals the
potential outcome under the treatment she actually received: if Patient
47 takes the Cox-2 drug, her observed $Y$ equals $Y^1$, not some
average or blend involving $Y^0$. This sounds almost too obvious to
state, but it is doing real work — it is what licenses the line
`y = if_else(a == 1, y1, y0)` in the simulation above, and it can fail
in practice when "treatment" is not well defined enough to pin down a
single potential outcome. If "Cox-2 initiation" secretly bundles several
different doses, durations, or concurrent gastroprotective strategies,
then $Y^1$ is not one well-defined quantity but several, and consistency
is violated before any statistical issue even arises.

**Exchangeability** (also called ignorability) says that the treated
and untreated groups would have had the same average outcome had they
received the same treatment — formally, $Y^1, Y^0 \perp A$, potential
outcomes are statistically independent of the treatment actually
received. This is exactly what fails in our simulation: because `a`
depends on `severity`, and `severity` also drives `y0` and `y1`, treated
and untreated patients are not exchangeable, and that is the entire
reason the naive contrast diverged from the ATE. Exchangeability is what
a randomized trial buys you by construction — Chapter 4 makes that
precise — and it is what confounding by indication takes away in
observational data like `ns`, where physicians route sicker-looking
patients toward the Cox-2 drug for reasons connected to the very outcome
we are studying.

Full exchangeability is usually too much to hope for in observational
data, which is why most of this book works with a weaker version,
**conditional exchangeability**: $Y^1, Y^0 \perp A \mid L$ for some set
of measured covariates $L$. Treatment need not be independent of
potential outcomes overall, only *within levels* of $L$ — among
patients who share the same measured `severity`, say, treatment
assignment is as good as random. This is a much more plausible
assumption whenever we can measure the variables driving treatment
choice, and it is exactly what standardization, propensity scores, and
IPTW in later chapters are built to exploit.

We can see conditional exchangeability rescue our toy example directly,
because we happen to have measured the one variable, `severity`, that
was confounding the naive comparison. If we compute the same treated-
minus-untreated contrast *separately within each level of severity*,
using `group_by(severity)`, we are conditioning on the confounder
exactly the way conditional exchangeability requires:

```{r cond-exch}
by_severity <- truth |>
  group_by(severity) |>
  summarise(
    naive_within = mean(y[a == 1]) - mean(y[a == 0]),
    .groups = "drop"
  )
by_severity
```

```{r cond-exch-check, include=FALSE}
weighted_avg <- with(
  truth |> count(severity) |> mutate(w = n / sum(n)) |>
    left_join(by_severity, by = "severity"),
  sum(w * naive_within)
)
```

Within each level of `severity`, the treated-versus-untreated contrast
comes out close to `r sprintf("%.3f", sim_summary$ate)` — the true ATE —
rather than the badly biased `r sprintf("%.3f", sim_summary$naive)` we
got by ignoring severity altogether. Averaging these two
severity-specific contrasts back together, weighted by how many
patients fall in each severity stratum, recovers
`r sprintf("%.3f", weighted_avg)`, essentially the ATE. Nothing about the
data changed; we only stopped comparing all treated patients to all
untreated patients, and started comparing treated to untreated patients
*within* the variable that was confounding the comparison. This
hand-computed stratification is the seed of standardization, which
Chapter 7 develops into a general tool for covariate sets far too large
to stratify by hand.

**Positivity** (or overlap) says that every combination of covariates
that occurs in the population has a nonzero probability of receiving
either treatment level: $0 < P(A = 1 \mid L = l) < 1$ for every value
$l$ of $L$ with positive probability. This is a logical precondition,
not a statistical nicety — if every patient with `severity == 1` in our
simulation had probability zero of getting the nonselective drug, there
would be no untreated severe patients to compare to, and no amount of
adjustment could recover what their outcome under nonselective
treatment would have been. Positivity can fail structurally, when a
covariate combination makes one treatment clinically unthinkable, or
only practically, when the combination is simply rare in a finite
sample; the second, less absolute failure is common enough in real
cohorts like `ns` that later chapters return to it when diagnosing
propensity score overlap.

Together, consistency, exchangeability (or its conditional version),
and positivity are what let us write an equals sign between a causal
estimand defined on unobservable potential outcomes and a statistical
quantity computable from an observed data set like `ns`. Every
adjustment method in the rest of this book — standardization, propensity
scores, weighting — is a different computational strategy for the same
identified quantity; none of them can substitute for these three
conditions actually holding.

## Exercises

1. Modify the `sim-po` simulation so that the Cox-2 drug is *harmful*
   rather than beneficial — for instance, swap the coefficients so `y1`
   has a higher baseline risk than `y0` — and recompute the ATE, ATT,
   and naive contrast. Does the naive estimate still point in the wrong
   direction, and if so, which way?
2. Starting from the original simulation, change how `a` is generated so
   that treatment no longer depends on `severity` at all (for example,
   `rbinom(100000, 1, 0.4)`, a fixed probability for everyone). Recompute
   the three summary statistics. What happens to the gap between the
   naive contrast and the ATE, and why does that make sense in light of
   the exchangeability condition?
3. Modify `y1` so that the treatment effect itself differs by
   `severity` — for example, make the drug strongly protective for
   severe patients but do nothing for non-severe patients. Recompute the
   ATE and the ATT. Are they still close to each other? Relate what you
   find to the discussion of when ATE and ATT coincide, and to which
   estimand a clinician deciding whether to keep prescribing the drug to
   her current, already-treated patients should care about most.
