# Potential outcomes and causal effects

```{r setup-ch03, include=FALSE}
source("R/setup.R")
```

Chapter 1 asked whether Cox-2 selective NSAIDs cause less peptic ulcer
disease than nonselective NSAIDs, and then showed that the obvious way of
answering the question, comparing crude risks between the two groups,
does not actually answer it, because Cox-2 initiators and nonselective
initiators differ in ways connected to ulcer risk before either drug is
ever taken. In this chapter we build the vocabulary needed to say
precisely what went wrong in that comparison, and what an answer to the
question would have to look like. The framework is called **potential
outcomes**, and it allows a causal effect to be defined as a specific
quantity that we can write down, simulate in code, and, under conditions
we will state carefully, estimate from data.

## Two futures for every patient

Picture a single patient in the `ns` cohort — call her Patient 47 — who
shows up at her physician's office with joint pain that warrants an
NSAID. Her physician has a choice: prescribe a nonselective NSAID, or
prescribe a Cox-2 selective inhibitor. Whichever one is chosen, Patient
47 either does or does not go on to develop peptic ulcer disease over
the following year. Although only one prescription is written, the
question of what would have happened under the other still has a
sensible answer. If she was in fact given the Cox-2 drug and stayed
ulcer-free, we can ask coherently what would have happened to her, with
everything else about her held fixed, had she instead received the
nonselective drug.

Potential outcomes provide notation for that question. Let $A$
denote the treatment Patient 47 receives, coded $A = 1$ for the Cox-2
selective drug and $A = 0$ for the nonselective drug, and let $Y$ denote
whether she develops incident PUD. Then $Y(1)$ is the outcome she would
have if she were treated with $A = 1$, and $Y(0)$ is the outcome she
would have if she were treated with $A = 0$. These are **potential**
outcomes, one for each treatment value she could in principle receive,
and both are well-defined quantities before we know which treatment she
actually gets. Once treatment is assigned, one of them becomes the
**observed** outcome and the other becomes counterfactual: contrary to
the fact of what happened. If Patient 47 actually receives the Cox-2
drug, we observe $Y(1)$, and $Y(0)$ is what her chart would show in the
world where her physician wrote the other prescription.

A note on notation is in order, since conventions differ across sources.
We write these potential outcomes as $Y(1)$ and $Y(0)$, matching the PHS
921 lectures and exams, whereas many books, including *Causal Inference:
What If*, use superscripts, $Y^1$ and $Y^0$ or $Y^a$, and others write
$Y_a$ for the outcome under a generic treatment value $a$. All three
notations denote the same objects, and it is worth becoming comfortable
moving between them rather than attaching the concept to any one of
them.

With both potential outcomes in hand for a single patient, her
**individual causal effect** of Cox-2 selectivity on PUD is the contrast
between them, $Y(1) - Y(0)$. If $Y(1) = 0$ and $Y(0) = 1$, the Cox-2
drug prevented her ulcer; if both are 0, or both are 1, the choice of
drug made no difference to her outcome, even though the drugs may differ
substantially for other patients. Although this quantity is well
defined, for any real patient it is unobservable. We only ever see one of
Patient 47's two potential outcomes, because she can only actually take
one drug. This is what Holland (1986) called the **fundamental problem of
causal inference**: for any individual, at most one potential outcome is
ever realized, and the individual causal effect can never be computed
from data on that individual alone, no matter how much we know about her
or how long we follow her. Every method in this book is therefore a
strategy for working around this problem. None of them recovers the
missing potential outcome for any one patient; instead, they use groups
of patients to estimate causal effects that are well defined on average,
although they remain unknowable one person at a time.

## Simulating both potential outcomes

The fundamental problem of causal inference is a permanent limitation of
real data: no data set on real patients ever contains both of anyone's
potential outcomes. In a simulated population, however, we are the ones
writing the code, and so we decide both potential outcomes for every
patient and can then observe what happens when only one of them is
recorded. Such a simulation reproduces the causal question Chapter 1
raised, and it lets us watch confounding arise mechanically rather than
merely describe it in words.

Imagine 100,000 patients with a severity indicator, `severity`, that
plays the role of an unmeasured driver of both treatment and outcome,
capturing whatever led some patients toward the Cox-2 drug and also
independently raised their ulcer risk. For each patient we generate
*both* potential outcomes, `y0` (the outcome under nonselective NSAID)
and `y1` (the outcome under Cox-2 selective drug), with the Cox-2
potential outcome built to carry a lower baseline risk, so that the drug
is genuinely protective in this simulated world. We then assign actual
treatment `a` so that more severe patients are more likely to receive the
Cox-2 drug, which is the channeling Chapter 1 described, and construct
the single outcome `y` actually recorded: the potential outcome
matching the treatment each patient actually got.

```{r sim-po}
set.seed(921)
truth <- tibble(
  severity = rbinom(100000, 1, 0.3),                       # confounder
  y0 = rbinom(100000, 1, 0.05 + 0.15 * severity),          # risk untreated
  y1 = rbinom(100000, 1, 0.03 + 0.15 * severity),          # risk treated
  a  = rbinom(100000, 1, 0.2 + 0.5 * severity),            # treated more if severe
  y  = if_else(a == 1, y1, y0)                             # consistency
)
truth |> summarise(
  ate   = mean(y1) - mean(y0),
  att   = mean(y1[a == 1]) - mean(y0[a == 1]),
  naive = mean(y[a == 1]) - mean(y[a == 0])
)
```

```{r sim-po-values, include=FALSE}
sim_summary <- truth |> summarise(
  ate   = mean(y1) - mean(y0),
  att   = mean(y1[a == 1]) - mean(y0[a == 1]),
  naive = mean(y[a == 1]) - mean(y[a == 0])
)
```

Because we generated `y0` and `y1` ourselves, we can compute the true
average causal effect directly from columns no real data set would ever
have both of at once: `mean(y1) - mean(y0)` is
`r sprintf("%.3f", sim_summary$ate)`, a negative number, confirming that
treatment is on average beneficial in this simulated population, lowering
the risk of the outcome by about
`r sprintf("%.1f", -100 * sim_summary$ate)` percentage points. The last
line of the summary, `naive`, is the only contrast we could ever compute
from a real observational data set, being the mean of the *observed*
outcome `y` among the treated minus the mean of `y` among the untreated,
and it comes out to `r sprintf("%.3f", sim_summary$naive)`, positive and
on the wrong side of zero. The naive comparison therefore suggests that
Cox-2 selectivity is harmful, although we built the simulation so that it
is beneficial. The gap between `r sprintf("%.3f", sim_summary$naive)`
and `r sprintf("%.3f", sim_summary$ate)` is confounding by `severity`,
visible here only because in this artificial world we know how the data
were generated: severe patients were both more likely to be treated
*and* more likely to have a bad outcome regardless of treatment, so the
treated group looks worse than it would have if severity had been
balanced across the two groups. Real data never provide `y0` and `y1`
side by side, and the remainder of the chapter takes up the conditions
under which the gap can nonetheless be closed.

## Average causal effects and estimands

The individual causal effect $Y(1) - Y(0)$ cannot be observed, although
its *average* over a population can be estimated once we have the right
tools, which is why causal inference is framed in terms of
population-level summaries called **estimands**. The one that shows up
most often in this book is
the **average treatment effect (ATE)**,
$$\text{ATE} = E[Y(1)] - E[Y(0)],$$
the average, over everyone in the population of interest, of what would
happen if everyone were treated minus what would happen if no one were.
In the simulation above, `mean(y1) - mean(y0)` computed over all 100,000
patients *is* the ATE, because we happened to have both potential
outcomes for every one of them.

A second estimand, the **average treatment effect in the treated
(ATT)**,
$$\text{ATT} = E[Y(1) \mid A = 1] - E[Y(0) \mid A = 1],$$
restricts that same contrast to the patients who were actually treated:
it asks what treatment did for the people who received it, not for the
population as a whole. In the simulation, that is
`mean(y1[a == 1]) - mean(y0[a == 1])`, and in our output it comes out
very close to, but not exactly equal to, the ATE. That small remaining
gap is Monte Carlo noise, not confounding: in this particular
simulation the treatment effect `y1 - y0` is built to be the same
constant regardless of `severity`, so the *true* ATT equals the *true*
ATE exactly, and the two only look different because both are estimated
from a single finite sample of 100,000 patients rather than computed on
the population itself. Rerunning with a different seed would shift the
gap around, and often shrink it, without ever making it disappear
completely.

The two estimands diverge genuinely, rather than merely by chance, when
the *effect itself* varies with whatever determines who gets treated.
Which estimand is wanted then depends on the question being asked. If a
health system is deciding whether to switch every future NSAID
prescription for arthritis patients to the Cox-2 selective drug, the ATE
is the relevant quantity, since it describes what would happen if
everyone were treated. If a clinician is instead trying to understand the
effect of the prescriptions she has already written, on the patients who
actually received them, the ATT is closer to what she wants, since it
describes whether that decision benefited the patients who received it.
The two estimands coincide only when treatment
effects do not vary with whatever determines who gets treated, which is
the case built into this simulation; Exercise 3 asks you to relax that
condition by making the effect depend on `severity`, so that the ATE and
ATT genuinely differ.

Any of these estimands can be reported on more than one scale. With $Y$
binary, as PUD is, the most direct scale is the **risk difference**,
$E[Y(1)] - E[Y(0)]$, interpretable as "how many fewer (or more) cases per
100 patients treated." The **risk ratio**, $E[Y(1)] / E[Y(0)]$, expresses
the same contrast multiplicatively, so that a risk ratio of 0.7 means
treatment reduces risk by 30% regardless of baseline risk, which often
makes it more portable across populations than a difference. The **odds
ratio**,
$$\frac{E[Y(1)]/(1-E[Y(1)])}{E[Y(0)]/(1-E[Y(0)])},$$
is the scale most familiar from logistic regression and case-control
studies, but it approximates the risk ratio only when the outcome is
rare, and can otherwise diverge from it substantially. None of the
identification conditions we turn to next, consistency, exchangeability,
and positivity, depends on which scale is eventually reported. The choice
of scale is a separate decision, made after identification, about how
best to communicate a causal contrast.

## The identification conditions

Estimands like the ATE are defined in terms of potential outcomes we
never fully observe. **Identification** is the step of showing that,
under stated assumptions, an estimand written in potential outcomes is
equal to some quantity we actually could compute from observed data.
Three conditions do this work, and we state each one in terms of the
NSAID cohort so that they read as claims about real patients.

**Consistency** says that a patient's observed outcome equals the
potential outcome under the treatment she actually received: if Patient
47 takes the Cox-2 drug, her observed $Y$ equals $Y(1)$, not some
average or blend involving $Y(0)$. Although this sounds almost too
obvious to state, it is doing real work, since it is what licenses the
line `y = if_else(a == 1, y1, y0)` in the simulation above, and it can
fail in practice when "treatment" is not well defined enough to pin down
a single potential outcome. If "Cox-2 initiation" in fact bundles several
different doses, durations, or concurrent gastroprotective strategies,
then $Y(1)$ refers to several distinct quantities rather than one, and
consistency fails for reasons that have nothing to do with statistical
estimation.

**Exchangeability** (also called ignorability) says that the treated
and untreated groups would have had the same average outcome had they
received the same treatment — formally, $Y(1), Y(0) \perp A$, potential
outcomes are statistically independent of the treatment actually
received. This is the condition that fails in our simulation: because `a`
depends on `severity`, and `severity` also drives `y0` and `y1`, treated
and untreated patients are not exchangeable, which is why the naive
contrast diverged from the ATE. A randomized trial delivers
exchangeability by construction, as Chapter 4 makes precise, whereas
confounding by indication removes it in observational data like `ns`,
where physicians route sicker-looking patients toward the Cox-2 drug for
reasons connected to the very outcome we are studying.

Full exchangeability is usually too much to hope for in observational
data, which is why most of this book works with a weaker version,
**conditional exchangeability**: $Y(1), Y(0) \perp A \mid L$ for some set
of measured covariates $L$. Treatment need not be independent of
potential outcomes overall, only *within levels* of $L$, so that among
patients who share the same measured `severity`, say, treatment
assignment is as good as random. This is a considerably more plausible
assumption whenever we can measure the variables driving treatment
choice, and it is the assumption that standardization, propensity scores,
and IPTW in later chapters are built to exploit.

Conditional exchangeability can be seen to work in our simulated example,
since we have measured the one variable, `severity`, that was confounding
the naive comparison. If we compute the same treated-minus-untreated
contrast *separately within each level of severity*, using
`group_by(severity)`, we are conditioning on the confounder in the way
conditional exchangeability requires:

```{r cond-exch}
by_severity <- truth |>
  group_by(severity) |>
  summarise(
    naive_within = mean(y[a == 1]) - mean(y[a == 0]),
    .groups = "drop"
  )
by_severity
```

```{r cond-exch-check, include=FALSE}
weighted_avg <- with(
  truth |> count(severity) |> mutate(w = n / sum(n)) |>
    left_join(by_severity, by = "severity"),
  sum(w * naive_within)
)
```

Within each level of `severity`, the treated-versus-untreated contrast
comes out close to `r sprintf("%.3f", sim_summary$ate)`, the true ATE,
rather than the badly biased `r sprintf("%.3f", sim_summary$naive)` we
obtained by ignoring severity altogether. Averaging these two
severity-specific contrasts back together, weighted by how many
patients fall in each severity stratum, recovers
`r sprintf("%.3f", weighted_avg)`, essentially the ATE. The data
themselves are unchanged; the only difference is that treated and
untreated patients are now being compared *within* levels of the variable
that was confounding the comparison. This hand-computed stratification is
a simple case of standardization, which Chapter 7 develops into a general
tool for covariate sets far too large to stratify by hand.

**Positivity** (or overlap) says that every combination of covariates
that occurs in the population has a nonzero probability of receiving
either treatment level: $0 < P(A = 1 \mid L = l) < 1$ for every value
$l$ of $L$ with positive probability. Positivity is a logical
precondition rather than a technical convenience. If every patient with
`severity == 1` in our simulation had probability zero of receiving the
nonselective drug, there would be no untreated severe patients available
for comparison, and no amount of adjustment could recover what their
outcome under nonselective treatment would have been. Positivity can fail
structurally, when a covariate combination makes one treatment clinically
unthinkable, or only practically, when the combination is merely rare in
a finite sample. The second, less absolute failure is common enough in
real cohorts like `ns` that later chapters return to it when diagnosing
propensity score overlap.

Together, consistency, exchangeability (or its conditional version),
and positivity are what let us write an equals sign between a causal
estimand defined on unobservable potential outcomes and a statistical
quantity computable from an observed data set like `ns`. The adjustment
methods in the rest of this book, including standardization, propensity
scores, and weighting, are different computational strategies for the
same identified quantity, and none of them can substitute for these three
conditions holding.

## Exercises

1. Modify the `sim-po` simulation so that the Cox-2 drug is *harmful*
   rather than beneficial — for instance, swap the coefficients so `y1`
   has a higher baseline risk than `y0` — and recompute the ATE, ATT,
   and naive contrast. Does the naive estimate still point in the wrong
   direction, and if so, which way?
2. Starting from the original simulation, change how `a` is generated so
   that treatment no longer depends on `severity` at all (for example,
   `rbinom(100000, 1, 0.4)`, a fixed probability for everyone). Recompute
   the three summary statistics. What happens to the gap between the
   naive contrast and the ATE, and why does that make sense in light of
   the exchangeability condition?
3. Modify `y1` so that the treatment effect itself differs by
   `severity` — for example, make the drug strongly protective for
   severe patients but do nothing for non-severe patients. Recompute the
   ATE and the ATT. Are they still close to each other? Relate what you
   find to the discussion of when ATE and ATT coincide, and to which
   estimand a clinician deciding whether to keep prescribing the drug to
   her current, already-treated patients should care about most.
