# The randomized trial as the benchmark

```{r setup-ch04, include=FALSE}
source("R/setup.R")
```

Chapter 3 named exchangeability as the condition that made the naive
comparison in Chapter 1 fail: Cox-2 initiators and nonselective
initiators were not exchangeable, because physicians routed patients
toward one drug or the other based on characteristics connected to
ulcer risk. That chapter also promised, in passing, that a randomized
trial buys exchangeability "by construction," rather than by an
assumption you have to defend with subject-matter argument. This
chapter cashes in that promise. It shows, first in code and then in
words, why handing the treatment decision to a random number generator
rather than to a physician's judgment produces two groups that are, on
average, alike in every way that matters — known confounders,
unmeasured ones, and everything in between. It then works out what a
randomized trial can and cannot still promise once patients stop doing
exactly what they were assigned to do, through the distinction between
intention-to-treat and per-protocol analysis. It closes with the idea
that organizes the rest of this book: the target trial, the discipline
of writing down the trial you wish you could run before you try to
emulate it with the data you actually have.

## Why randomization works

Confounding by indication, as Chapter 1 described it, arises because
treatment assignment is a function of patient characteristics connected
to the outcome: a physician who suspects a patient is at elevated
gastrointestinal risk is more likely to reach for the Cox-2 selective
drug for exactly that patient, so Cox-2 initiators end up looking sicker
on the very dimension the drug is meant to protect. Randomization
attacks this at its source by making assignment a function of *nothing*
about the patient at all — not age, not arthritis history, not anything
a physician can see, and, crucially, not anything a physician cannot
see either. That last point is what makes randomization strictly more
powerful than any adjustment strategy in this book: matching,
standardization, and propensity scores can only balance covariates
someone thought to measure, while a coin flip balances everything,
including confounders no one ever recorded, because the coin does not
consult any characteristic of the patient, known or unknown, before
it lands. Formally, when $A$ is generated by a physical randomization
device rather than by clinical judgment, $A$ is statistically
independent of baseline covariates $L$ and of both potential outcomes
$Y(1), Y(0)$ by the mechanics of assignment — the exchangeability
condition from Chapter 3, delivered as a design guarantee rather than
an assumption.

We do not have an actual randomized trial of Cox-2 selectivity — only
the `ns` cohort, in which treatment was a physician's choice. But we can
ask a narrower question of the data we do have: what would the Table 1
comparison and the crude risk comparison look like if the label
"Cox-2 initiator" had been handed out by a coin flip instead of by a
physician? `mutate(a_rand = sample(cox2_init))` answers exactly that —
it shuffles which of the `r nrow(ns)` rows get called "1," so that the
same `r sum(ns$cox2_init)` Cox-2 labels in the data land on that many
patients chosen completely at
random, while every actual covariate and every actual outcome stays
exactly as recorded. This is a relabeling exercise, not a new trial:
nobody's peptic ulcer disease status changes, because `pud` is not
regenerated. The only thing the random draw decides is which of the
real patients we call "arm 1" versus "arm 0" for the purpose of the
comparison that follows.

```{r real-imbalance}
source("R/table1.R")
table1_categorical(
  ns,
  vars = c("sex", "arthritis", "hypertension", "diabetes"),
  by = "cox2_init"
) |> knitr::kable(
  caption = "Table 1 under the real, physician-chosen assignment (0 = nonselective, 1 = Cox-2 selective)"
)
```

```{r rand-demo}
set.seed(8)
ns_rand <- ns |> mutate(a_rand = sample(cox2_init))
table1_categorical(
  ns_rand,
  vars = c("sex", "arthritis", "hypertension", "diabetes"),
  by = "a_rand"
) |> knitr::kable(
  caption = "Table 1 under the same patients, with the Cox-2 label reassigned by a coin flip (0/1 = random label)"
)
```

```{r gap-values, include=FALSE}
gap_pp <- function(data, var, group) {
  data |>
    group_by(.data[[group]]) |>
    summarise(p = mean(.data[[var]] == "Yes"), .groups = "drop") |>
    pull(p) |>
    diff() |>
    abs() |>
    (\(x) x * 100)()
}
arth_real <- gap_pp(ns, "arthritis", "cox2_init")
arth_rand <- gap_pp(ns_rand, "arthritis", "a_rand")
arth_pct <- ns |>
  group_by(cox2_init) |>
  summarise(p = 100 * mean(arthritis == "Yes"), .groups = "drop") |>
  pull(p)
arth_pct_0 <- arth_pct[1]
arth_pct_1 <- arth_pct[2]
htn_real  <- gap_pp(ns, "hypertension", "cox2_init")
htn_rand  <- gap_pp(ns_rand, "hypertension", "a_rand")
diab_real <- gap_pp(ns, "diabetes", "cox2_init")
diab_rand <- gap_pp(ns_rand, "diabetes", "a_rand")
```

Look at arthritis first: under the real, physician-chosen assignment
the gap between groups is `r sprintf("%.0f", arth_real)` percentage
points (`r sprintf("%.1f", arth_pct_1)`% among Cox-2 initiators versus
`r sprintf("%.1f", arth_pct_0)`% among nonselective
initiators) — large, and exactly the kind of systematic imbalance
Chapter 2 already flagged when it introduced this Table 1. Under the
randomly permuted label, the same comparison shrinks to
`r sprintf("%.1f", arth_rand)` points. Diabetes tells the same story:
a real gap of `r sprintf("%.1f", diab_real)` points falls to
`r sprintf("%.1f", diab_rand)` under random assignment. Hypertension is
the instructive exception: its gap under the random label happens to
come out to `r sprintf("%.1f", htn_rand)` points, identical to the real
gap of `r sprintf("%.1f", htn_real)`. That coincidence is not evidence
that hypertension is somehow still confounded after randomization —
it is a reminder that any single random split is itself only one draw
among many possible draws, and one draw is not guaranteed to look
perfectly balanced on every variable, purely by chance. What
randomization guarantees is that this kind of mismatch is *unsystematic*:
rerun the shuffle with a different seed and the hypertension coincidence
disappears, while arthritis and diabetes keep coming back close to
balanced, because "close to balanced" is what coin flips deliver on
average, not what any one flip is obligated to produce.

The same logic applies to the outcome itself, and it is the sharper
point. Because `pud` was never regenerated, the treated and untreated
groups defined by `a_rand` are being compared on outcomes that already
happened, for reasons that had nothing to do with `a_rand`.

```{r rand-outcome}
contrasts <- bind_rows(
  ns |> summarise(
    assignment = "real (cox2_init)",
    risk_0 = mean(pud[cox2_init == 0]),
    risk_1 = mean(pud[cox2_init == 1])
  ),
  ns_rand |> summarise(
    assignment = "randomly permuted (a_rand)",
    risk_0 = mean(pud[a_rand == 0]),
    risk_1 = mean(pud[a_rand == 1])
  )
) |> mutate(diff = risk_1 - risk_0)
contrasts |> knitr::kable(digits = 4)
```

```{r rand-values, include=FALSE}
real_diff <- contrasts$diff[1]
rand_diff <- contrasts$diff[2]
```

The real crude contrast reproduces Chapter 1's finding: Cox-2
initiators have a `r sprintf("%.1f", 100 * real_diff)`-point higher
crude risk of PUD than nonselective initiators, the wrong-signed,
confounded comparison this whole book exists to fix. The randomly
permuted contrast, computed from the identical set of recorded `pud`
values, comes out to `r sprintf("%.2f", 100 * rand_diff)` points —
indistinguishable from zero. Nothing about any patient's actual outcome
changed between the two rows of that table; the only thing that changed
is which label we used to split the data into two groups. When the
label carries no information about who received it — which is exactly
what a coin flip guarantees — the crude contrast is an honest, unbiased
estimate of nothing, and it finds nothing. When the label is a
physician's judgment instead, correlated with everything that judgment
was trying to account for, the same arithmetic turns up a gap that has
nothing to do with the drug. This is why the randomized trial is called
the *benchmark*: not because any one trial is guaranteed a perfectly
matched Table 1 — chance imbalance, like the hypertension coincidence
above, can and does happen — but because whatever mismatch remains is
due to chance alone, the kind of uncertainty ordinary confidence
intervals are built to quantify. Confounding by indication is a
different kind of error entirely; no amount of additional data makes it
smaller, because it is not noise. Every design and every adjustment
method in the rest of this book is judged by how close it comes to the
zero this section produced for free.

## Intention-to-treat and per-protocol

A coin flip decides the arm a patient is assigned to; it does not
decide what that patient actually does afterward. In a real Cox-2 trial,
some patients randomized to the Cox-2 arm would stop the drug early
because of side effects or cost, some randomized to nonselective NSAIDs
would switch to Cox-2 therapy if their joint pain persisted, and some in
either arm would take over-the-counter NSAIDs the protocol never
intended to capture. Once assignment and actual treatment can diverge,
a trial supports (at least) two different analyses, and they answer two
different questions.

The **intention-to-treat (ITT)** analysis compares patients according to
the arm they were *randomized* to, regardless of what treatment they
actually went on to receive. Writing $A$ for randomized assignment, the
ITT contrast is $E[Y \mid A = 1] - E[Y \mid A = 0]$, computed on the
groups the randomization device itself created. Because $A$ here is
literally the coin flip, ITT preserves exchangeability exactly as
constructed — the comparison remains a contrast between groups that
differ, on average, only by that coin flip, no matter how much
nonadherence occurs afterward. This is why ITT is the default primary
analysis in essentially every large randomized trial: it is the one
contrast that inherits randomization's guarantee without any further
qualification.

That guarantee comes at a cost of interpretation. ITT does not estimate
the effect of *taking* the assigned drug; it estimates the effect of
being *assigned* to a strategy, which is a blend of the effect among
patients who followed it and the effect among patients who did not.
When nonadherence is common, ITT effects are typically attenuated
toward the null relative to the effect of full adherence — a drug that
would clearly help if everyone took it as directed can produce a modest
or null ITT result if enough assigned patients quietly stopped taking
it. The **per-protocol** analysis tries to recover the sharper question
by restricting to, or reclassifying patients according to, what they
actually did — comparing adherers to the Cox-2 strategy against
adherers to the nonselective strategy, say, rather than comparing the
arms as randomized. This is often the clinically more interesting
question, but the moment patients are grouped by what they actually did
rather than by what the coin flip assigned them, the randomization at
baseline stops protecting the comparison. Whatever led a given patient
to adhere rather than switch or discontinue — symptom severity, general
health, health literacy, tolerance of side effects — is exactly the kind
of characteristic that can also be connected to the outcome, which is
the same mechanism Chapter 1 called the healthy-user effect, now showing
up *inside* a randomized trial rather than across an observational
cohort. A naive per-protocol comparison can therefore be confounded even
though the trial that produced the data was, at the point of
randomization, perfectly randomized.

The `sta` cohort makes the stakes concrete without needing a real trial
at all. `sta` is observational — nobody was randomized to start a
statin — so any comparison of patients who initiated and *stayed on* a
statin for some minimum duration against everyone else is already
per-protocol in spirit, and already vulnerable to the same
self-selection just described. It is also vulnerable to something
subtler: misclassifying the follow-up time before a patient meets the
"adherer" criterion can quietly build survival into the definition of
one of the groups, a distortion called immortal time bias that Chapter 6
introduces on its own terms. Chapter 12 develops the clone-censor-weight
approach specifically to answer a per-protocol-style question — what
would happen if patients followed a specific treatment strategy for its
full duration — using observational data like `sta`, without letting
either kind of bias back in.

## The target trial

Every method in this book adjusts observational data to stand in for
something it is not: a randomized trial. Hernán and Robins (2016)
turned that observation into a discipline, called the **target trial**
framework:
for any causal question you want to answer with observational data,
first specify, in full protocol detail, the randomized trial you wish
you could run to answer it — as though funding, ethics, and recruitment
were not obstacles — and only then describe how each piece of that
protocol will be *emulated* using the data you actually have. Almost
none of the trials specified this way will ever be run. The value of
writing the protocol anyway is that it forces every design decision a
trial would have to make — who is eligible, what exactly counts as each
treatment strategy, when follow-up starts, what counts as the outcome —
into the open, before any adjustment method is chosen, rather than
leaving those decisions implicit in a modeling script.

Here is that protocol, written out for the Cox-2 question this book has
been asking since Chapter 1:

| Protocol component | Target trial | Emulation using `ns` |
|---|---|---|
| Eligibility | Adults beginning NSAID therapy, with no NSAID use in the preceding period | Same criterion, applied when `ns` was assembled: new users only |
| Treatment strategies | (1) Initiate a Cox-2 selective NSAID; (2) initiate a nonselective NSAID | `cox2_init == 1` versus `cox2_init == 0` at the qualifying prescription |
| Assignment | Random allocation to strategy (1) or (2) | Not randomized; approximated by conditioning on measured confounders (`ns_covs`) under the conditional exchangeability assumption from Chapter 3 |
| Follow-up | Begins at randomization ("time zero") and continues to the outcome, loss to follow-up, or a fixed administrative end | Begins at the qualifying prescription date, ends at incident PUD, loss to follow-up, or the administrative end |
| Outcome | Incident peptic ulcer disease | `pud == 1` observed during follow-up |
| Causal contrast | Intention-to-treat comparison of the two assigned strategies | ITT analog: compare all patients who initiated each strategy at time zero, regardless of later switching or discontinuation |
| Analysis plan | Risk difference or risk ratio between arms, estimated directly from randomized groups | Same contrast, estimated with an adjustment method (standardization, propensity scores, or IPTW — Chapters 7 through 9) that stands in for randomization |

Six of these seven rows can already be written down for `ns` with no
controversy, in the same language a trial protocol would use: the
eligibility criteria are exactly how new users were identified, the two
strategies are already coded in `cox2_init`, follow-up already has a
defined start and end, the outcome is already defined as `pud`, and
even the causal contrast and analysis plan can be stated without
needing to resolve anything about confounding first. Only the
**Assignment** row cannot be filled in as "randomized," because it was
not — and that single row is exactly the site where the rest of this
book does its work. Standardization and outcome regression, propensity
scores and inverse probability weighting, instrumental variables and
sensitivity analysis: every one of these is, at bottom, a different
technical strategy for filling in that one row of the table with
something that behaves enough like a coin flip to make the causal
contrast in the last row trustworthy. Chapters 6, 10, and 12 return to
a version of this same protocol — adapted for study design choices, for
survival outcomes, and for the dynamic treatment strategies studied in
`sta` — but the shape of the argument never changes: specify the trial
first, then argue, row by row, for how well the observational data
emulates it.

## Exercises

1. Draft a target trial protocol table, in the same seven-row format
   used above, for the statin question this book has been asking
   alongside the Cox-2 question. Use the `sta` cohort and `sta_covs` to
   fill in eligibility, treatment strategies, assignment, follow-up,
   outcome, causal contrast, and analysis plan in your own words.
2. Immortal time bias was named in this chapter as a hazard for
   per-protocol-style comparisons in `sta`, and Chapter 6 takes it up in
   full. Looking at the seven rows of the target trial table above,
   which single protocol component does immortal time bias violate, and
   why does misclassifying a patient's follow-up before they meet a
   treatment-defining criterion break that component specifically,
   rather than, say, the eligibility or outcome rows?
