# The randomized trial as the benchmark

```{r setup-ch04, include=FALSE}
source("R/setup.R")
```

Chapter 3 named exchangeability as the condition that made the naive
comparison in Chapter 1 fail: Cox-2 initiators and nonselective
initiators were not exchangeable, because physicians routed patients
toward one drug or the other based on characteristics connected to
ulcer risk. That chapter also promised, in passing, that a randomized
trial buys exchangeability "by construction," rather than by an
assumption you have to defend with subject-matter argument. In this
chapter we make good on that promise, first in code and then in words,
by showing why handing the treatment decision to a random number
generator rather than to a physician's judgment produces two groups that
are, on average, alike with respect to both measured and unmeasured
characteristics. We then consider what a randomized trial can and cannot
still promise once patients stop doing what they were assigned to do,
using the distinction between intention-to-treat and per-protocol
analysis. Finally, we introduce the idea that organizes the rest of this
book, the target trial, which is the discipline of writing down the trial
one wishes one could run before attempting to emulate it with the data
actually available.

## Why randomization works

Confounding by indication, as Chapter 1 described it, arises because
treatment assignment is a function of patient characteristics connected
to the outcome: a physician who suspects a patient is at elevated
gastrointestinal risk is more likely to reach for the Cox-2 selective
drug for exactly that patient, so Cox-2 initiators end up looking sicker
on the very dimension the drug is meant to protect. Randomization
addresses this at its source by making assignment a function of nothing
about the patient at all, neither age nor arthritis history nor anything
else a physician can observe, and, equally important, nothing a physician
cannot observe. It is this last property that makes randomization more
powerful than any adjustment strategy in this book. Matching,
standardization, and propensity scores can balance only those covariates
someone thought to measure, whereas a coin flip balances everything,
including confounders no one ever recorded, because the assignment
mechanism does not consult any characteristic of the patient, known or
unknown. Formally, when $A$ is generated by a physical randomization
device rather than by clinical judgment, $A$ is statistically
independent of baseline covariates $L$ and of both potential outcomes
$Y(1), Y(0)$ by the mechanics of assignment, so that the exchangeability
condition from Chapter 3 is delivered as a feature of the design rather
than assumed.

We do not have an actual randomized trial of Cox-2 selectivity, only the
`ns` cohort, in which treatment was a physician's choice. We can,
however, ask a narrower question of the data we do have: what would the
Table 1 comparison and the crude risk comparison look like if the label
"Cox-2 initiator" had been handed out by a coin flip instead of by a
physician? `mutate(a_rand = sample(cox2_init))` answers that question by
shuffling which of the `r nrow(ns)` rows get called "1," so that the
same `r sum(ns$cox2_init)` Cox-2 labels in the data land on that many
patients chosen completely at
random, while every actual covariate and every actual outcome stays
exactly as recorded. This is a relabeling exercise, not a new trial:
nobody's peptic ulcer disease status changes, because `pud` is not
regenerated. The only thing the random draw decides is which of the
real patients we call "arm 1" versus "arm 0" for the purpose of the
comparison that follows.

```{r real-imbalance}
source("R/table1.R")
table1_categorical(
  ns,
  vars = c("sex", "arthritis", "hypertension", "diabetes"),
  by = "cox2_init"
) |> knitr::kable(
  caption = "Table 1 under the real, physician-chosen assignment (0 = nonselective, 1 = Cox-2 selective)"
)
```

```{r rand-demo}
set.seed(8)
ns_rand <- ns |> mutate(a_rand = sample(cox2_init))
table1_categorical(
  ns_rand,
  vars = c("sex", "arthritis", "hypertension", "diabetes"),
  by = "a_rand"
) |> knitr::kable(
  caption = "Table 1 under the same patients, with the Cox-2 label reassigned by a coin flip (0/1 = random label)"
)
```

```{r gap-values, include=FALSE}
gap_pp <- function(data, var, group) {
  data |>
    group_by(.data[[group]]) |>
    summarise(p = mean(.data[[var]] == "Yes"), .groups = "drop") |>
    pull(p) |>
    diff() |>
    abs() |>
    (\(x) x * 100)()
}
arth_real <- gap_pp(ns, "arthritis", "cox2_init")
arth_rand <- gap_pp(ns_rand, "arthritis", "a_rand")
arth_pct <- ns |>
  group_by(cox2_init) |>
  summarise(p = 100 * mean(arthritis == "Yes"), .groups = "drop") |>
  pull(p)
arth_pct_0 <- arth_pct[1]
arth_pct_1 <- arth_pct[2]
htn_real  <- gap_pp(ns, "hypertension", "cox2_init")
htn_rand  <- gap_pp(ns_rand, "hypertension", "a_rand")
diab_real <- gap_pp(ns, "diabetes", "cox2_init")
diab_rand <- gap_pp(ns_rand, "diabetes", "a_rand")
```

Look at arthritis first: under the real, physician-chosen assignment
the gap between groups is `r sprintf("%.0f", arth_real)` percentage
points (`r sprintf("%.1f", arth_pct_1)`% among Cox-2 initiators versus
`r sprintf("%.1f", arth_pct_0)`% among nonselective
initiators), which is large, and is the kind of systematic imbalance
Chapter 2 already flagged when it introduced this Table 1. Under the
randomly permuted label, the same comparison shrinks to
`r sprintf("%.1f", arth_rand)` points. Diabetes behaves similarly:
a real gap of `r sprintf("%.1f", diab_real)` points falls to
`r sprintf("%.1f", diab_rand)` under random assignment. Hypertension is
an instructive exception, in that its gap under the random label happens
to come out to `r sprintf("%.1f", htn_rand)` points, identical to the
real gap of `r sprintf("%.1f", htn_real)`. Rather than indicating that
hypertension remains confounded after randomization, this coincidence is
a reminder that any single random split is only one draw among many
possible draws, and that a single draw will not, by chance alone, look
perfectly balanced on every variable. What randomization guarantees is
that mismatches of this kind are *unsystematic*: if we rerun the shuffle
with a different seed, the hypertension coincidence disappears, while
arthritis and diabetes continue to come back close to balanced, since
approximate balance is what randomization delivers on average rather than
in any particular draw.

The same logic applies to the outcome itself. Because `pud` was never
regenerated, the treated and untreated groups defined by `a_rand` are
being compared on outcomes that already happened, for reasons that had
nothing to do with `a_rand`.

```{r rand-outcome}
contrasts <- bind_rows(
  ns |> summarise(
    assignment = "real (cox2_init)",
    risk_0 = mean(pud[cox2_init == 0]),
    risk_1 = mean(pud[cox2_init == 1])
  ),
  ns_rand |> summarise(
    assignment = "randomly permuted (a_rand)",
    risk_0 = mean(pud[a_rand == 0]),
    risk_1 = mean(pud[a_rand == 1])
  )
) |> mutate(diff = risk_1 - risk_0)
contrasts |> knitr::kable(digits = 4)
```

```{r rand-values, include=FALSE}
real_diff <- contrasts$diff[1]
rand_diff <- contrasts$diff[2]
```

The real crude contrast reproduces Chapter 1's finding: Cox-2
initiators have a `r sprintf("%.1f", 100 * real_diff)`-point higher
crude risk of PUD than nonselective initiators, the wrong-signed,
confounded comparison this whole book exists to fix. The randomly
permuted contrast, computed from the identical set of recorded `pud`
values, comes out to `r sprintf("%.2f", 100 * rand_diff)` points, which is
indistinguishable from zero. No patient's actual outcome changed between
the two rows of that table; the only difference is which label we used to
split the data into two groups. When the label carries no information
about who received it, as a coin flip guarantees, the crude contrast is
an unbiased estimate of a null effect, and it recovers one. When the
label is instead a physician's judgment, correlated with everything that
judgment was trying to account for, the same arithmetic turns up a gap
that has nothing to do with the drug. This is why the randomized trial is
described as the *benchmark*. The reason is not that any one trial is
guaranteed a perfectly matched Table 1, since chance imbalance of the
sort seen for hypertension above does occur, but rather that whatever
mismatch remains is due to chance alone, which is the kind of uncertainty
that ordinary confidence intervals are built to quantify. Confounding by
indication is a different kind of error, and no amount of additional data
makes it smaller, because it is not noise. Every design and every
adjustment method in the rest of this book is judged by how close it comes
to the null contrast that randomization produced in this section without
any adjustment at all.

## Intention-to-treat and per-protocol

A coin flip decides the arm a patient is assigned to; it does not
decide what that patient actually does afterward. In a real Cox-2 trial,
some patients randomized to the Cox-2 arm would stop the drug early
because of side effects or cost, some randomized to nonselective NSAIDs
would switch to Cox-2 therapy if their joint pain persisted, and some in
either arm would take over-the-counter NSAIDs the protocol never
intended to capture. Once assignment and actual treatment can diverge,
a trial supports (at least) two different analyses, and they answer two
different questions.

The **intention-to-treat (ITT)** analysis compares patients according to
the arm they were *randomized* to, regardless of what treatment they
actually went on to receive. Writing $A$ for randomized assignment, the
ITT contrast is $E[Y \mid A = 1] - E[Y \mid A = 0]$, computed on the
groups the randomization device itself created. Because $A$ is here the
coin flip, ITT preserves exchangeability as constructed, and the
comparison remains a contrast between groups that differ, on average,
only by that coin flip, however much nonadherence occurs afterward. For
this reason ITT is the default primary analysis in nearly every large
randomized trial, being the one contrast that inherits randomization's
guarantee without further qualification.

That guarantee, however, comes at a cost of interpretation, because ITT
does not estimate the effect of *taking* the assigned drug. It estimates
the effect of being *assigned* to a strategy, which is a blend of the
effect among patients who followed it and the effect among patients who
did not. When nonadherence is common, ITT effects are typically
attenuated toward the null relative to the effect of full adherence, so
that a drug that would be beneficial if everyone took it as directed can
produce a modest or null ITT result if enough assigned patients
discontinue treatment. The **per-protocol** analysis attempts to recover
the more specific question by restricting to, or reclassifying patients
according to, what they actually did, comparing adherers to the Cox-2
strategy against adherers to the nonselective strategy rather than
comparing the arms as randomized. Although this is often the clinically
more interesting question, once patients are grouped by what they
actually did rather than by what the coin flip assigned them, the
randomization at baseline no longer protects the comparison. Whatever led
a given patient to adhere rather than switch or discontinue (eg, symptom
severity, general health, health literacy, or tolerance of side effects)
may also be connected to the outcome, which is the mechanism Chapter 1
called the healthy-user effect, now operating *inside* a randomized trial
rather than across an observational cohort. A naive per-protocol
comparison can therefore be confounded even though the trial that
produced the data was, at the point of randomization, perfectly
randomized.

The `sta` cohort makes these concerns concrete without requiring a real
trial. Because `sta` is observational, and nobody was randomized to start
a statin, any comparison of patients who initiated and *stayed on* a
statin for some minimum duration against everyone else is already
per-protocol in spirit, and already vulnerable to the same
self-selection just described. It is vulnerable to a further problem as
well: misclassifying the follow-up time before a patient meets the
"adherer" criterion can build survival into the definition of one of the
groups, a distortion called immortal time bias that Chapter 6 introduces
on its own terms. Chapter 12 develops the clone-censor-weight
approach specifically to answer a per-protocol-style question — what
would happen if patients followed a specific treatment strategy for its
full duration — using observational data like `sta`, without letting
either kind of bias back in.

## The target trial

Every method in this book adjusts observational data so that it can stand
in for a randomized trial. Hernán and Robins (2016) turned that
observation into a discipline, called the **target trial** framework: for
any causal question one wants to answer with observational data, one
first specifies, in full protocol detail, the randomized trial that would
answer it, as though funding, ethics, and recruitment were not obstacles,
and only then describes how each piece of that protocol will be
*emulated* using the data actually available. Almost none of the trials
specified this way will ever be run. The value of writing the protocol
anyway is that it brings every design decision a trial would have to make
(who is eligible, what counts as each treatment strategy, when follow-up
starts, and how the outcome is defined) into the open before any
adjustment method is chosen, instead of leaving those decisions implicit
in a modeling script.

Here is that protocol, written out for the Cox-2 question this book has
been asking since Chapter 1:

| Protocol component | Target trial | Emulation using `ns` |
|---|---|---|
| Eligibility | Adults beginning NSAID therapy, with no NSAID use in the preceding period | Same criterion, applied when `ns` was assembled: new users only |
| Treatment strategies | (1) Initiate a Cox-2 selective NSAID; (2) initiate a nonselective NSAID | `cox2_init == 1` versus `cox2_init == 0` at the qualifying prescription |
| Assignment | Random allocation to strategy (1) or (2) | Not randomized; approximated by conditioning on measured confounders (`ns_covs`) under the conditional exchangeability assumption from Chapter 3 |
| Follow-up | Begins at randomization ("time zero") and continues to the outcome, loss to follow-up, or a fixed administrative end | Begins at the qualifying prescription date, ends at incident PUD, loss to follow-up, or the administrative end |
| Outcome | Incident peptic ulcer disease | `pud == 1` observed during follow-up |
| Causal contrast | Intention-to-treat comparison of the two assigned strategies | ITT analog: compare all patients who initiated each strategy at time zero, regardless of later switching or discontinuation |
| Analysis plan | Risk difference or risk ratio between arms, estimated directly from randomized groups | Same contrast, estimated with an adjustment method (standardization, propensity scores, or IPTW — Chapters 7 through 9) that stands in for randomization |

Six of these seven rows can already be written down for `ns` with no
controversy, in the same language a trial protocol would use: the
eligibility criteria are those used to identify new users, the two
strategies are already coded in `cox2_init`, follow-up already has a
defined start and end, the outcome is already defined as `pud`, and
even the causal contrast and analysis plan can be stated without
needing to resolve anything about confounding first. Only the
**Assignment** row cannot be filled in as "randomized," because
assignment was not randomized, and that single row is where the rest of
this book does its work. Standardization and outcome regression,
propensity scores and inverse probability weighting, and instrumental
variables and sensitivity analysis are each a different technical
strategy for filling in that one row with something that behaves enough
like a coin flip to make the causal contrast in the last row credible.
Chapters 6, 10, and 12 return to a version of this same protocol, adapted
for study design choices, for survival outcomes, and for the dynamic
treatment strategies studied in `sta`, although the structure of the
argument is the same in each case: specify the trial first, then argue,
row by row, for how well the observational data emulate it.

One application of this logic has become common enough in regulatory
submissions to deserve a name here. In an **external control arm**
design, a single-arm trial — typically in a rare disease, or in an
oncology setting where randomizing patients to the standard of care is
hard to justify — is compared against a comparator group assembled from
observational data: a disease registry, an electronic health record
extract, or the control arm of an earlier trial. The target trial is the
two-arm randomized trial that was not run, and the emulation problem is
the one Part II of this book spends its chapters on, since the trial
patients and the external controls were selected by different criteria,
treated in different calendar periods, and followed by different means.
Every row of the protocol table therefore has to be argued for rather
than assumed, and the Assignment row carries the same weight it carries
above, which is why the confounding machinery of the next several
chapters applies here without alteration. The course's session on
external control arms works through examples of submissions that were
accepted and of submissions that were not.

## Exercises

1. Draft a target trial protocol table, in the same seven-row format
   used above, for the statin question this book has been asking
   alongside the Cox-2 question. Use the `sta` cohort and `sta_covs` to
   fill in eligibility, treatment strategies, assignment, follow-up,
   outcome, causal contrast, and analysis plan in your own words.
2. Immortal time bias was named in this chapter as a hazard for
   per-protocol-style comparisons in `sta`, and Chapter 6 takes it up in
   full. Looking at the seven rows of the target trial table above,
   which single protocol component does immortal time bias violate, and
   why does misclassifying a patient's follow-up before they meet a
   treatment-defining criterion break that component specifically,
   rather than, say, the eligibility or outcome rows?
