# Causal diagrams

```{r setup-ch05, include=FALSE}
source("R/setup.R")
```

Chapter 1 named confounding by indication in words: physicians steer
Cox-2 selective NSAIDs toward patients they already judge to be at
higher gastrointestinal risk, which is why the crude comparison came out
backwards. Chapter 3 gave that idea a formal home inside potential
outcomes, as a failure of exchangeability, and named the covariates that
would need to be equal on average across treatment groups for a
comparison to be believable. Neither chapter offered a way to *look* at a
causal assumption and read off which variables belong on that list, or,
equally important, which variables do not belong on it and would damage
the analysis if included. In this chapter we develop that language. A
causal diagram translates a set of substantive beliefs about how
variables relate to one another into a picture that can be reasoned about
with a small number of rules, checked with software, and reviewed by
colleagues who may not share our subject-matter expertise but can still
read a graph.

## Drawing causal assumptions

A causal diagram, in the sense we use throughout this book, is a directed
acyclic graph (DAG): a collection of **nodes**, one for each variable
that matters to the question, connected by **arrows** that represent
direct causal effects. An arrow from $A$ to $B$ asserts that $A$ has
some effect on $B$, holding everything else in the picture fixed; the
absence of an arrow is an equally substantive claim, that no such direct
effect exists. "Acyclic" rules out feedback loops drawn as a single
static picture, so that $A$ cannot cause $B$ which causes $A$ within the
same diagram, although time can be represented by drawing $A$ at one
point in time and $A$ at a later point as two separate nodes. Every
arrow in a DAG represents a claim that someone is prepared to defend on
scientific grounds, rather than a quantity estimated from data, and the
diagram is where such assumptions are written down and debated before any
model is fit.

A **path** between two nodes is any sequence of arrows connecting them,
regardless of direction. Some paths are **causal paths**: every arrow
along the way points forward, from the exposure toward the outcome,
tracing an actual mechanism by which the exposure could affect the
outcome. Other paths are **backdoor paths**: they connect exposure to
outcome but enter the exposure node through an arrow pointing *into*
it, meaning they trace a route through some third variable that causes
both, rather than an effect of the exposure itself. A crude comparison
of exposed and unexposed patients pools every open path between them,
causal and backdoor alike, which is why Chapter 1's comparison
of Cox-2 initiators and nonselective initiators mixed whatever real
effect Cox-2 selectivity has on ulcer risk with whatever backdoor paths
ran through GI risk, age, and steroid use. The goal of adjustment,
translated into this graphical language, is to close every backdoor
path while leaving every causal path open. This criterion is precise
enough that software can check it, as the next section demonstrates for
the NSAID question. The formal name for the state we are
aiming at is **d-separation**: two variables are d-separated, given a set
of conditioning variables, when every path between them is blocked. It is
the term used in the course's DAG workshop and throughout the
graphical-models literature, and it is what `dagitty` is checking when we
ask it whether an adjustment set suffices.

## The NSAID--PUD diagram

We now encode in a diagram the causal story Chapters 1 and 3 have been
telling about Cox-2 selectivity and peptic ulcer disease: GI risk
factors, age, and corticosteroid use each influence a physician's choice
between a Cox-2 selective and a nonselective NSAID, and each also
influences ulcer risk directly, alongside whatever effect the drug
itself has.

```{r nsaid-dag, message=FALSE}
library(ggdag)
dag_nsaid <- dagify(
  pud ~ cox2 + gi_risk + age + steroids,
  cox2 ~ gi_risk + age + steroids,
  exposure = "cox2",
  outcome = "pud",
  labels = c(cox2 = "Cox-2 selective", pud = "Peptic ulcer",
             gi_risk = "GI risk factors", age = "Age",
             steroids = "Corticosteroids")
)
ggdag(dag_nsaid, text = FALSE, use_labels = "label") + theme_dag()
```

The diagram makes the confounding structure explicit: `gi_risk`, `age`,
and `steroids` each sit upstream of both `cox2` and `pud`, so each one
opens a backdoor path from `cox2` to `pud` that has nothing to do with
the drug's own effect — `cox2 <- gi_risk -> pud`, `cox2 <- age -> pud`,
and `cox2 <- steroids -> pud`, alongside the single causal path
`cox2 -> pud` that is the object of interest. Rather than reason through
each backdoor path by hand, we can let `dagitty` do it directly:

```{r nsaid-adjustment-sets}
dagitty::adjustmentSets(dag_nsaid)
```

`adjustmentSets()` reports a single minimal set here: `{age, gi_risk,
steroids}`. That result tells us two things. First, adjusting for all
three variables together closes every backdoor path in the diagram, so
that a comparison of Cox-2 and nonselective initiators that conditions on
age, GI risk factors, and steroid use, whether by stratification,
standardization, or a propensity score built from them (the subjects of
Chapters 7 through 9), is no longer contaminated by the confounding this
diagram encodes. Second, no smaller set will suffice: because each of the
three variables opens its own separate backdoor path with no shared
intermediate node, dropping any one of them leaves that path open, and
the comparison remains confounded through whichever variable was left
out. This is the graphical counterpart of the conditional exchangeability
condition from Chapter 3, $Y(a)
\perp A \mid L$ for $L$ equal to this adjustment set, with the difference
that the set is here read off a picture rather than asserted, and that
the diagram remains available for a colleague to challenge by redrawing
an arrow.

## Confounders, colliders, and mediators

Real diagrams can have dozens of nodes, although nearly everything about
how adjustment behaves follows from three small structures that recur
throughout them. Recognizing each structure, and adjusting only for the
first of the three, accounts for most of what this chapter is meant to
teach.

A **confounder** is a common cause of exposure and outcome, the pattern
`gi_risk`, `age`, and `steroids` all instantiate above: a variable with
arrows pointing into both the exposure and the outcome, opening a
backdoor path that adjustment is meant to close.

```{r confounder-dag}
dag_confounder <- dagify(
  y ~ x + c,
  x ~ c,
  exposure = "x", outcome = "y",
  labels = c(x = "Exposure", y = "Outcome", c = "Confounder")
)
ggdag(dag_confounder, text = FALSE, use_labels = "label") + theme_dag()
```

A **collider** is the mirror image: a common *effect* of two other
variables, with arrows pointing *into* it from both sides rather than
out of it.

```{r collider-dag}
dag_collider <- dagify(
  m ~ x + z,
  exposure = "x", outcome = "z",
  labels = c(x = "Cause 1", z = "Cause 2", m = "Collider")
)
ggdag(dag_collider, text = FALSE, use_labels = "label") + theme_dag()
```

Left alone, a collider does nothing to the association between its two
causes, so that if `x` and `z` are independent they remain independent
whether or not anyone ever looks at `m`. Difficulties arise once `m` is
conditioned on, whether deliberately, by restricting a sample to `m`'s
observed value, or inadvertently, by studying only the patients who
survived to be observed at all. Conditioning on a common effect induces
an association between its causes where none existed, because within any
fixed value of `m`, learning that one cause was small makes the other
cause more likely to have been large enough to produce that value of `m`
on its own. We can make this concrete with a short simulation, in which
we draw two variables that are causally unrelated to each other, build a
third variable as their common effect, and compare the association
between the first two before and after restricting to one level of the
third.

```{r collider-sim}
set.seed(5)
n <- 5000
gi_risk_indep <- rnorm(n)    # cause 1, unrelated to cause 2 by construction
frailty       <- rnorm(n)    # cause 2, unrelated to cause 1 by construction

# collider: a variable such as "hospitalized" driven by both causes plus noise
hosp_score   <- gi_risk_indep + frailty + rnorm(n)
hospitalized <- as.integer(hosp_score > quantile(hosp_score, 0.75))

collider_df <- tibble(gi_risk_indep, frailty, hospitalized)

fit_full <- lm(frailty ~ gi_risk_indep, data = collider_df)
fit_cond <- lm(frailty ~ gi_risk_indep, data = collider_df |> filter(hospitalized == 1))
```

```{r collider-sim-values, include=FALSE}
b_full <- coef(fit_full)["gi_risk_indep"]
p_full <- summary(fit_full)$coefficients["gi_risk_indep", "Pr(>|t|)"]
b_cond <- coef(fit_cond)["gi_risk_indep"]
p_cond <- summary(fit_cond)$coefficients["gi_risk_indep", "Pr(>|t|)"]
```

We simulated `gi_risk_indep` and `frailty` as independent draws, so the
true association between them is zero by construction, and the
regression run on the full sample confirms it: the slope is
`r sprintf("%.3f", b_full)` (p = `r sprintf("%.2f", p_full)`),
indistinguishable from no relationship at all. When we restrict that
same regression to the `r sum(collider_df$hospitalized)` patients in the
top quartile of the collider score — patients who were "hospitalized"
in this simulation — we turn that null slope into
`r sprintf("%.3f", b_cond)` (p `r if (p_cond < 0.001) "< 0.001" else sprintf("= %.3f", p_cond)`),
a strong and highly significant association between two variables that
have no causal connection at all. Nothing about the data-generating
process differs between the two regressions; only the sample does. This
is collider bias, and it illustrates that adjustment applied to the wrong
kind of node does not merely fail to help but induces the sort of
spurious association that the methods in this book are intended to
avoid.

The third structure is a **mediator**: a variable that sits on the
causal path itself, transmitting some or all of the exposure's effect
on the way to the outcome rather than confounding it.

```{r mediator-dag}
dag_mediator <- dagify(
  y ~ m,
  m ~ x,
  exposure = "x", outcome = "y",
  labels = c(x = "Exposure", m = "Mediator", y = "Outcome")
)
ggdag(dag_mediator, text = FALSE, use_labels = "label") + theme_dag()
```

Adjusting for a mediator is a different error from adjusting for a
collider, since a mediator is genuinely caused by the exposure and
genuinely causes the outcome, although it remains inappropriate for the
question we ask throughout this book, which concerns the *total* effect
of exposure on outcome. Conditioning on `m` blocks the causal path
`x -> m -> y` that carries the effect we are trying to estimate, so a
regression of `y` on `x` that also includes `m` answers a narrower
question, namely how much of the effect operates independently of that
pathway, while resembling the same kind of adjusted analysis that closes
a backdoor path. Confounders therefore belong in the adjustment set,
whereas colliders and mediators, for their different reasons, do not.

## Selection bias as collider stratification

The collider simulation above used "hospitalized" as a placeholder,
although the same mechanism underlies a problem known by a more familiar
name, selection bias. The structural reading of selection bias, in which
it is understood as conditioning on a collider rather than as a catalog
of study-design mishaps, is the account given by Hernán,
Hernández-Díaz, and Robins (2004), and it is the one the rest of this
section follows. Suppose Cox-2 selectivity has some effect on the chance
a patient tolerates therapy long enough to stay in a study, and,
independently, that some unmeasured element of a patient's underlying
prognosis also affects whether they remain in follow-up long enough to
have their outcome observed. "Remained in the study" is then a collider,
with an arrow into it from treatment and an arrow into it from unmeasured
prognosis.

```{r selection-dag}
dag_selection <- dagify(
  pud ~ u,
  s ~ cox2 + u,
  exposure = "cox2", outcome = "pud",
  labels = c(cox2 = "Cox-2 selective", pud = "Peptic ulcer",
             u = "Unmeasured prognosis", s = "Remained in study")
)
ggdag(dag_selection, text = FALSE, use_labels = "label") + theme_dag()
```

Any analysis that sees only patients who remained in the study, which
describes every complete-case analysis unless something deliberate is
done about it, is conditioning on `s` whether or not a `filter()` call
was ever written, because patients who were lost to follow-up do not
appear in the rows being summed. By the same logic as the simulation
above, that conditioning can open an association between `cox2` and `pud`
running through `u`, even in a world where the diagram already grants
that GI risk, age, and steroid use have been fully adjusted for and the
drug has no effect on ulcer risk. Measuring additional baseline
covariates does not remove this bias, since it does not arise from
confounding but from which patients are present to be analyzed at all,
after treatment has had a chance to influence who stays and who leaves.
Chapter 11 takes up this problem under its clinical name, informative
censoring, and develops inverse probability of censoring weights, which
reweight the patients who remain so that they stand in, on average, for
everyone who should have been followed.

## Exercises

1. Add a node for `otc_nsaid` (over-the-counter NSAID use, unmeasured in
   `ns`) to `dag_nsaid`, with arrows into both `cox2` and `pud`, and
   re-run `dagitty::adjustmentSets()` on the result. Does an adjustment
   set still exist? If `otc_nsaid` cannot be included because it is
   never recorded, what does that imply about the conditional
   exchangeability assumption from Chapter 3 as applied to the `ns`
   cohort?
2. Draw a DAG for the statin question using `dagify()`, with `statin`
   as the exposure and cardiovascular event as the outcome, that
   includes a node for the healthy-user behaviors described in Chapter 1
   (exercise, diet, medication adherence generally) as a common cause of
   both statin initiation and the outcome. Run `adjustmentSets()` on
   your diagram and compare the set it returns to `sta_covs`.
3. Modify `collider_df` in the `collider-sim` chunk so that
   `hospitalized` is defined by restricting to the *bottom* quartile of
   `hosp_score` instead of the top quartile. Re-fit the conditional
   regression and report the new slope. Is the sign the same as in the
   chapter? Explain, in terms of the collider mechanism, why
   conditioning on either tail of a common effect induces an
   association, while conditioning on nothing (the full sample) does
   not.
4. Consider a study of SSRI use and lung cancer. Write down a DAG with
   `dagify()` in which SSRI use is caused by depression, lung cancer is
   caused by smoking, smoking is caused by depression, and coronary artery
   disease is caused by smoking, taking SSRI use as the exposure and lung
   cancer as the outcome. Coronary artery disease is the variable of the
   four most reliably recorded in claims data, so an analyst who could
   adjust for only one thing would be tempted by it. Using the backdoor
   rules of this chapter, say what adjusting for coronary artery disease
   alone does to the exposure–outcome association, and what changes when
   depression is added alongside it. Then run `adjustmentSets()` on your
   diagram and report the minimal sufficient set. Which of the three
   strategies — coronary artery disease alone, coronary artery disease
   with depression, or the set `dagitty` returns — would you defend, and
   on what grounds?
