# Study design for causal questions

```{r setup-ch06, include=FALSE}
source("R/setup.R")
```

Chapter 3 gave us estimands, Chapter 4 gave us the target trial as the
protocol every observational analysis is implicitly compared against,
and Chapter 5 gave us a way to read a confounding structure off a
diagram. All three chapters, though, took for granted that the rows of
the target trial table — who is eligible, what counts as each treatment
strategy, when follow-up starts — had already been settled before any
adjustment method entered the picture. This chapter is about settling
them. We look at three design choices that determine what a causal
comparison is even capable of estimating, independent of which model
eventually gets fit to the data: choosing new users over prevalent
users, choosing an active comparator over no treatment at all, and
aligning the moment a patient becomes eligible with the moment their
follow-up clock starts. Get any of the three wrong and no amount of
standardization, propensity-score weighting, or careful modeling in the
chapters ahead can put it right, because the bias is baked into which
rows and which person-time entered the analysis in the first place.

## Design before analysis

An estimand, in the language of Chapter 3, is a contrast between
potential outcomes in some population — the ATE if we mean everyone,
the ATT if we mean the patients who were actually treated. Which
contrast we can identify from a data set is decided well before any
regression is fit, by choices that belong entirely to design: who is
eligible to enter the cohort, what exactly distinguishes the treatment
groups, when each patient's clock starts, and what serves as the
yardstick the treated group is measured against. The target trial
protocol from Chapter 4 is really a checklist of exactly these
decisions, written out in the same seven rows a trial's own protocol
would use, and it is deliberately silent on adjustment: eligibility,
treatment strategies, and follow-up can all be specified in full before
we have said a single word about standardization or weighting.

That ordering matters because no statistical method can repair a
design that answers the wrong question. If eligibility quietly mixes
patients who have been on a drug for years with patients just starting
it, the sample no longer corresponds to any one estimand — not the
effect of starting the drug, not the effect of continuing it — no
matter how well the eventual model fits. Standardization, propensity
scores, and inverse probability weighting, the subjects of Part II, are
strategies for closing a *measured* gap in exchangeability once the
design has already fixed who is compared to whom, at what moment, on
what scale; they have nothing to offer a design that never posed an
answerable question in the first place. This chapter works through
three of the design decisions most responsible for that failure in
pharmacoepidemiology, and shows how each is already built into the two
cohorts we have used since Chapter 1.

## New-user designs

Chapter 1 named prevalent-user bias in passing; here we work out why it
distorts a comparison and how the new-user design fixes it. A
**prevalent-user** comparison enrolls patients who are already on a
drug, of whatever duration, and compares them to patients who are not.
Two problems follow from that choice, and they are different problems
even though they often travel together. The first is **depletion of
susceptibles**: patients who could not tolerate a drug, or for whom it
plainly was not working, tend to stop taking it early and so drop out
of the pool of "current users" long before anyone samples that pool.
The current users left behind are enriched for exactly the people who
tolerated the drug well, which biases any comparison toward the drug
looking safer and more effective than it is for the population that
ever started it. The second problem is that covariates measured at the
same moment as prevalent exposure may already have been changed by the
treatment itself. A blood pressure reading taken after several years on
a drug reflects, in part, whatever the drug did to blood pressure; using
it to "adjust for confounding" adjusts away part of the treatment
effect, or worse, opens exactly the kind of collider path Chapter 5
warned against, if some other unmeasured factor also drives both that
downstream covariate and the outcome.

The **new-user design** avoids both problems by anchoring the cohort to
the moment of treatment initiation rather than to some arbitrary later
moment of observation. A patient enters the cohort only when we observe
them starting the drug, with no recorded use of that drug or its class
immediately beforehand, and every covariate used for adjustment is
measured at or before that same moment — before treatment has had any
chance to change it. This is exactly how `ns` was built. Reviewing NAMCS
office visits from 2005 through 2009, a patient enters `ns` at the visit
where they are observed initiating an NSAID, with no current NSAID use
recorded at that visit, and their initiating prescription — Cox-2
selective or nonselective — fixes which arm, `cox2_initiation`, they
belong to for the rest of the cohort. Of the `r nrow(ns)` patients in
`ns`, `r sum(ns$cox2_initiation == "Yes")` are captured this way as
Cox-2 initiators and the rest as nonselective initiators; nobody in
either group had already been taking an NSAID before the qualifying
visit, and nobody's baseline covariates were measured after the drug
had a chance to act on them. `ns` is, in other words, already a
new-user cohort by construction — we have simply not yet named the
design principle it follows.

Pharmacoepidemiology has accumulated names for the specific ways these
comparisons go wrong, and the names are worth knowing because they are how
reviewers will describe your study. Each is a special case of confounding
or of misaligned time, and each is answered by some part of the new-user,
active-comparator design.

> **Healthy user effect.** People who start and continue preventive
> therapy also exercise, screen, and adhere to other advice, so the drug
> inherits credit for the behavior (Chapter 1). An active comparator helps
> because both arms have consulted a physician and accepted a
> prescription.
>
> **Healthy adherer effect.** The same phenomenon inside a trial or cohort
> arm: adherence itself marks health, so comparing adherent to nonadherent
> patients confounds the drug with the disposition. Chapter 12's protocol
> weights, not the design, are the answer here.
>
> **Sick stopper effect.** Patients who become ill discontinue treatment,
> so the untreated person-time is enriched for people about to have
> outcomes. New-user follow-up from a fixed time zero keeps the
> discontinuation downstream of the exposure assignment rather than
> allowed to redefine it.
>
> **Protopathic bias.** A drug prescribed for the earliest symptoms of an
> undiagnosed disease appears to cause the disease it was reaching for.
> Anchoring time zero at initiation and, where needed, lagging the outcome
> window prevents the drug from taking credit for what prompted it.
>
> **Depletion of susceptibles.** Described above: prevalent users are the
> survivors of the early attrition, so the pool is enriched for
> tolerators. Enrolling at initiation samples everyone who ever started.

## Active-comparator designs

A new-user design still leaves open a further choice: new users of the
drug of interest, compared to *what*? The most obvious comparator is
patients who did not start any NSAID at all. It is also, for this
question, a poor one. Whether a physician prescribes an NSAID in the
first place is itself a treatment decision, driven by whether a patient
has arthritis, joint pain, or another condition that calls for one — and
that same decision is entangled with everything connected to needing an
NSAID at all, most of which has nothing to do with GI risk specifically.
Comparing Cox-2 initiators to non-initiators folds confounding by
indication for *needing an NSAID* together with whatever separately
drives the choice *between* the two drug classes, and the resulting
mixture is much harder to reason about than either piece alone.

An **active-comparator design** sidesteps the first piece entirely by
choosing the comparison group from among patients who also started an
NSAID, just the other class. Every patient in `ns`, regardless of arm,
already needed and received NSAID therapy, so whatever led a physician
to prescribe an NSAID in the first place is held fixed across both
arms — it cannot be a source of imbalance between them, because it is
shared by construction. What remains to differ between Cox-2 initiators
and nonselective initiators is the narrower, more nameable set of
factors that push a physician toward one NSAID rather than the other
once the decision to prescribe some NSAID has already been made: age,
GI risk history, concurrent steroid use — precisely the adjustment set
`{age, gi_risk, steroids}` that Chapter 5's DAG identified. An active
comparator does not eliminate confounding; it shrinks the dimension we
have to worry about from "everything connected to needing pain
treatment" down to "whatever distinguishes one NSAID from the other,"
which is both more plausible to measure in full and more amenable to
the adjustment methods in Part II.

We can see exactly how much confounding remains, and along which
covariates, by rebuilding Chapter 2's Table 1 for every covariate in
`ns_covs` rather than the handful used there for illustration.
`ns_covs` mixes two kinds of variables — `age` and `year` are numeric,
everything else is categorical — so we route each kind through its own
helper from `R/table1.R`, casting `year` to numeric first since it
arrives from NAMCS as a factor.

```{r ch06-table1-continuous}
source("R/table1.R")
ns_year_num <- ns |> mutate(year = as.numeric(as.character(year)))
table1_continuous(
  ns_year_num,
  vars = c("age", "year"),
  by = "cox2_initiation"
) |> knitr::kable(
  col.names = c("Variable", "Nonselective (No)", "Cox-2 (Yes)"),
  caption = "Table 1, continuous covariates, ns cohort by cox2_initiation"
)
```

```{r ch06-table1-categorical}
table1_categorical(
  ns,
  vars = setdiff(ns_covs, c("age", "year")),
  by = "cox2_initiation"
) |> knitr::kable(
  col.names = c("Variable", "Level", "Nonselective (No)", "Cox-2 (Yes)"),
  caption = "Table 1, categorical covariates, ns cohort by cox2_initiation"
)
```

```{r ch06-table1-values, include=FALSE}
gap_pct <- function(data, var, level = "Yes", by = "cox2_initiation") {
  data |>
    group_by(.data[[by]]) |>
    summarise(p = 100 * mean(.data[[var]] == level, na.rm = TRUE), .groups = "drop") |>
    pull(p)
}
age_grp   <- ns |> group_by(cox2_initiation) |> summarise(m = mean(age), .groups = "drop") |> pull(m)
arth_grp  <- gap_pct(ns, "arthritis")
htn_grp   <- gap_pct(ns, "hypertension")
diab_grp  <- gap_pct(ns, "diabetes")
osteo_grp <- gap_pct(ns, "osteoporosis")
cort_grp  <- gap_pct(ns, "corticosteroid_use")
ppi_grp   <- gap_pct(ns, "ppi_use")
sex_grp   <- gap_pct(ns, "sex", level = "Female")
south_grp <- gap_pct(ns, "region", level = "South")
year_grp  <- ns_year_num |> group_by(cox2_initiation) |> summarise(m = mean(year), .groups = "drop") |> pull(m)
```

Even with the comparator restricted to a shared class of drugs, several
covariates still separate the two arms by a wide margin. The starkest
gap is arthritis: `r sprintf("%.1f", arth_grp[1])`% of nonselective
initiators carry an arthritis diagnosis against
`r sprintf("%.1f", arth_grp[2])`% of Cox-2 initiators, a
`r sprintf("%.0f", arth_grp[2] - arth_grp[1])`-percentage-point gap that
dwarfs anything else in either table. Age moves the same way — Cox-2
initiators average `r sprintf("%.1f", age_grp[2])` years against
`r sprintf("%.1f", age_grp[1])` among nonselective initiators, a gap of
roughly `r sprintf("%.0f", age_grp[2] - age_grp[1])` years — and
corticosteroid use, though rare overall, is more than four times as
common among Cox-2 initiators (`r sprintf("%.1f", cort_grp[2])`% versus
`r sprintf("%.1f", cort_grp[1])`%). Hypertension, diabetes, osteoporosis,
and PPI use all run a few points higher among Cox-2 initiators as well
(`r sprintf("%.1f", htn_grp[1])`% vs `r sprintf("%.1f", htn_grp[2])`%,
`r sprintf("%.1f", diab_grp[1])`% vs `r sprintf("%.1f", diab_grp[2])`%,
`r sprintf("%.1f", osteo_grp[1])`% vs `r sprintf("%.1f", osteo_grp[2])`%,
and `r sprintf("%.1f", ppi_grp[1])`% vs `r sprintf("%.1f", ppi_grp[2])`%).

None of this is a coincidence. Arthritis and the comorbidities that
travel with it — hypertension, diabetes, osteoporosis, PPI use — are
the substance behind the diagram's `gi_risk` node; age lines up with
the DAG's own `age` node; corticosteroid use lines up with its
`steroids` node. Three separate confounders, three separate arrows into
both `cox2` and `pud` in Chapter 5's diagram, and each still shows up
here as a real imbalance. Sex (`r sprintf("%.1f", sex_grp[1])`% female
vs `r sprintf("%.1f", sex_grp[2])`%) and calendar year
(`r sprintf("%.1f", year_grp[1])` vs `r sprintf("%.1f", year_grp[2])`)
sit close to balanced, which fits the diagram: neither has an arrow
into `cox2`. Region fits neither pattern:
`r sprintf("%.1f", south_grp[1])`% of nonselective initiators are from
the South against `r sprintf("%.1f", south_grp[2])`% of Cox-2
initiators, a `r sprintf("%.0f", south_grp[2] - south_grp[1])`-point
gap on par with hypertension's — real, but with no node in Chapter 5's
diagram and no obvious clinical mechanism tying geography to drug
choice. A Table 1 gap and a DAG arrow are not the same claim; this one
more likely reflects regional prescribing habits than confounding by GI
risk, and it deserves its own scrutiny rather than being waved away as
either explained or safely ignorable. The gaps that do map onto the
diagram — arthritis and its fellow travelers, age, and corticosteroid
use — are what Part II's adjustment methods exist to close.

## Immortal time and other time-zero problems

A design can get eligibility and the comparator right and still fail on
a third dimension: when, exactly, each patient's follow-up clock starts.
**Aligned time zero** means that three things happen on the same date
for every patient: the date they are confirmed eligible, the date they
are classified into a treatment strategy, and the date their
person-time starts counting toward the outcome. `ns` satisfies this
automatically, because eligibility, drug class, and the start of
follow-up are all pinned to the same qualifying visit. Aligned time zero
is easy to state and surprisingly easy to violate the moment a study
defines "exposed" using any information that could only be known after
follow-up has already begun.

Here is a worked hypothetical that shows exactly how the violation
bites, built around a definition of treatment that shows up constantly
in pharmacoepidemiology: calling someone a "user" only once they have
filled a second prescription, some grace period after the first.
Suppose we simulate 20,000 hypothetical patients, each with a true time
to death that is, by construction, unrelated to anything about their
future drug-taking behavior, and suppose "persistent user" status
requires two things: surviving past a 90-day grace period, and an
independent adherence habit that is fixed for each patient and has no
effect on survival whatsoever.

```{r ch06-immortal-sim}
set.seed(6)
n <- 20000
grace <- 90
follow_up <- 730

hyp <- tibble(
  id = 1:n,
  death_time = rexp(n, rate = 1 / 2000),
  adherent   = rbinom(n, 1, 0.6)   # a habit unrelated to survival, by construction
) |>
  mutate(
    alive_grace = death_time > grace,
    persistent  = as.integer(alive_grace & adherent == 1),
    died_by_end = as.integer(death_time <= follow_up)
  )
```

A naive analysis, of the kind a misaligned time-zero produces, starts
every patient's clock on day zero and simply compares mortality risk
between "persistent users" and everyone else over the full
`r follow_up`-day window:

```{r ch06-immortal-naive}
naive_tab <- hyp |>
  group_by(persistent) |>
  summarise(n = n(), risk = mean(died_by_end), .groups = "drop")
naive_tab |> knitr::kable(digits = 3)
```

```{r ch06-immortal-values, include=FALSE}
naive_diff <- naive_tab$risk[naive_tab$persistent == 0] -
  naive_tab$risk[naive_tab$persistent == 1]

landmark_tab <- hyp |>
  filter(alive_grace) |>
  group_by(adherent) |>
  summarise(n = n(), risk = mean(died_by_end), .groups = "drop")
landmark_diff <- landmark_tab$risk[landmark_tab$adherent == 1] -
  landmark_tab$risk[landmark_tab$adherent == 0]
```

Persistent users appear dramatically protected —
`r sprintf("%.1f", 100 * naive_diff)` percentage points lower mortality
risk than everyone else — even though nothing in the simulation gave
adherence any causal effect on survival at all. The reason is
mechanical rather than clinical: to be labeled "persistent," a patient
had to survive at least `r grace` days, so the persistent group starts
its `r follow_up`-day clock with `r grace` guaranteed event-free days
already banked, days that could never have contained a death no matter
who was in that group. That banked, guaranteed-survival stretch is
**immortal time**, and it is entirely an artifact of letting the
exposure definition depend on information — surviving the grace
period — that was not yet available at the true time zero, the date
treatment actually started.

The fix is to give every patient the same time zero regardless of
group: restrict the comparison to patients who survived to the end of
the grace period, and start everyone's clock there, at day
`r grace`, rather than at day zero.

```{r ch06-immortal-landmark}
landmark_tab |> knitr::kable(digits = 3)
```

Among the `r sum(hyp$alive_grace)` patients who survived to day
`r grace`, adherent and non-adherent patients now have essentially
identical mortality risk over the remaining follow-up
(`r sprintf("%.3f", landmark_tab$risk[landmark_tab$adherent == 1])`
versus `r sprintf("%.3f", landmark_tab$risk[landmark_tab$adherent == 0])`,
a difference of `r sprintf("%.3f", landmark_diff)`) — indistinguishable
from the null effect we built into the simulation. Nothing about any
patient changed between the two analyses; only the point at which each
patient's clock was allowed to start changed, and that alone was enough
to manufacture, and then erase, a seven-point "protective" effect.

Chapter 4 flagged this exact hazard for the `sta` cohort, where any
comparison built around staying on a statin for some minimum duration
is vulnerable to precisely this mechanism, and left the question of
which row of the target trial table it violates for this chapter to
settle. The answer is the **Follow-up** row: the row that specifies
when person-time starts accruing toward the outcome. Immortal time
bias arises whenever the Assignment row — how a patient gets classified
into a treatment strategy — is allowed to depend on something observed
after the Follow-up row's start date, rather than at it. A short
checklist captures the requirement for any design in this book,
`ns`, `sta`, or a design of your own:

| Component | Must be fixed as of the same date as... |
|---|---|
| Eligibility | the date all inclusion criteria are first met |
| Assignment | the date the patient's treatment strategy is classified |
| Start of follow-up | the date person-time begins accruing toward the outcome |

Whenever those three dates coincide, as they do in `ns` by
construction, immortal time cannot arise, because there is no gap
between becoming eligible and starting the clock in which future
information could leak backward into a treatment label. Whenever they
are allowed to drift apart, as in the persistent-user hypothetical
above, the resulting comparison can manufacture an effect, hide a real
one, or both, entirely independent of anything an adjustment method in
Part II could repair. The argument that this is what specifying a
target trial buys you — that making eligibility, assignment, and the
start of follow-up coincide forecloses immortal time bias rather than
leaving it to be corrected afterward — is Hernán and colleagues' (2016,
*Journal of Clinical Epidemiology*).

## Exercises

1. Using what `R/setup.R` tells you about `sta` — that it follows
   patients who did or did not initiate a statin, with time variables
   `cv_time`, `death_time`, and `loss_fu_time` measured from cohort
   entry — describe whether `sta` follows a new-user or a
   prevalent-user design, and state what its time zero is. Would your
   answer change if `sta` had instead been built by sampling patients
   already filling statin prescriptions at some arbitrary visit, rather
   than at their first qualifying prescription?
2. Look back at the two Table 1 chunks built from `ns_covs` in this
   chapter. Besides arthritis and age, name two more covariates where
   the gap between Cox-2 and nonselective initiators is largest, and
   propose a clinical reason each one might still differ even after
   restricting the comparison to two active NSAID options.
3. `sta` currently compares statin initiators to patients who did not
   initiate a statin at all — not an active-comparator design. Propose
   a plausible active comparator for the statin question (a different
   medication prescribed for a similar indication, initiated at a
   similar point in a patient's care), and discuss one advantage and
   one disadvantage of switching to it, using the reasoning this
   chapter used for Cox-2 versus nonselective NSAIDs.
