# Study design for causal questions

```{r setup-ch06, include=FALSE}
source("R/setup.R")
```

Chapter 3 introduced estimands, Chapter 4 developed the target trial as
the protocol against which any observational analysis is implicitly
compared, and Chapter 5 showed how a confounding structure can be read
off a diagram. Each of those chapters, however, assumed that the rows of
the target trial table — who is eligible, what counts as each treatment
strategy, when follow-up starts — had already been settled before any
adjustment method entered the picture. In this chapter we take up those
decisions. We consider three design choices that determine what a causal
comparison is capable of estimating, independent of which model
eventually gets fit to the data: choosing new users over prevalent
users, choosing an active comparator over no treatment at all, and
aligning the moment a patient becomes eligible with the moment their
follow-up clock starts. If any of the three is misspecified, no
amount of standardization, propensity-score weighting, or careful
modeling in the chapters ahead will repair it, because the bias has
entered through which rows and which person-time were analyzed.

## Design before analysis

An estimand, in the language of Chapter 3, is a contrast between
potential outcomes in some population — the ATE if we mean everyone,
the ATT if we mean the patients who were actually treated. Which
contrast we can identify from a data set is decided well before any
regression is fit, by choices that belong entirely to design: who is
eligible to enter the cohort, what distinguishes the treatment groups,
when each patient's clock starts, and which group the treated patients
are measured against. The target trial protocol of Chapter 4 is
essentially a checklist of these decisions, written out in the same
seven rows a trial's own protocol would use, and it is deliberately
silent on adjustment: eligibility, treatment strategies, and follow-up
can all be specified in full before standardization or weighting is
mentioned.

That ordering matters because no statistical method can repair a
design that answers the wrong question. If eligibility quietly mixes
patients who have been on a drug for years with patients just starting
it, the sample no longer corresponds to any one estimand — neither the
effect of starting the drug nor the effect of continuing it — no
matter how well the eventual model fits. Standardization, propensity
scores, and inverse probability weighting, the subjects of Part II, are
strategies for closing a *measured* gap in exchangeability once the
design has fixed which patients are being compared and at what moment;
they cannot rescue a design that never posed an answerable question.
This chapter works through three of the design decisions most
responsible for that failure in pharmacoepidemiology, and shows how each
is already built into the two cohorts we have used since Chapter 1.

## New-user designs

Chapter 1 mentioned prevalent-user bias in passing. Here we consider why
it distorts a comparison and how a new-user design addresses it. A
**prevalent-user** comparison enrolls patients who are already on a
drug, of whatever duration, and compares them to patients who are not.
Two distinct problems follow from that choice, although they usually
occur together. The first is **depletion of susceptibles**: patients who
could not tolerate a drug, or for whom it was clearly ineffective, tend
to stop taking it early and so drop out of the pool of "current users"
long before anyone samples that pool. The current users who remain are
enriched with patients who tolerated the drug well, which biases any
comparison toward the drug appearing safer and more effective than it is
for the population that ever started it. The second problem is that
covariates measured at the same moment as prevalent exposure may already
have been changed by the treatment itself. A blood pressure reading
taken after several years on a drug reflects, in part, whatever the drug
did to blood pressure; using it to "adjust for confounding" adjusts away
part of the treatment effect, or, if some other unmeasured factor drives
both that downstream covariate and the outcome, opens the kind of
collider path Chapter 5 warned against.

The **new-user design** avoids both problems by anchoring the cohort to
the moment of treatment initiation rather than to some arbitrary later
moment of observation. A patient enters the cohort only when we observe
them starting the drug, with no recorded use of that drug or its class
immediately beforehand, and every covariate used for adjustment is
measured at or before that same moment — before treatment has had any
chance to change it. This is how `ns` was built. Among NAMCS office
visits from 2005 through 2009, a patient enters `ns` at the visit at
which they are observed initiating an NSAID, with no current NSAID use
recorded at that visit, and their initiating prescription — Cox-2
selective or nonselective — fixes which arm, `cox2_initiation`, they
belong to for the rest of the cohort. Of the `r nrow(ns)` patients in
`ns`, `r sum(ns$cox2_initiation == "Yes")` are captured this way as
Cox-2 initiators and the rest as nonselective initiators; no patient in
either group had already been taking an NSAID before the qualifying
visit, and no patient's baseline covariates were measured after the drug
had a chance to act on them. `ns` is therefore already a new-user cohort
by construction, although we had not previously named the design
principle it follows.

The idea generalizes beyond initiation. Starting a drug is one clinical
decision, but stopping one, switching, and intensifying a dose are
decisions as well, and each can anchor a comparison in the same way: enroll
patients at the moment the decision is made, classify them by what was
decided, and follow them forward. This framing — the **treatment
decision design** (Brookhart 2015) — is the general form of the
new-user design, and it is what allows the target trials of Chapter 4 to
be emulated at decision points other than a first prescription; Chapter
12's treatment strategies rest on this view.

Pharmacoepidemiology has accumulated names for the specific ways these
comparisons go wrong, and the names are worth knowing because they are how
reviewers will describe your study. Each is a special case of confounding
or of misaligned time, and each is addressed by some part of the new-user,
active-comparator design.

> **Healthy user effect.** Patients who start and continue a preventive
> therapy also tend to exercise, undergo screening, and follow other
> medical advice, so an apparent benefit of the drug may instead reflect
> the behavior of the patients taking it (Chapter 1). An active comparator
> mitigates this, since patients in both arms have consulted a physician
> and accepted a prescription.
>
> **Healthy adherer effect.** The same phenomenon operating within a trial
> or a cohort arm: because adherence is itself a marker of health, a
> comparison of adherent with nonadherent patients confounds the effect of
> the drug with the disposition of the patients who take it. Design alone
> does not address this; the protocol weights of Chapter 12 do.
>
> **Sick stopper effect.** Patients who become ill tend to discontinue
> treatment, so untreated person-time becomes enriched with patients who
> are about to experience outcomes. Following new users from a fixed time
> zero keeps discontinuation downstream of the exposure assignment rather
> than letting it redefine that assignment.
>
> **Protopathic bias.** A drug prescribed for the earliest symptoms of an
> as-yet-undiagnosed disease will appear to cause the disease it was
> intended to treat. Anchoring time zero at initiation, and where
> necessary lagging the outcome window, keeps the drug from being blamed
> for the condition that prompted the prescription.
>
> **Depletion of susceptibles.** Described above: prevalent users are
> those who survived the early attrition, so the pool is enriched with
> patients who tolerate the drug. Enrolling patients at initiation instead
> samples everyone who ever started.

## Active-comparator designs

A new-user design still leaves open a further choice: new users of the
drug of interest, compared to *what*? The most obvious comparator is
patients who did not start any NSAID at all, although for this question
that is a poor choice. Whether a physician prescribes an NSAID in the
first place is itself a treatment decision, driven by whether a patient
has arthritis, joint pain, or another condition that calls for one — and
that same decision is entangled with everything connected to needing an
NSAID at all, most of which has nothing to do with GI risk specifically.
Comparing Cox-2 initiators to non-initiators folds confounding by
indication for *needing an NSAID* together with whatever separately
drives the choice *between* the two drug classes, and the resulting
mixture is much harder to reason about than either piece alone.

An **active-comparator design** removes the first of these by drawing
the comparison group from patients who also started an NSAID, of the
other class. Every patient in `ns`, regardless of arm, already needed
and received NSAID therapy, so whatever led a physician to prescribe an
NSAID in the first place is held approximately fixed across both arms
and cannot by itself produce imbalance between them. What can still
differ between Cox-2 initiators and nonselective initiators is the
narrower set of factors that lead a physician to prefer one NSAID over
the other once the decision to prescribe an NSAID has been made: age,
history of GI risk, and concurrent steroid use, which is the adjustment
set `{age, gi_risk, steroids}` identified by Chapter 5's DAG. An active
comparator does not eliminate confounding. It narrows the set of
variables we have to worry about from everything connected to needing
pain treatment down to whatever distinguishes one NSAID from the other,
a set that is more plausibly measured in full and more amenable to the
adjustment methods of Part II.

We can see how much confounding remains, and along which
covariates, by rebuilding Chapter 2's Table 1 for every covariate in
`ns_covs` rather than the handful used there for illustration.
`ns_covs` mixes two kinds of variables — `age` and `year` are numeric,
everything else is categorical — so we route each kind through its own
helper from `R/table1.R`, casting `year` to numeric first since it
arrives from NAMCS as a factor.

```{r ch06-table1-continuous}
source("R/table1.R")
ns_year_num <- ns |> mutate(year = as.numeric(as.character(year)))
table1_continuous(
  ns_year_num,
  vars = c("age", "year"),
  by = "cox2_initiation"
) |> knitr::kable(
  col.names = c("Variable", "Nonselective (No)", "Cox-2 (Yes)"),
  caption = "Table 1, continuous covariates, ns cohort by cox2_initiation"
)
```

```{r ch06-table1-categorical}
table1_categorical(
  ns,
  vars = setdiff(ns_covs, c("age", "year")),
  by = "cox2_initiation"
) |> knitr::kable(
  col.names = c("Variable", "Level", "Nonselective (No)", "Cox-2 (Yes)"),
  caption = "Table 1, categorical covariates, ns cohort by cox2_initiation"
)
```

```{r ch06-table1-values, include=FALSE}
gap_pct <- function(data, var, level = "Yes", by = "cox2_initiation") {
  data |>
    group_by(.data[[by]]) |>
    summarise(p = 100 * mean(.data[[var]] == level, na.rm = TRUE), .groups = "drop") |>
    pull(p)
}
age_grp   <- ns |> group_by(cox2_initiation) |> summarise(m = mean(age), .groups = "drop") |> pull(m)
arth_grp  <- gap_pct(ns, "arthritis")
htn_grp   <- gap_pct(ns, "hypertension")
diab_grp  <- gap_pct(ns, "diabetes")
osteo_grp <- gap_pct(ns, "osteoporosis")
cort_grp  <- gap_pct(ns, "corticosteroid_use")
ppi_grp   <- gap_pct(ns, "ppi_use")
sex_grp   <- gap_pct(ns, "sex", level = "Female")
south_grp <- gap_pct(ns, "region", level = "South")
year_grp  <- ns_year_num |> group_by(cox2_initiation) |> summarise(m = mean(year), .groups = "drop") |> pull(m)
```

Even with the comparator restricted to a shared class of drugs, several
covariates still separate the two arms by a wide margin. The largest
gap is arthritis: `r sprintf("%.1f", arth_grp[1])`% of nonselective
initiators carry an arthritis diagnosis against
`r sprintf("%.1f", arth_grp[2])`% of Cox-2 initiators, a gap of
`r sprintf("%.0f", arth_grp[2] - arth_grp[1])` percentage points, larger
than any other in either table. Age moves the same way — Cox-2
initiators average `r sprintf("%.1f", age_grp[2])` years against
`r sprintf("%.1f", age_grp[1])` among nonselective initiators, a gap of
roughly `r sprintf("%.0f", age_grp[2] - age_grp[1])` years — and
corticosteroid use, though rare overall, is more than four times as
common among Cox-2 initiators (`r sprintf("%.1f", cort_grp[2])`% versus
`r sprintf("%.1f", cort_grp[1])`%). Hypertension, diabetes, osteoporosis,
and PPI use all run a few points higher among Cox-2 initiators as well
(`r sprintf("%.1f", htn_grp[1])`% vs `r sprintf("%.1f", htn_grp[2])`%,
`r sprintf("%.1f", diab_grp[1])`% vs `r sprintf("%.1f", diab_grp[2])`%,
`r sprintf("%.1f", osteo_grp[1])`% vs `r sprintf("%.1f", osteo_grp[2])`%,
and `r sprintf("%.1f", ppi_grp[1])`% vs `r sprintf("%.1f", ppi_grp[2])`%).

These imbalances are not arbitrary. Arthritis and the comorbidities that
accompany it — hypertension, diabetes, osteoporosis, PPI use — are what
stands behind the `gi_risk` node in the diagram, while age and
corticosteroid use correspond to the diagram's `age` and `steroids`
nodes. Each of these three confounders has an arrow into both `cox2` and
`pud` in Chapter 5's diagram, and each appears here as a substantial
imbalance. Sex (`r sprintf("%.1f", sex_grp[1])`% female
vs `r sprintf("%.1f", sex_grp[2])`%) and calendar year
(`r sprintf("%.1f", year_grp[1])` vs `r sprintf("%.1f", year_grp[2])`)
sit close to balanced, which fits the diagram: neither has an arrow
into `cox2`. Region fits neither pattern:
`r sprintf("%.1f", south_grp[1])`% of nonselective initiators are from
the South against `r sprintf("%.1f", south_grp[2])`% of Cox-2
initiators, a gap of `r sprintf("%.0f", south_grp[2] - south_grp[1])`
points comparable to that for hypertension, although region has no node
in Chapter 5's diagram and no obvious clinical mechanism ties geography
to drug choice. An imbalance in Table 1 is not by itself evidence of a
confounding path; this one probably reflects regional prescribing habits
rather than confounding by GI risk, although it warrants scrutiny rather
than being dismissed as either explained or safely ignorable. The gaps
that do map onto the diagram — arthritis and the conditions that
accompany it, age, and corticosteroid use — are the ones that Part II's
adjustment methods are meant to close.

## Immortal time and other time-zero problems

A design can specify eligibility and the comparator correctly and still
fail on a third dimension: when each patient's follow-up clock starts.
**Aligned time zero** means that three things happen on the same date
for every patient: the date they are confirmed eligible, the date they
are classified into a treatment strategy, and the date their
person-time starts counting toward the outcome. `ns` satisfies this
automatically, because eligibility, drug class, and the start of
follow-up are all pinned to the same qualifying visit. The requirement
is easy to state, but it is violated whenever a study defines "exposed"
using information that could only be known after follow-up had already
begun.

A worked hypothetical shows how the violation operates. It is built
around a definition of treatment that is common in
pharmacoepidemiology: a patient is called a "user" only once a second
prescription has been filled, some grace period after the first.
Suppose we simulate 20,000 hypothetical patients, each with a true time
to death that is, by construction, unrelated to anything about their
future drug-taking behavior, and suppose "persistent user" status
requires two things: surviving past a 90-day grace period, and an
independent adherence habit that is fixed for each patient and has no
effect on survival.

```{r ch06-immortal-sim}
set.seed(6)
n <- 20000
grace <- 90
follow_up <- 730

hyp <- tibble(
  id = 1:n,
  death_time = rexp(n, rate = 1 / 2000),
  adherent   = rbinom(n, 1, 0.6)   # a habit unrelated to survival, by construction
) |>
  mutate(
    alive_grace = death_time > grace,
    persistent  = as.integer(alive_grace & adherent == 1),
    died_by_end = as.integer(death_time <= follow_up)
  )
```

A naive analysis, of the kind a misaligned time zero produces, starts
every patient's clock on day zero and compares mortality risk
between "persistent users" and everyone else over the full
`r follow_up`-day window:

```{r ch06-immortal-naive}
naive_tab <- hyp |>
  group_by(persistent) |>
  summarise(n = n(), risk = mean(died_by_end), .groups = "drop")
naive_tab |> knitr::kable(digits = 3)
```

```{r ch06-immortal-values, include=FALSE}
naive_diff <- naive_tab$risk[naive_tab$persistent == 0] -
  naive_tab$risk[naive_tab$persistent == 1]

landmark_tab <- hyp |>
  filter(alive_grace) |>
  group_by(adherent) |>
  summarise(n = n(), risk = mean(died_by_end), .groups = "drop")
landmark_diff <- landmark_tab$risk[landmark_tab$adherent == 1] -
  landmark_tab$risk[landmark_tab$adherent == 0]
```

Persistent users appear strongly protected, with mortality risk
`r sprintf("%.1f", 100 * naive_diff)` percentage points lower than
everyone else, even though adherence was given no causal effect on
survival in the simulation. The explanation is mechanical rather than
clinical. To be labeled "persistent," a patient had to survive at least
`r grace` days, so the persistent group begins its
`r follow_up`-day follow-up with `r grace` days during which a death
could not have occurred, whatever the composition of that group. This
guaranteed event-free interval is **immortal time**, and it arises
entirely because the exposure definition depends on information —
survival through the grace period — that was not available at the true
time zero, the date treatment actually started.

The remedy is to give every patient the same time zero regardless of
group, restricting the comparison to patients who survived to the end of
the grace period and starting follow-up at day
`r grace` rather than at day zero.

```{r ch06-immortal-landmark}
landmark_tab |> knitr::kable(digits = 3)
```

Among the `r sum(hyp$alive_grace)` patients who survived to day
`r grace`, adherent and non-adherent patients now have essentially
identical mortality risk over the remaining follow-up
(`r sprintf("%.3f", landmark_tab$risk[landmark_tab$adherent == 1])`
versus `r sprintf("%.3f", landmark_tab$risk[landmark_tab$adherent == 0])`,
a difference of `r sprintf("%.3f", landmark_diff)`) — indistinguishable
from the null effect we built into the simulation. Nothing about any
patient differed between the two analyses. Only the point at which
follow-up was allowed to begin changed, and that change alone was
enough to produce, and then remove, a seven-point "protective" effect.

Chapter 4 flagged this hazard for the `sta` cohort, where any
comparison built around remaining on a statin for some minimum duration
is vulnerable to the same mechanism, and left for this chapter the
question of which row of the target trial table is violated. It is the
**Follow-up** row, which specifies when person-time starts accruing
toward the outcome. Immortal time
bias arises whenever the Assignment row — how a patient gets classified
into a treatment strategy — is allowed to depend on something observed
after the Follow-up row's start date, rather than at it. A short
checklist captures the requirement for any design in this book,
`ns`, `sta`, or a design of your own:

| Component | Must be fixed as of the same date as... |
|---|---|
| Eligibility | the date all inclusion criteria are first met |
| Assignment | the date the patient's treatment strategy is classified |
| Start of follow-up | the date person-time begins accruing toward the outcome |

When those three dates coincide, as they do in `ns` by construction,
immortal time cannot arise, because no interval separates the point at
which a patient becomes eligible from the start of follow-up in which
later information could influence the treatment label. If they are
allowed to drift apart, as in the persistent-user hypothetical above,
the resulting comparison can produce a spurious association or conceal a
real one, and no adjustment method in Part II will correct it. This is
much of what specifying a target trial buys: making eligibility,
assignment, and the start of follow-up coincide forecloses immortal time
bias rather than leaving it to be corrected after the fact, an argument
developed by Hernán and colleagues (2016, *Journal of Clinical
Epidemiology*).

## Exercises

1. Using what `R/setup.R` tells you about `sta` — that it follows
   patients who did or did not initiate a statin, with time variables
   `cv_time`, `death_time`, and `loss_fu_time` measured from cohort
   entry — describe whether `sta` follows a new-user or a
   prevalent-user design, and state what its time zero is. Would your
   answer change if `sta` had instead been built by sampling patients
   already filling statin prescriptions at some arbitrary visit, rather
   than at their first qualifying prescription?
2. Look back at the two Table 1 chunks built from `ns_covs` in this
   chapter. Besides arthritis and age, name two more covariates where
   the gap between Cox-2 and nonselective initiators is largest, and
   propose a clinical reason each one might still differ even after
   restricting the comparison to two active NSAID options.
3. `sta` currently compares statin initiators to patients who did not
   initiate a statin at all — not an active-comparator design. Propose
   a plausible active comparator for the statin question (a different
   medication prescribed for a similar indication, initiated at a
   similar point in a patient's care), and discuss one advantage and
   one disadvantage of switching to it, using the reasoning this
   chapter used for Cox-2 versus nonselective NSAIDs.
