# (PART) Core Methods {-}

# Outcome regression and standardization

```{r setup-ch07, include=FALSE}
source("R/setup.R")
```

Part I built the vocabulary: potential outcomes and estimands in
Chapter 3, the target trial in Chapter 4, a diagram and an adjustment
set in Chapter 5, and a design that earns the right to compare two arms
in Chapter 6. What Part I did not do was produce an estimate, and that
is the work of Part II.
Over the next three chapters we answer the question we have been
circling since Chapter 1 — does initiating a Cox-2 selective inhibitor,
rather than a nonselective NSAID, change the risk of peptic ulcer
disease? — three separate times, with three estimators that make the
same identification assumptions and reach the same estimand by
different routes: standardization here, propensity scores in Chapter 8,
and inverse probability of treatment weighting in Chapter 9. The three
are not competing philosophies. Each uses the covariates in `ns_covs` to
address the imbalance documented in Chapter 6's Table 1, and one reason
to run all three is to see how little the answer moves when the
estimator changes. We also start building the toolkit promised
in the preface, with the book's first estimator function,
`standardize()`.

## Adjusting with regression

Let us start where most epidemiologic analyses start, with a regression
model for the outcome. Chapter 6's Table 1 told us that Cox-2
initiators are older, sicker, and more likely to be on corticosteroids
than nonselective initiators, so the crude comparison of PUD risk
between arms is confounded — and confounded in a direction we can
anticipate, since patients were directed toward the Cox-2 drug partly
*because* their physicians judged them to be at higher GI risk.

```{r ch07-crude}
ns |>
  group_by(cox2_init) |>
  summarise(n = n(), cases = sum(pud), risk = mean(pud), .groups = "drop") |>
  knitr::kable(digits = 3)
```

```{r ch07-crude-values, include=FALSE}
crude <- ns |> group_by(cox2_init) |> summarise(r = mean(pud), .groups = "drop") |> pull(r)
crude_rd <- crude[2] - crude[1]
crude_rr <- crude[2] / crude[1]
crude_or <- exp(coef(glm(pud ~ cox2_init, family = binomial(), data = ns))[["cox2_init"]])
n_cox2 <- sum(ns$cox2_init)
```

The crude risks run `r sprintf("%.1f", 100 * crude[2])`% in the Cox-2 arm
against `r sprintf("%.1f", 100 * crude[1])`% among nonselective
initiators — a crude risk difference of
`r sprintf("%+.1f", 100 * crude_rd)` percentage points and a crude risk
ratio of `r sprintf("%.2f", crude_rr)`. Taken at face value, this
suggests that the Cox-2 drug *raises* ulcer risk, which runs against its
clinical rationale. It is a familiar pattern of confounding by
indication, and it is the crude association that adjustment has to work
on.

The standard first step is a logistic model for `pud` that includes the
treatment indicator alongside every covariate in the adjustment set.
Rather than typing nineteen covariate names, we let `reformulate()`
assemble the formula from `ns_covs`, which keeps `R/setup.R` as the
single source of truth for what "adjusted" means in this book.

```{r ch07-outcome-model}
library(broom)

fit <- glm(
  reformulate(c("cox2_init", ns_covs), "pud"),
  family = binomial(),
  data = ns
)

tidy(fit, exponentiate = TRUE, conf.int = TRUE) |>
  filter(term %in% c("cox2_init", "age", "arthritisYes", "corticosteroid_useYes")) |>
  knitr::kable(digits = 3)
```

```{r ch07-model-values, include=FALSE}
tt <- tidy(fit, exponentiate = TRUE, conf.int = TRUE)
cond_or    <- tt$estimate[tt$term == "cox2_init"]
cond_or_lo <- tt$conf.low[tt$term == "cox2_init"]
cond_or_hi <- tt$conf.high[tt$term == "cox2_init"]
age_or     <- tt$estimate[tt$term == "age"]
```

Adjustment moves the estimate as expected. The coefficient on
`cox2_init`,
exponentiated, is an odds ratio of `r sprintf("%.2f", cond_or)` (95% CI
`r sprintf("%.2f", cond_or_lo)` to `r sprintf("%.2f", cond_or_hi)`) —
a shift from the crude odds ratio of `r sprintf("%.2f", crude_or)` to just
below the null, though with an interval far too wide to rule out either
a modest benefit or a modest harm. Age carries an odds ratio of
`r sprintf("%.2f", age_or)` per year, and accounts for much of the
confounding.

There is a complication, however. That `r sprintf("%.2f", cond_or)` is a
**conditional** odds ratio: the odds ratio comparing two patients who
are identical on all nineteen covariates and differ only in which NSAID
they started. It is not the marginal contrast we defined in Chapter 3.
The ATE was written $E[Y(1)] - E[Y(0)]$, an average over the whole
population of what would happen under each treatment, and therefore a
quantity attached to a population rather than to a stratum. No
coefficient in the regression provides it, and the discrepancy is not
merely a matter of rescaling from odds to risks.

The reason is **noncollapsibility**. If treatment truly has no effect,
the conditional and marginal odds ratios agree at 1. Otherwise, even
with no confounding at all, the odds ratio averaged over strata is
pulled toward the null relative to the odds ratio within any stratum,
simply because the odds scale is nonlinear and the strata have different
baseline risks. The conditional odds ratio and the marginal odds ratio
are answering different questions, and the difference between them does
not indicate bias in either; it is a property of the scale. Greenland,
Robins, and Pearl (1999) are the standard reference for keeping this
distinction straight, and for the point that noncollapsibility and
confounding are different phenomena that the same numerical gap can
express. Risk differences and risk ratios do not behave this way; they
collapse. Since the difficulty lies in reading the estimand off a
coefficient rather than in the model itself, we can keep the fitted
model and extract the marginal contrast from it directly, which is what
standardization does.

## Standardization: predict, then average

The recipe, in words, is four steps. Fit an outcome model that includes
treatment and the covariates. Then take the entire cohort and set every
patient's treatment to 1 — including the patients who actually took the
nonselective drug — and ask the model to predict each patient's risk in
that hypothetical world. Take the whole cohort again, set everyone's
treatment to 0, and predict again. Average each set of predictions and
contrast the two averages. This is the **parametric g-formula**, due to
Robins (1986): the model is used only as a device for interpolating what
each individual's risk would have been under each treatment, and the
averaging — over the observed covariate distribution of the whole
cohort — is what turns those individual predictions into a marginal,
population-level contrast.

In code, the entire recipe takes five lines.

```{r ch07-hand-rolled}
ns1 <- ns |> mutate(cox2_init = 1)
ns0 <- ns |> mutate(cox2_init = 0)
risk1 <- mean(predict(fit, newdata = ns1, type = "response"))
risk0 <- mean(predict(fit, newdata = ns0, type = "response"))
c(risk0 = risk0, risk1 = risk1, rd = risk1 - risk0, rr = risk1 / risk0)
```

Notice what the two copies of the data are doing. `ns1` is the cohort as
it would look if every one of these `r nrow(ns)` patients had started a
Cox-2 inhibitor, with every covariate left exactly as observed; `ns0` is
the same cohort under the nonselective strategy. Because both averages
are taken over the *same* covariate distribution — every patient, once,
in each world — the difference between them cannot be attributed to a
difference in case mix. This is the same logic as classical direct
standardization to a reference population, with the fitted model
standing in for the stratum-specific rates.

Every step of that reasoning rests on Chapter 3's identification
conditions. Consistency lets us read a predicted risk under
`cox2_init = 1` as a potential outcome; conditional exchangeability,
which Chapter 5's diagram is our argument for, is what makes a model fit
partly to treated patients usable for predicting untreated ones.
Positivity bears on whether the resulting predictions can be trusted:
with only `r n_cox2` Cox-2 initiators among `r nrow(ns)` patients, the
model extrapolates substantially when it predicts a Cox-2 world for
covariate patterns in which almost no Cox-2 initiator was observed.
Standardization will return an estimate regardless, which is a reason to
keep Chapter 6's Table 1 in view rather than relying on the estimator
alone.

We will want this estimator again in Chapters 8 and 9, so let us
package it.
The function takes the data, a model formula, and the name of the
treatment column, and returns one row per estimand. Two details are
worth pausing on. The `"{treat}" := 1` syntax inside `mutate()` is
`dplyr`'s glue notation for assigning to a column whose name is held in
a variable, which is what lets the same function work for `cox2_init`
here and `statin` later. And the point estimates are wrapped in their
own inner function, `point_estimates()`, because the bootstrap in the
next section needs to run exactly the same computation on a resampled
data set — fitting the model *inside* the resample, not reusing the
original fit.

```{r ch07-standardize-def, eval=FALSE}
standardize <- function(data, outcome_model, treat, family = binomial(),
                        n_boot = 200, conf_level = 0.95, seed = NULL) {
  if (!is.null(seed)) set.seed(seed)

  point_estimates <- function(d) {
    fit <- glm(outcome_model, family = family, data = d)
    d1 <- d |> mutate("{treat}" := 1)
    d0 <- d |> mutate("{treat}" := 0)
    risk1 <- mean(predict(fit, newdata = d1, type = "response"))
    risk0 <- mean(predict(fit, newdata = d0, type = "response"))
    c(risk0 = risk0, risk1 = risk1, rd = risk1 - risk0, rr = risk1 / risk0)
  }

  est <- point_estimates(data)
  boot_est <- replicate(n_boot, {
    point_estimates(slice_sample(data, prop = 1, replace = TRUE))
  })
  alpha <- 1 - conf_level

  tibble(
    estimand  = names(est),
    estimate  = unname(est),
    conf_low  = apply(boot_est, 1, quantile, probs = alpha / 2),
    conf_high = apply(boot_est, 1, quantile, probs = 1 - alpha / 2)
  )
}
```

That code lives in `R/standardize.R`, so from here on we source it
rather than retype it.

```{r standardize}
source("R/standardize.R")
std <- standardize(
  ns,
  outcome_model = reformulate(c("cox2_init", ns_covs), "pud"),
  treat = "cox2_init", n_boot = 200, seed = 921
)
std |> knitr::kable(digits = 3)
```

```{r ch07-std-values, include=FALSE}
g <- function(what, col = "estimate") std[[col]][std$estimand == what]
std_r0 <- g("risk0"); std_r1 <- g("risk1")
std_rd <- g("rd"); std_rr <- g("rr")
marg_or <- (std_r1 / (1 - std_r1)) / (std_r0 / (1 - std_r0))
```

Here is the estimate, on the scale Chapter 3 called for. Had every patient
in `ns` started a nonselective NSAID, the model puts PUD risk at
`r sprintf("%.1f", 100 * std_r0)`%; had every one of them started a
Cox-2 inhibitor, `r sprintf("%.1f", 100 * std_r1)`%. The standardized
risk difference is `r sprintf("%+.1f", 100 * std_rd)` percentage points
(95% CI `r sprintf("%.1f", 100 * g("rd", "conf_low"))` to
`r sprintf("%+.1f", 100 * g("rd", "conf_high"))`) and the standardized
risk ratio `r sprintf("%.2f", std_rr)`
(`r sprintf("%.2f", g("rr", "conf_low"))` to
`r sprintf("%.2f", g("rr", "conf_high"))`). Adjustment has moved the
crude risk ratio of `r sprintf("%.2f", crude_rr)` to
`r sprintf("%.2f", std_rr)` — from apparent harm to a point estimate
slightly on the protective side of the null — but the interval
comfortably includes 1, so these data support no more than the absence of
a large effect in either direction, and certainly not a benefit. With
`r sum(ns$pud)` ulcer cases and `r n_cox2` Cox-2 initiators, this is
close to the precision the cohort can provide.

Note also what noncollapsibility looks like in the arithmetic. The
standardized risks imply a *marginal* odds ratio of
`r sprintf("%.3f", marg_or)`, while the regression coefficient gave a
conditional odds ratio of `r sprintf("%.3f", cond_or)`, the conditional
one sitting slightly further from the null, as expected. The
gap is small here only because PUD is uncommon; with a common outcome it
can be substantial.

## The bootstrap

The point estimates came out of a `glm()` fit, so it is tempting to
reach for that model's standard errors, but they do not apply here. The
standardized risk difference is not a parameter of the logistic model at
all — it is a nonlinear function of *every* coefficient in the model,
combined with the empirical covariate distribution of the cohort. There
is no line in the `summary()` output whose standard error describes it,
and the delta-method calculation that would produce one by hand is
tedious and specification-specific.

Resampling avoids the algebra. The sampling variability of our estimate
arises because `ns` is one draw of
`r nrow(ns)` patients from some larger population, so we mimic that draw
by sampling `r nrow(ns)` patients from `ns` *with replacement*, and
recompute the whole estimator — refit the model, re-predict both worlds,
re-average, re-contrast — on each resample. The spread of those
recomputed estimates approximates the spread we would see across
repeated samples from the population, and the 2.5th and 97.5th
percentiles of that spread give a 95% percentile confidence interval.
It is important that everything downstream of sampling be redone within
each replicate, model fitting included; bootstrapping
only the averaging step, with the coefficients held fixed, would ignore
the largest source of uncertainty in the procedure.

`n_boot` trades precision for time. Each replicate is a full `glm()` fit
on `r nrow(ns)` rows, and our default of 200 is enough to place the
interval approximately while keeping the build time tolerable for a book
that re-renders often. The percentile endpoints are themselves noisy at that size,
though, so a different seed will shift them in the second decimal place;
for a result you intend to publish, 1,000 or 2,000 is the right order of
magnitude, and `seed` is what makes whichever number you choose
reproducible.

## Effect modification

So far we have estimated a single effect for the whole cohort. However,
the effect of a Cox-2 inhibitor might plausibly differ by whether a patient is also
taking aspirin, since aspirin damages gastric mucosa through the same
prostaglandin pathway the Cox-2 drug was designed to spare. When a
causal effect differs across levels of a covariate, we call that
**effect modification** (or heterogeneity of treatment effect), and we
look for it by letting the treatment coefficient depend on that
covariate — an interaction term.

```{r ch07-interaction}
fit_int <- glm(
  reformulate(c("cox2_init", ns_covs, "cox2_init:aspirin_use"), "pud"),
  family = binomial(), data = ns
)

tidy(fit_int, exponentiate = TRUE, conf.int = TRUE) |>
  filter(str_detect(term, "cox2_init")) |>
  knitr::kable(digits = 3)
```

```{r ch07-int-values, include=FALSE}
ti <- tidy(fit_int, exponentiate = TRUE, conf.int = TRUE)
int_or <- ti$estimate[ti$term == "cox2_init:aspirin_useYes"]
int_p  <- ti$p.value[ti$term == "cox2_init:aspirin_useYes"]
n_asa  <- sum(ns$aspirin_use == "Yes")
n_asa_cox2 <- sum(ns$aspirin_use == "Yes" & ns$cox2_init == 1)
```

The interaction term is a ratio of odds ratios,
`r sprintf("%.2f", int_or)` with a p-value of
`r sprintf("%.2f", int_p)` — no evidence of modification, which is
unsurprising given that only `r n_asa` of the `r nrow(ns)` patients use
aspirin and only `r n_asa_cox2` of those started a Cox-2 inhibitor. The
coefficient is in any case not the quantity of interest, for the reason
given above: it is a conditional, noncollapsible quantity on the odds
scale. The subgroup effects of interest are two standardized risk
differences, one within each aspirin stratum.

Standardizing within a subgroup means changing the population we average
over: instead of predicting both worlds for the whole cohort, we predict
both worlds for the aspirin users only, then repeat for the non-users.
Fitting a separate model within each stratum is the fully flexible
version of the interaction above — it lets every coefficient, not just
treatment's, differ by aspirin use — so we can get subgroup effects by
calling `standardize()` on each subset. One adjustment is forced on us:
`aspirin_use` is constant within a stratum by construction and must
leave the covariate list, and among aspirin users
`anti_coagulant_use` turns out to be constant too. With only
`r n_asa` aspirin users we cannot support nineteen covariates in that
stratum in any case. We fit a deliberately lean model in both strata so the
two are comparable.

```{r ch07-subgroups, warning=FALSE}
lean <- reformulate(c("cox2_init", "age", "sex", "arthritis"), "pud")

sub <- bind_rows(
  standardize(filter(ns, aspirin_use == "No"),  lean, "cox2_init", seed = 922) |>
    mutate(aspirin = "No", .before = 1),
  standardize(filter(ns, aspirin_use == "Yes"), lean, "cox2_init", seed = 923) |>
    mutate(aspirin = "Yes", .before = 1)
)
sub |> knitr::kable(digits = 3)
```

```{r ch07-sub-values, include=FALSE}
gs <- function(a, what, col = "estimate") {
  sub[[col]][sub$aspirin == a & sub$estimand == what]
}
```

Aspirin users start from a much higher baseline: a standardized
nonselective-NSAID risk of `r sprintf("%.1f", 100 * gs("Yes", "risk0"))`%
against `r sprintf("%.1f", 100 * gs("No", "risk0"))`% among non-users,
which is consistent with the pathway described above. The standardized risk
differences, though, are
`r sprintf("%+.1f", 100 * gs("Yes", "rd"))` points among aspirin users
and `r sprintf("%+.1f", 100 * gs("No", "rd"))` points among non-users,
and the aspirin-user interval runs from
`r sprintf("%.1f", 100 * gs("Yes", "rd", "conf_low"))` to
`r sprintf("%+.1f", 100 * gs("Yes", "rd", "conf_high"))` percentage
points — wide enough to contain nearly any clinically meaningful effect
in either direction. The risk ratio row is worse still, with a lower
bound at essentially zero: in some resamples the handful of
aspirin-using Cox-2 initiators produced a predicted risk near zero, and
the bootstrap reports that instability faithfully. The wide interval
reflects the sparsity of the data rather than anything about the drug. We
suppressed R's separation warnings in that chunk to keep the output
readable, although they convey the same information.

It is worth being precise about how effect modification differs from
confounding, because both are described loosely as "a covariate that
matters." Confounding is a problem of identification: a common cause of
treatment and outcome makes the observed contrast fail to equal the
causal one, and we adjust for it to recover a valid estimate. Effect
modification, in contrast, is not a threat to validity; it is a feature
of the causal effect itself. A confounder must be addressed or the
estimate will be biased, whereas an effect modifier is something we
choose to report, and the estimand changes when we do, from one ATE for
the cohort to a set of subgroup-specific ATEs. A variable can be one,
the other, or both:
aspirin use here is a candidate for either role. Note also that the
overall standardized risk difference of
`r sprintf("%+.1f", 100 * std_rd)` points is a weighted average of the
subgroup effects, weighted by how common each subgroup is — which is why
a cohort-level ATE can be small while an effect in a small subgroup is
large, and why the population one standardizes to is part of the
estimand rather than an implementation choice.

## Exercises

1. The model above enters `age` linearly on the log-odds scale. Refit it
   with `poly(age, 2)` in place of `age` and rerun `standardize()` with
   the same seed. How much do the standardized risks and the risk
   difference move? Does the confidence interval change enough to alter
   what you would conclude, and what does your answer suggest about how
   sensitive this estimate is to the functional form of the strongest
   confounder?
2. `ppi_use` and `h2_antagonist_use` are gastroprotective drugs, so they
   may be markers of a physician's concern about GI risk rather than
   causes of it. Drop both from the covariate list and rerun
   `standardize()`. Report the standardized risk difference with and
   without them. This is an informal sensitivity analysis to the choice
   of adjustment set; using Chapter 5's language, sketch the DAG under
   which including them is right and the DAG under which it is wrong.
3. Everything in this chapter estimated the ATE, averaging predictions
   over the whole cohort. The ATT averages over the treated only.
   Modify `point_estimates()` inside a copy of `standardize()` so that
   the two predicted risks are averaged over `filter(d, cox2_init == 1)`
   while the model is still fit on all of `d`, and estimate the ATT for
   the Cox-2 question. Why must the *model* still be fit on the full
   data even though the *averaging* is restricted, and would you expect
   the ATT here to sit above or below the ATE given Chapter 6's Table 1?
