# Propensity scores

```{r setup-ch08, include=FALSE}
source("R/setup.R")
```

Chapter 7 adjusted for confounding by modeling the outcome. We wrote
down a logistic regression for `pud` that included the treatment
indicator and all nineteen covariates in `ns_covs`, then averaged its
predictions over the cohort twice. It worked, but notice how much weight
that single model was carrying: nineteen covariates and their functional
forms, all of it resting on `r sum(ns$pud)` ulcer cases. This chapter
takes a second route to the same question, and it moves the modeling
burden to the other side of the problem. Instead of modeling how
covariates produce the outcome, we model how covariates produce
*treatment*, and then use that one fitted probability — rather than the
nineteen covariates themselves — to make the arms comparable. In many
studies the appeal is arithmetic, because exposure is far more common
than the outcome and the treatment model therefore has more information
to work with; here, with `r sum(ns$cox2_init)` Cox-2 initiators against
`r sum(ns$pud)` cases, the two are comparably rare and the appeal is
conceptual instead. What we gain is a separation between the work of
making groups comparable and the work of estimating an effect.

## One number that summarizes confounding

The **propensity score** is the probability of receiving treatment given
the measured covariates,
$$e(L) = P(A = 1 \mid L),$$
a single number between 0 and 1 attached to each patient. In our cohort
it is the probability that a patient with this age, sex, region,
comorbidity profile, and medication history was started on a Cox-2
selective inhibitor rather than a nonselective NSAID.

What makes this useful is a result of Rosenbaum and Rubin (1983): the
propensity score is a **balancing score**. Among patients who share the
same value of $e(L)$, the distribution of $L$ itself is the same in the
treated and the untreated. Two patients with a propensity score of 0.20
may look nothing alike — one old with arthritis and no other illness,
the other younger with diabetes and hypertension — but if we collect
*everyone* whose score is 0.20, the treated members of that collection
and the untreated members will have, on average, the same age
distribution, the same arthritis prevalence, the same steroid use, and
so on for every covariate that went into the score. Within a level of
the score, treatment is as though randomized with respect to those
covariates.

The consequence for us is the second half of the theorem. If
conditional exchangeability holds given the full covariate vector $L$ —
Chapter 3's assumption, for which Chapter 5's diagram is the argument —
then it also holds given $e(L)$ alone. So we may condition on one
scalar in place of nineteen variables and lose nothing. Nineteen
dimensions of confounding collapse into one number.

Two cautions belong here at the outset. First, the propensity score is
a **nuisance parameter**: nobody wants to know a patient's probability
of getting a Cox-2 inhibitor for its own sake, and we should resist
interpreting the treatment model's coefficients causally. It exists only
to be conditioned on. Second, the balancing property is a property of
the *true* score, which we do not have. We estimate it, and the honest
justification for the estimate is not that the model is correctly
specified but that it demonstrably balances the covariates — something
we can check directly, and will.

## Estimating the score

Estimating a propensity score is a logistic regression like any other,
so the function that does it is short. It fits the treatment model and
returns the data with the fitted probabilities attached as a column
named `ps`, which is what every method later in the chapter consumes.

```{r ch08-est-ps-def, eval=FALSE}
est_ps <- function(data, treat_model) {
  fit <- glm(treat_model, family = binomial(), data = data)
  data |> mutate(ps = predict(fit, type = "response"))
}
```

`predict()` with `type = "response"` is what puts the predictions on the
probability scale rather than the log-odds scale; without it we would
get linear predictors, which balance just as well but are harder to
read. That code and the matching function below live in
`R/propensity.R`, so from here on we source it.

```{r ch08-est-ps}
source("R/propensity.R")

ps_model <- reformulate(ns_covs, "cox2_init")
ns_ps <- est_ps(ns, ps_model) |> mutate(id = row_number())

ns_ps |>
  group_by(cox2_init) |>
  summarise(n = n(), mean_ps = mean(ps), min = min(ps), max = max(ps),
            .groups = "drop") |>
  knitr::kable(digits = 3)
```

```{r ch08-ps-values, include=FALSE}
ps1 <- ns_ps$ps[ns_ps$cox2_init == 1]
ps0 <- ns_ps$ps[ns_ps$cox2_init == 0]
prev <- mean(ns$cox2_init)
n_ctrl_above <- sum(ps0 > max(ps1))
n_trt_below  <- sum(ps1 < min(ps0))
lo <- ns_ps |> filter(ps < quantile(ps, 0.01))
hi <- ns_ps |> filter(ps > quantile(ps, 0.99))
```

Notice the formula. `reformulate(ns_covs, "cox2_init")` puts treatment
on the left and the covariates on the right — the mirror image of
Chapter 7's outcome model, and with no mention of `pud` anywhere. That
is more than a technicality. The whole model can be built, inspected,
and revised without ever looking at the outcome, so we cannot tune it,
however unconsciously, toward an answer we like. It restores something
outcome modeling gives up: a **design** stage, prior to and separate
from analysis, in which we work only with treatment and covariates. It
is the closest an observational study comes to writing a randomization
scheme. Rubin (2008) makes this the central argument for propensity
scores — that in observational work design trumps analysis, and that the
discipline of settling the comparison before the outcome is ever
examined is most of what the score is for.

The fitted scores average `r sprintf("%.3f", mean(ps1))` in the Cox-2
arm against `r sprintf("%.3f", mean(ps0))` among nonselective
initiators, straddling the cohort-wide treatment prevalence of
`r sprintf("%.1f", 100 * prev)`%. That separation is the confounding of
Chapter 6's Table 1, now expressed as one number: the model can tell the
arms apart because the arms genuinely differ. What matters more is how
much the two distributions overlap.

```{r ch08-overlap}
ggplot(ns_ps, aes(x = ps, fill = factor(cox2_init))) +
  geom_density(alpha = 0.45, colour = NA) +
  scale_fill_discrete(name = NULL,
                      labels = c("Nonselective NSAID", "Cox-2 selective")) +
  labs(x = "Estimated propensity score", y = "Density") +
  theme_minimal()
```

This plot is the practical face of the positivity assumption, and the
first thing to look at before any propensity-score analysis: we want the
region where treated patients live to be a region where untreated
patients also live, and vice versa, because conditioning on the score
only makes sense where both arms are represented. Formally the ranges
nest: the count of nonselective initiators above the highest treated
score is `r n_ctrl_above`, and the count of Cox-2 initiators below the
lowest untreated score is `r n_trt_below`. But the tails are where the
assumption gets thin. The bottom percentile of the score holds
`r nrow(lo)` patients, of whom `r sum(lo$cox2_init)` started a Cox-2
inhibitor; the top percentile holds `r nrow(hi)` patients, of whom
`r sum(hi$cox2_init)` did. The
left tail, then, contains patients for whom we have essentially no
information about what a Cox-2 inhibitor would have done, and the right
tail is thin in the other direction. Chapter 7's standardization ran
into exactly this problem but ran into it silently, extrapolating
wherever it had to. Here the difficulty is visible on a plot, which is
one of the plainer arguments for looking at a propensity score even when
one intends to adjust some other way.

## Stratification

The most direct use of a balancing score is to stratify on it. If
treated and untreated patients within a level of the score are
comparable, then a risk difference computed inside a narrow band of the
score is unconfounded, and we can pool those within-band differences
into one estimate. Bands of exactly equal score would be empty, so in
practice we cut the score into quantiles — quintiles, conventionally.

```{r ch08-strata}
strata <- ns_ps |>
  mutate(ps_q = ntile(ps, 5)) |>
  group_by(ps_q) |>
  summarise(
    n_treat = sum(cox2_init),
    n_ctrl  = sum(1 - cox2_init),
    risk1   = mean(pud[cox2_init == 1]),
    risk0   = mean(pud[cox2_init == 0]),
    rd      = risk1 - risk0,
    var_rd  = risk1 * (1 - risk1) / n_treat + risk0 * (1 - risk0) / n_ctrl,
    .groups = "drop"
  )

strata |> select(-var_rd) |> knitr::kable(digits = 3)
```

```{r ch08-strata-values, include=FALSE}
w <- 1 / strata$var_rd
rd_prec <- sum(w * strata$rd) / sum(w)
se_prec <- sqrt(1 / sum(w))
rd_size <- with(strata, sum((n_treat + n_ctrl) * rd) / sum(n_treat + n_ctrl))
fit_y <- glm(reformulate(c("cox2_init", ns_covs), "pud"),
             family = binomial(), data = ns)
std_rd <- mean(predict(fit_y, mutate(ns, cox2_init = 1), type = "response")) -
  mean(predict(fit_y, mutate(ns, cox2_init = 0), type = "response"))
crude <- tapply(ns$pud, ns$cox2_init, mean)
```

Read down the `risk0` column first. Baseline risk climbs steadily from
`r sprintf("%.1f", 100 * strata$risk0[1])`% in the lowest-propensity
quintile to `r sprintf("%.1f", 100 * strata$risk0[5])`% in the highest,
which is the confounding made concrete: the covariates that pushed
physicians toward the Cox-2 drug are the same covariates that predict
ulcers. Read the `n_treat` column next and the cost of stratification
becomes clear. The lowest quintile holds only
`r strata$n_treat[1]` Cox-2 initiators against
`r strata$n_ctrl[1]` nonselective ones, so its risk difference of
`r sprintf("%+.1f", 100 * strata$rd[1])` percentage points rests on a
handful of patients and should not be taken seriously on its own.

Pooling is therefore essential, and *how* we pool is a choice about the
estimand rather than a computational detail. Weighting each stratum by
the inverse of its variance — precision weighting, the logic behind the
Mantel–Haenszel estimator — gives the most statistically efficient
combination:

```{r ch08-pooled}
c(rd_precision_weighted = rd_prec, se = se_prec)
```

That is `r sprintf("%+.1f", 100 * rd_prec)` percentage points with a
standard error of `r sprintf("%.1f", 100 * se_prec)`, so a 95% interval
running roughly `r sprintf("%.1f", 100 * (rd_prec - 1.96 * se_prec))` to
`r sprintf("%+.1f", 100 * (rd_prec + 1.96 * se_prec))` points. Weighting
each stratum by its *size* instead gives
`r sprintf("%+.1f", 100 * rd_size)` points, considerably closer to
Chapter 7's standardized risk difference of
`r sprintf("%+.1f", 100 * std_rd)`. Neither is wrong, but they answer
different questions. The size-weighted average is standardization to
the cohort's own covariate distribution, and so estimates the ATE of
Chapter 3. The precision-weighted average is standardized to a
population defined by where the information happens to be, which
coincides with the ATE only if the risk difference is constant across
strata. When effects vary, efficiency and interpretability pull apart,
and it is worth being explicit about which one has been bought.

Either way, both sit on the protective side of the null while the crude
risk difference was `r sprintf("%+.1f", 100 * (crude[2] - crude[1]))`
points, and both have intervals wide enough to include no effect.

Does the stratification actually balance anything? Because the score is
a balancing score, covariates should look similar between arms *within*
a stratum even though they differ sharply overall. Chapter 2's
`table1_categorical()` will show us, applied to the top quintile.

```{r ch08-strata-balance}
source("R/table1.R")

top_q <- ns_ps |> mutate(ps_q = ntile(ps, 5)) |> filter(ps_q == 5)
table1_categorical(top_q, c("arthritis", "corticosteroid_use", "hypertension"),
                   by = "cox2_init") |>
  knitr::kable()
table1_continuous(top_q, "age", by = "cox2_init") |> knitr::kable()
```

```{r ch08-balance-values, include=FALSE}
age_gap_all <- diff(tapply(ns$age, ns$cox2_init, mean))
age_gap_q5  <- diff(tapply(top_q$age, top_q$cox2_init, mean))
arth_all <- 100 * tapply(ns$arthritis == "Yes", ns$cox2_init, mean)
```

It does. Across the whole cohort, Cox-2 initiators are
`r sprintf("%.1f", age_gap_all)` years older on average and
`r sprintf("%.0f", diff(arth_all))` percentage points more likely to
carry an arthritis diagnosis. Inside the top propensity quintile the age
gap has shrunk to `r sprintf("%.1f", abs(age_gap_q5))` years — and in
the opposite direction — and the arthritis, steroid, and hypertension
percentages are within a few points of each other. Both arms of this
stratum are high-propensity patients, so both are old and arthritic;
what distinguished them has been absorbed into the stratum itself.

Balance within quintiles is good but not perfect, and it cannot be.
Patients with scores of 0.12 and 0.50 land in the same top quintile
despite being quite different, so some confounding survives inside each
band — Cochran's (1968) classic result is that five strata remove roughly 90%
of the bias from a single continuous confounder, leaving the rest.
Matching addresses that residual directly, by insisting that comparisons
happen between individuals rather than within bands.

## Matching

Matching pairs each treated patient with an untreated patient whose
propensity score is close, and analyzes the paired sample. The version
below is the simplest one that is still respectable: **greedy 1:1
nearest-neighbor matching without replacement, within a caliper**.

```{r ch08-ps-match-def, eval=FALSE}
# Greedy 1:1 nearest-neighbor matching without replacement.
ps_match <- function(data, treat, ps = "ps", caliper = 0.05) {
  treated <- data |> filter(.data[[treat]] == 1) |> arrange(desc(.data[[ps]]))
  controls <- data |> filter(.data[[treat]] == 0)
  used <- rep(FALSE, nrow(controls))
  pairs <- vector("list", nrow(treated))
  for (i in seq_len(nrow(treated))) {
    dist <- abs(controls[[ps]] - treated[[ps]][i])
    dist[used] <- Inf
    j <- which.min(dist)
    if (dist[j] <= caliper) {
      used[j] <- TRUE
      pairs[[i]] <- bind_rows(treated[i, ], controls[j, ])
    }
  }
  bind_rows(pairs, .id = "pair_id")
}
```

Four decisions are encoded in those lines. *Greedy* means we take
treated patients one at a time and give each its closest available
control, never revisiting an earlier choice; it is not guaranteed to
minimize total distance the way optimal matching would, but it is
transparent and fast. *Without replacement* is enforced by the `used`
vector, which sets already-matched controls to infinite distance so no
control appears twice — keeping the paired analysis honest at the cost
of using fewer controls. The `arrange(desc())` orders treated
patients from the highest score down, so the hardest patients to match,
the ones out at the sparse end of the distribution, get first pick of
the control pool. And the *caliper* is the refusal: if the nearest
available control is further than 0.05 on the propensity scale, that
treated patient is left unmatched rather than paired with someone
implausible. Unmatched patients are how a matched analysis reports a
positivity problem instead of extrapolating through one.

```{r ch08-match}
matched <- ps_match(ns_ps, treat = "cox2_init", caliper = 0.05)
n_pairs <- n_distinct(matched$pair_id)
c(pairs = n_pairs, treated_available = sum(ns_ps$cox2_init),
  unmatched = sum(ns_ps$cox2_init) - n_pairs)
```

`r n_pairs` pairs from `r sum(ns_ps$cox2_init)` Cox-2 initiators, a
shortfall of `r sum(ns_ps$cox2_init) - n_pairs`: out at the extreme
right of the density plot sits a treated patient for whom no
nonselective initiator falls inside the caliper. Losing so little is a
consequence of having a
control pool `r sprintf("%.0f", (nrow(ns) - sum(ns$cox2_init)) / sum(ns$cox2_init))`
times the size of the treated group; in a cohort with a more common
exposure the loss would be larger and more consequential.

Now the check that matters. The `cobalt` package computes standardized
mean differences before and after matching and plots them. We pass the
full cohort along with a 0/1 weight flagging which rows survived
matching, which lets `bal.tab()` compute both columns at once.

```{r ch08-baltab, message=FALSE}
library(cobalt)

ns_ps <- ns_ps |> mutate(matched_wt = as.integer(id %in% matched$id))

bt <- bal.tab(ps_model, data = ns_ps, weights = ns_ps$matched_wt,
              estimand = "ATT", un = TRUE, binary = "std",
              thresholds = c(m = 0.1))
bt
```

```{r ch08-bal-values, include=FALSE}
b <- bt$Balance
smd_un  <- abs(b$Diff.Un)
smd_adj <- abs(b$Diff.Adj)
worst_un <- rownames(b)[which.max(smd_un)]
cad <- table(matched$cox2_init, matched$coronory_artery_disease == "Yes")
```

```{r ch08-loveplot, fig.height=6}
love.plot(bt, abs = TRUE, var.order = "unadjusted",
          thresholds = c(m = 0.1),
          sample.names = c("Before matching", "After matching")) +
  theme(legend.position = "bottom")
```

The love plot is the standard way to present this, and it reads at a
glance: each covariate is a row, absolute standardized mean difference
runs along the horizontal axis, and the dashed line at 0.1 marks the
conventional threshold below which imbalance is usually called
negligible. Before matching, `r sum(smd_un > 0.1)` of the
`r nrow(b)` terms exceed that line, led by `r worst_un` at
`r sprintf("%.2f", max(smd_un))`. After matching, most collapse toward
zero and `r worst_un` falls to
`r sprintf("%.2f", smd_adj[which.max(smd_un)])`; the average absolute
standardized difference drops from `r sprintf("%.3f", mean(smd_un))` to
`r sprintf("%.3f", mean(smd_adj))`.

The exceptions deserve attention rather than concealment, since
`r sum(smd_adj > 0.1)` terms still sit above the threshold after
matching, the worst of them the coronary artery disease indicator at
`r sprintf("%.2f", max(smd_adj))`. Look at the counts, though: coronary
artery disease appears in `r cad[1, 2]` matched controls and
`r cad[2, 2]` matched Cox-2 initiators. A gap of
`r cad[1, 2] - cad[2, 2]` patients out of
`r n_pairs` is a raw difference of about
`r sprintf("%.1f", 100 * (cad[1, 2] - cad[2, 2]) / n_pairs)` percentage
points; it looks large as a standardized difference only because
dividing by the standard deviation of a rare binary variable inflates
small absolute differences. This is a general hazard of reading the 0.1
rule mechanically for rare covariates, and the remedy is to look at both
the standardized and the raw scale before deciding whether an imbalance
is worth acting on.

With balance established, the effect estimate is a comparison of two
proportions in the matched sample.

```{r ch08-match-effect}
matched |>
  group_by(cox2_init) |>
  summarise(n = n(), cases = sum(pud), risk = mean(pud), .groups = "drop") |>
  knitr::kable(digits = 3)
```

```{r ch08-match-values, include=FALSE}
mr <- matched |> group_by(cox2_init) |> summarise(r = mean(pud), .groups = "drop") |> pull(r)
m_rd <- mr[2] - mr[1]
paired <- matched |>
  select(pair_id, cox2_init, pud) |>
  pivot_wider(names_from = cox2_init, values_from = pud, names_prefix = "y")
d <- paired$y1 - paired$y0
m_se <- sd(d) / sqrt(length(d))
```

The matched risk difference is `r sprintf("%+.1f", 100 * m_rd)`
percentage points, `r sprintf("%.1f", 100 * mr[2])`% against
`r sprintf("%.1f", 100 * mr[1])`%, with a risk ratio of
`r sprintf("%.2f", mr[2] / mr[1])`. Because the sample is paired, the
standard error should respect the pairing: we take the within-pair
difference in outcomes for each of the `r n_pairs` pairs and compute the
standard error of their mean, `r sprintf("%.1f", 100 * m_se)` percentage
points, giving an interval of roughly
`r sprintf("%.1f", 100 * (m_rd - 1.96 * m_se))` to
`r sprintf("%+.1f", 100 * (m_rd + 1.96 * m_se))` points. It comfortably
includes zero.

That point estimate is further from the null than the
`r sprintf("%+.1f", 100 * std_rd)` points standardization gave us, and
two things account for the gap. The obvious one is precision: we
discarded `r nrow(ns) - 2 * n_pairs` patients, so the matched interval
is much wider than Chapter 7's and the point estimate is free to wander
within it. The subtler one is that **matching estimates a different
estimand**. Every treated patient is kept and controls are recruited to
resemble them, so the covariate distribution we have averaged over is
that of the Cox-2 initiators — the ATT of Chapter 3, not the ATE. Given
that the Cox-2 arm is older and sicker, and therefore at higher baseline
risk, there is no reason to expect the ATT and the ATE to agree, and the
matched estimate should not be read as a failed replication of the
standardized one. It is an answer to a neighboring question, and
arguably the more clinically natural one: what did this drug do for the
patients who actually received it?

## Choosing the PS model

Everything above depends on which variables enter `ps_model`, and the
selection rule is not "whatever predicts treatment best." Judging a
propensity-score model by its c-statistic is actively misleading: a
model that separated the arms perfectly would push the two densities
apart until nothing overlapped, destroying the very comparability we
need. The model is good if it produces balance and preserves overlap,
which is what the love plot and the density plot were for.

Chapter 5's diagram is the guide to what belongs. Confounders — common
causes of treatment and outcome, such as age, arthritis, and steroid use
— must be included; that is the whole point. Pure outcome predictors,
variables that affect the outcome but not treatment, should generally be
included too: they cannot introduce bias, and conditioning on them
tightens the estimate. The variables to keep out are **instruments**:
things that influence treatment but have no effect on the outcome except
through treatment. Physician prescribing preference is the classic
example. Including an instrument sharpens the treatment model, which is
exactly why it hurts — it spreads the propensity scores apart, reduces
overlap, inflates the variance of the estimate, and, worse, amplifies
whatever bias unmeasured confounding was already contributing. And as in
Chapter 5, colliders and anything on the causal pathway from treatment
to outcome stay out; the propensity score offers no exemption from the
rules for reading a DAG.

That last consequence deserves its name, because it is the one that
surprises people. **Bias amplification** is not merely a loss of
precision: conditioning on an instrument-like variable can make the bias
from an unmeasured confounder *larger* than it was before adjustment.
Intuitively, the instrument explains away the part of treatment variation
that was innocent, leaving the residual variation more heavily loaded with
the unmeasured confounder, and the estimate is then contrasted across
patients who differ more on $U$ than the full cohort did. Brookhart et al.
(2006, AJE) demonstrated the effect in simulation, and it is why "include
everything that predicts treatment" is bad advice rather than merely
wasteful. The opposite extreme is also defensible in the right setting: in
large claims databases, high-dimensional proxy adjustment — the **hd-PS**
algorithm of Schneeweiss et al. (2009) — screens thousands of diagnosis
and dispensing codes for empirical proxies of unmeasured confounders and
selects a few hundred, on the argument that no investigator can enumerate
them by hand. That approach is beyond this book's scope, but it is where
the course's variable-selection session lands, and the tension between it
and the diagram-first rule above is a real one rather than a
misunderstanding.

Two habits follow. Choose the covariate set from substantive knowledge
and the diagram before looking at the outcome — the design stage exists
precisely so this is possible. And when a variable's role is genuinely
unclear, prefer including it if it plausibly affects the outcome and
excluding it if it plausibly affects only treatment, then report the
analysis both ways. `ns_covs` was assembled on that reasoning, and its
arguable members — `ppi_use` and `h2_antagonist_use` from Chapter 7's
exercises among them — are the ones worth varying.

## Exercises

1. Re-estimate the propensity score with `arthritis` removed from the
   covariate list, rematch, and compare the two love plots. How much
   arthritis imbalance survives in the matched sample, and do the other
   covariates stay balanced? Explain why leaving out a strong confounder
   can still yield a matched sample that looks acceptable on the
   remaining variables.
2. Rerun `ps_match()` with `caliper = 0.01` and report the number of
   pairs. How many treated patients are lost relative to the 0.05
   caliper? Compute the mean and maximum within-pair difference in
   propensity score under the looser caliper, and use those numbers to
   explain why tightening it costs so little here — then describe the
   cohort in which it would cost a great deal.
3. Build one table comparing three estimates of the effect of Cox-2
   initiation on peptic ulcer disease: the precision-weighted stratified
   risk difference, the matched risk difference, and the standardized
   risk difference from `standardize()` in `R/standardize.R` (use
   `n_boot = 200` and a fixed seed). Report each with an interval. State
   which estimand each one targets and rank them by precision, then say
   which you would report as the primary analysis and why.
