# Inverse probability of treatment weighting

```{r setup-ch09, include=FALSE}
source("R/setup.R")
```

Chapter 8 built a propensity score and used it in two ways, each of
which discarded information: stratification coarsened the score into
five bands, and matching dropped every nonselective initiator who failed
to pair with a Cox-2 patient. This chapter puts the same fitted score to
a third use that retains every patient. Although the earlier estimators
worked by grouping people who resemble each other, here we leave the
cohort intact and assign each patient a *weight*, larger for
those who received the treatment their covariates made unlikely and
smaller for those who received the expected one. In the reweighted cohort
treatment and covariates are unrelated by construction, so a comparison
of two weighted means estimates the causal contrast. This is **inverse
probability of treatment weighting**, the last of Part II's three
estimators and the one that generalizes most readily, since it extends
cleanly to treatments that change over time.

## Reweighting to a pseudo-population

The intuition is easiest to see in a cohort small enough to work by
hand. Suppose the only confounder is frailty, present in 400 of 1,000
patients. Frail patients are treated with probability 0.60 and
non-frail patients with probability 0.15, and the true risk of the
outcome is 20% in frail patients and 10% in non-frail patients
regardless of treatment — so the treatment does nothing at all.

```{r ch09-toy}
toy <- tibble(
  frail = c(  1,   1,   0,   0),
  arm   = c(  1,   0,   1,   0),
  n     = c(240, 160,  90, 510),
  cases = c( 48,  32,   9,  51)
) |>
  mutate(
    e        = if_else(frail == 1, 0.60, 0.15),
    wt       = if_else(arm == 1, 1 / e, 1 / (1 - e)),
    pseudo_n = n * wt,
    pseudo_y = cases * wt
  )

toy |> knitr::kable(digits = 3)
```

```{r ch09-toy-values, include=FALSE}
crude_toy <- with(toy, tapply(cases, arm, sum) / tapply(n, arm, sum))
wtd_toy   <- with(toy, tapply(pseudo_y, arm, sum) / tapply(pseudo_n, arm, sum))
```

Compared crudely, treatment looks harmful: risk
`r sprintf("%.1f", 100 * crude_toy[2])`% among the treated against
`r sprintf("%.1f", 100 * crude_toy[1])`% among the untreated, a risk
difference of
`r sprintf("%+.1f", 100 * (crude_toy[2] - crude_toy[1]))` percentage
points for a drug with no effect. The reason is entirely in the `n`
column: the treated arm is nearly three-quarters frail, the untreated
arm three-quarters not.

Now read the `wt` column. A frail treated patient had probability 0.60
of ending up where she did, so she stands in for $1/0.60 = 1.67$
patients like her; a frail *untreated* patient had probability 0.40, so
she stands in for two and a half. Look at `pseudo_n`: both arms now
contain 400 frail and 600 non-frail patients, 1,000 apiece — each arm
inflated into a copy of the whole cohort. In this
**pseudo-population** — Hernán and Robins's term for the reweighted
cohort — frailty is distributed identically across arms; the association
between frailty and treatment has been broken by duplicating patients
rather than by conditioning on frailty, and the weighted risks come out at
`r sprintf("%.1f", 100 * wtd_toy[2])`% and
`r sprintf("%.1f", 100 * wtd_toy[1])`%, a risk difference of exactly
zero.

That is the essential idea, and the general form of the weight follows
directly. For a patient with covariates $L$ who received treatment $A$
the weight is

$$W = \frac{1}{P(A \mid L)} = \frac{A}{e(L)} + \frac{1-A}{1 - e(L)},$$

the reciprocal of the probability of the treatment actually received,
which is where the estimator's name comes from. Each patient represents
herself plus everyone with her covariates who *would* have gone the
other way, and Chapter 3's identification conditions license that
substitution: consistency, conditional exchangeability given $L$, and
positivity. Positivity is doing conspicuous work here, because as $e(L)$
approaches zero for a treated patient her weight grows without bound, and
a handful of patients can come to speak for the entire pseudo-population.

## Building the weights

The weight above, $1/P(A \mid L)$, is the **unstabilized** weight. It
averages to about 2, because each arm is inflated to the size of the
full cohort and the pseudo-population is therefore roughly twice the
real one. Nothing is wrong with that, although it makes the weights
larger and more variable than they need to be, and variability in the
weights inflates the variance of the estimate.

**Stabilized** weights fix this by multiplying by the marginal
probability of the treatment received,

$$W^{s} = \frac{P(A)}{P(A \mid L)} = A\,\frac{\Pr(A=1)}{e(L)} + (1-A)\,\frac{1 - \Pr(A=1)}{1 - e(L)}.$$

The numerator does not depend on covariates, so it cannot undo the
balancing the denominator achieves; it does no more than rescale each arm so that
the pseudo-population is about the size of the real cohort. The
diagnostic consequence is convenient: stabilized weights should average
close to 1, and a mean far from 1 is a sign that the treatment model is
misbehaving. Cole and Hernán (2008) are the practical reference for
constructing these weights and for reading their mean this way.

Our estimator packages the whole procedure — fit the treatment model,
form the weights, optionally truncate them, and fit a weighted
regression of the outcome on treatment alone.

```{r ch09-iptw-def, eval=FALSE}
iptw_estimator <- function(data, treat_model, outcome, treat,
                           family = gaussian(), stabilize = TRUE,
                           truncate_at = NULL) {
  data <- est_ps(data, treat_model)
  p_treat <- mean(data[[treat]])

  data <- data |>
    mutate(
      wt = if_else(.data[[treat]] == 1, 1 / ps, 1 / (1 - ps)),
      wt = if (stabilize) {
        if_else(.data[[treat]] == 1, p_treat * wt, (1 - p_treat) * wt)
      } else wt
    )

  if (!is.null(truncate_at)) {
    data <- data |> mutate(wt = pmin(wt, quantile(wt, truncate_at)))
  }

  fit <- geepack::geeglm(
    reformulate(treat, outcome),
    id = seq_len(nrow(data)), weights = data$wt,
    family = family, data = data
  )
  broom::tidy(fit, conf.int = TRUE)
}
```

Three details deserve comment. The function calls Chapter 8's `est_ps()`
rather than fitting its own logistic model, which keeps one definition of
"propensity score" in the book. The outcome regression is
`reformulate(treat, outcome)` — treatment and nothing else, because the
weights have already dealt with the covariates; a weighted regression
that also adjusted for `ns_covs` would be a different, doubly adjusted
estimator. And the fit goes through `geepack::geeglm()` with
`id = seq_len(nrow(data))`, making every patient her own cluster under an
independent working correlation. That is an elaborate way to reproduce a
`glm()` point estimate, but it delivers the sandwich standard error,
which is what we need here. A weighted `glm()` treats the weights as
counts of real observations and reports intervals that are far too
narrow, whereas the robust variance from `geeglm()` is estimated
empirically rather than from the weighted likelihood.

That code lives in `R/iptw_estimator.R`. Before using it, we look at the
weights themselves.

```{r ch09-weights}
source("R/propensity.R")

ps_model <- reformulate(ns_covs, "cox2_init")
wts <- est_ps(ns, ps_model) |>
  mutate(
    p_treat  = mean(cox2_init),
    w_unstab = if_else(cox2_init == 1, 1 / ps, 1 / (1 - ps)),
    w_stab   = if_else(cox2_init == 1, p_treat * w_unstab,
                       (1 - p_treat) * w_unstab)
  )

summary(wts$w_unstab)
summary(wts$w_stab)
```

```{r ch09-wt-hist}
ggplot(wts, aes(x = w_stab)) +
  geom_histogram(bins = 40, fill = "grey35") +
  facet_wrap(
    ~ factor(cox2_init, labels = c("Nonselective NSAID", "Cox-2 selective")),
    scales = "free_y"
  ) +
  labs(x = "Stabilized weight", y = "Patients") +
  theme_minimal()
```

```{r ch09-wt-values, include=FALSE}
mean_stab <- mean(wts$w_stab)
max_un    <- max(wts$w_unstab)
max_st    <- max(wts$w_stab)
ess <- function(w) sum(w)^2 / sum(w^2)
ess_un <- ess(wts$w_unstab); ess_st <- ess(wts$w_stab)
n_big <- sum(wts$w_stab > 1.5)
big_trt <- sum(wts$w_stab > 1.5 & wts$cox2_init == 1)
ps_min_trt <- min(wts$ps[wts$cox2_init == 1])
```

The stabilized weights average `r sprintf("%.3f", mean_stab)`, close
enough to 1 to be reassuring, while the unstabilized ones average about
2 as advertised. The maxima are more informative:
`r sprintf("%.1f", max_un)` unstabilized against
`r sprintf("%.1f", max_st)` stabilized. That largest unstabilized weight
belongs to the Cox-2 initiator whose estimated propensity score is
`r sprintf("%.3f", ps_min_trt)` — a patient whose covariates made the
Cox-2 drug very unlikely, who received it anyway, and who is therefore
asked to represent thirty-odd patients like herself who did not. Kish's
effective sample size, $(\sum w)^2 / \sum w^2$, falls from
`r nrow(ns)` unweighted to `r sprintf("%.0f", ess_st)` under stabilized
weights and all the way to `r sprintf("%.0f", ess_un)` unstabilized: a
compact measure of how much information a few large weights consume.

The histogram shows where the asymmetry lies. Nonselective initiators
are the common case, so their weights cluster tightly around 1, while
the Cox-2 arm spreads out with a long right tail:
`r n_big` patients carry a stabilized weight above 1.5, of whom
`r big_trt` are Cox-2 initiators. Extreme weights are best read as a
symptom of a positivity problem rather than as a numerical nuisance to be
tidied away. The patient with the largest weight sits in the sparse left
tail of Chapter 8's density plot, where matching declined to find a
partner. Matching reported that difficulty by leaving her out, whereas
weighting reports it by making her highly influential, which is easier to
miss, so the weight distribution should be printed and plotted routinely.

## Estimating the effect

With the weights in hand the effect estimate is a weighted regression of
`pud` on `cox2_init` and nothing else. The link function chooses the
scale: an identity link on a 0/1 outcome — `family = gaussian()` — gives
the treatment coefficient as a risk difference, and a log link with
`poisson()` gives it as a log risk ratio, which we exponentiate.

```{r iptw}
source("R/propensity.R"); source("R/iptw_estimator.R")
ps_model <- reformulate(ns_covs, "cox2_init")
iptw_estimator(ns, ps_model, outcome = "pud", treat = "cox2_init",
               family = gaussian()) |> knitr::kable(digits = 4)
iptw_estimator(ns, ps_model, outcome = "pud", treat = "cox2_init",
               family = poisson(link = "log")) |>
  mutate(across(c(estimate, conf.low, conf.high), exp)) |> knitr::kable(digits = 3)
```

```{r ch09-iptw-values, include=FALSE}
g <- function(fam, col) {
  r <- iptw_estimator(ns, ps_model, outcome = "pud", treat = "cox2_init",
                      family = fam)
  r[[col]][r$term == "cox2_init"]
}
ipw_rd <- g(gaussian(), "estimate")
ipw_lo <- g(gaussian(), "conf.low"); ipw_hi <- g(gaussian(), "conf.high")
ipw_rr    <- exp(g(poisson(link = "log"), "estimate"))
ipw_rr_lo <- exp(g(poisson(link = "log"), "conf.low"))
ipw_rr_hi <- exp(g(poisson(link = "log"), "conf.high"))
```

The weighted risk difference is
`r sprintf("%+.2f", 100 * ipw_rd)` percentage points, with a 95%
interval from `r sprintf("%.1f", 100 * ipw_lo)` to
`r sprintf("%+.1f", 100 * ipw_hi)` points, and the weighted risk ratio
is `r sprintf("%.2f", ipw_rr)`
(`r sprintf("%.2f", ipw_rr_lo)` to `r sprintf("%.2f", ipw_rr_hi)`).
Both estimates sit essentially on the null. The Poisson fit deserves a
note. The outcome is binary rather than a count, so the likelihood is
deliberately wrong, although the coefficient of a log-link regression on
a 0/1 outcome still estimates a log risk ratio, and the sandwich variance
repairs the standard error that the wrong likelihood would give. This is
the modified Poisson approach to risk ratios, and it is a further reason
for the `geeglm()` machinery.

There is very little on the outcome side of this analysis: the regression
has one covariate. Everything we know about age, arthritis, steroids, and
the other seventeen variables entered through the weights, which were
built without ever looking at `pud`, so Chapter 8's design-stage
discipline carries through to the estimate.

## Truncation and diagnostics

The obvious response to a weight of `r sprintf("%.1f", max_un)` is to
cap it. **Truncation** replaces every weight above a chosen percentile
with the value at that percentile; our function does this with
`pmin(wt, quantile(wt, truncate_at))` when `truncate_at` is supplied.

```{r ch09-truncate}
iptw_estimator(ns, ps_model, outcome = "pud", treat = "cox2_init",
               family = gaussian(), truncate_at = 0.99) |>
  knitr::kable(digits = 4)
```

```{r ch09-trunc-values, include=FALSE}
tr <- iptw_estimator(ns, ps_model, outcome = "pud", treat = "cox2_init",
                     family = gaussian(), truncate_at = 0.99)
tr_rd <- tr$estimate[tr$term == "cox2_init"]
tr_se <- tr$std.error[tr$term == "cox2_init"]
un_se <- g(gaussian(), "std.error")
n_cap <- sum(wts$w_stab > quantile(wts$w_stab, 0.99))
```

Capping at the 99th percentile touches `r n_cap` of the
`r nrow(ns)` patients and moves the risk difference from
`r sprintf("%+.2f", 100 * ipw_rd)` to
`r sprintf("%+.2f", 100 * tr_rd)` percentage points, with the standard
error changing from `r sprintf("%.4f", un_se)` to
`r sprintf("%.4f", tr_se)`. Truncation therefore buys almost nothing in
this cohort, which is itself informative: the estimate is robust to the
handful of most influential patients.

The trade-off deserves to be stated explicitly, since truncation is a
deliberate exchange of bias for variance. If the treatment model is
correct, the untruncated estimator is consistent for the ATE and the
truncated one is not, because we have knowingly misrepresented how much
of the pseudo-population the extreme patients should represent. What we
get in return is stability, and when a single weight dominates the sum a
small amount of bias can be worth accepting. Truncation is therefore
better used as a sensitivity analysis than as a default: report the
untruncated estimate, report it capped at the 99th and 95th percentiles,
and if they disagree materially, say so. An estimate that depends on
whether one patient counts thirty times or ten is being driven by
extrapolation into a region where the data are thin, and no choice of cap
will remedy that.
Trimming, which drops patients outside a range of the propensity score
altogether, is the other common option; it changes the population and
therefore the estimand, and should be described as such. The distinction
is worth holding onto, since the two words are not always kept apart:
truncation caps a weight and keeps the patient, trimming discards the
patient — often those outside a percentile range of the propensity score
— and both the course and much of the literature use "trimming" loosely
for either operation, so it is worth saying which one you did.

## Standardized mortality ratio weights

The weights built so far blow both arms up into a copy of the whole
cohort, which is why they estimate the ATE. A small change to the weight
redirects the estimator at a different population. Leave the treated arm
alone — every treated patient keeps weight 1 — and give each untreated
patient $e(L) / (1 - e(L))$, the odds of treatment her covariates imply.
The untreated are then reweighted to resemble the treated rather than the
cohort, the treated covariate distribution becomes the standard, and the
estimand is the ATT: the effect of treatment among the treated. In the
course these are called **SMR (standardized mortality ratio) weights**,
after the indirect standardization that gives a standardized mortality
ratio its name.

```{r ch09-smr}
source("R/propensity.R")

smr <- est_ps(ns, ps_model) |>
  mutate(w_smr = if_else(cox2_init == 1, 1, ps / (1 - ps)))

smr_fit <- geepack::geeglm(
  pud ~ cox2_init,
  id = seq_len(nrow(smr)), weights = smr$w_smr,
  family = gaussian(), data = smr
)
broom::tidy(smr_fit, conf.int = TRUE) |> knitr::kable(digits = 4)
```

```{r ch09-smr-values, include=FALSE}
st <- broom::tidy(smr_fit, conf.int = TRUE)
gs <- function(col) st[[col]][st$term == "cox2_init"]
smr_rd <- gs("estimate"); smr_lo <- gs("conf.low"); smr_hi <- gs("conf.high")
w_unt     <- smr$w_smr[smr$cox2_init == 0]
smr_mean  <- mean(w_unt); smr_max <- max(w_unt)
smr_match <- ps_match(est_ps(ns, ps_model), treat = "cox2_init",
                      caliper = 0.05)
smr_mr    <- tapply(smr_match$pud, smr_match$cox2_init, mean)
smr_att   <- smr_mr[2] - smr_mr[1]
```

The SMR-weighted risk difference is
`r sprintf("%+.2f", 100 * smr_rd)` percentage points
(`r sprintf("%.1f", 100 * smr_lo)` to
`r sprintf("%+.1f", 100 * smr_hi)`). Set beside the ATE of
`r sprintf("%+.2f", 100 * ipw_rd)` points from the same score and
Chapter 8's matched estimate of `r sprintf("%+.2f", 100 * smr_att)`
points on the same cohort, it lands between them. That is unsurprising.
SMR weighting targets the ATT, as matching did, although it reaches the
ATT by retaining the unmatched controls at reduced weight rather than
discarding them, so the two ATT estimates need not coincide, and neither
sits far from the null. The weights themselves indicate what this costs:
among the untreated they average only `r sprintf("%.2f", smr_mean)` and
none exceeds `r sprintf("%.2f", smr_max)`, because the common arm has to
be shrunk substantially to resemble the rare one. The confidence interval
widens accordingly, which is the price of asking about the treated rather
than about everyone.

## MSMs and where weighting shines

The one-covariate regression we fit in the pseudo-population has a name.
A **marginal structural model** is a model for the marginal mean of a
potential outcome as a function of treatment,

$$E[Y(a)] = \beta_0 + \beta_1 a,$$

*structural* because its arguments are counterfactual treatments rather
than observed ones, *marginal* because it averages over the covariate
distribution instead of conditioning on it. The class, and the name, are
Robins, Hernán, and Brumback's (2000). Under the identification
conditions, fitting it in the weighted data estimates its parameters, so
$\beta_1$ is the ATE of Chapter 3 — the same estimand standardization
targeted, reached by another route. Ours is *saturated*: with a binary
treatment it has one parameter per arm and so cannot be misspecified.
Larger MSMs can be, and often carry baseline effect modifiers on the
right-hand side to allow subgroup effects.

The more important reason to learn weighting is that the framework
extends to time-varying treatment, whereas conditioning does not.
Suppose patients start and stop a drug over months of follow-up while
their covariates evolve. A confounder measured at month six may itself be
affected by treatment at month three, making it a mediator as well as a
confounder. If we adjust for it in an outcome regression we block part of
the effect we are trying to measure and may induce collider bias as well,
whereas if we leave it out the later treatment remains confounded. No
choice of covariate set resolves that conflict by conditioning, but
weighting can: build a weight from a treatment model at each time point,
multiply the per-visit weights along each patient's history, and fit a
marginal structural model in the resulting pseudo-population. The
time-varying covariates then appear only in the weight models, where they
cause no trouble, and never in the outcome model, so nothing is blocked.
Chapter 11 applies the same logic to censoring rather than treatment, and
Chapter 12 uses it to emulate trials with sustained strategies; much of
Part III consists of variations on this chapter.

## Three estimators, one estimand

Part II asked one question three times. Here are the answers, with the
crude comparison for reference. The standardization row re-runs
`standardize()` with Chapter 7's seed, so its bootstrap interval
reproduces that chapter's exactly.

```{r ch09-compare}
source("R/standardize.R")

crude <- ns |>
  group_by(cox2_init) |>
  summarise(r = mean(pud), .groups = "drop") |>
  pull(r)
std <- standardize(ns, reformulate(c("cox2_init", ns_covs), "pud"),
                   treat = "cox2_init", n_boot = 200, seed = 921)
ipw <- iptw_estimator(ns, ps_model, outcome = "pud", treat = "cox2_init")

pick <- function(d, what, col) d[[col]][d[[1]] == what]

comparison <- tibble(
  estimator = c("Crude", "Standardization (Ch. 7)", "IPTW, stabilized"),
  estimand  = c("none (confounded)", "ATE", "ATE"),
  rd_pp     = 100 * c(crude[2] - crude[1], pick(std, "rd", "estimate"),
                      pick(ipw, "cox2_init", "estimate")),
  lower     = 100 * c(NA, pick(std, "rd", "conf_low"),
                      pick(ipw, "cox2_init", "conf.low")),
  upper     = 100 * c(NA, pick(std, "rd", "conf_high"),
                      pick(ipw, "cox2_init", "conf.high"))
)

comparison |> knitr::kable(digits = 2)
```

```{r ch09-compare-values, include=FALSE}
crude_rd <- crude[2] - crude[1]
std_rd   <- pick(std, "rd", "estimate")
matched <- ps_match(est_ps(ns, ps_model), treat = "cox2_init", caliper = 0.05)
mr <- tapply(matched$pud, matched$cox2_init, mean)
att_rd <- mr[2] - mr[1]
```

The crude risk difference of
`r sprintf("%+.1f", 100 * crude_rd)` percentage points is the confounded
starting point, and both adjusted estimators pull it across the null:
standardization to `r sprintf("%+.2f", 100 * std_rd)` points and IPTW to
`r sprintf("%+.2f", 100 * ipw_rd)`. The agreement is expected rather than
coincidental. Both target the ATE, both use the same nineteen covariates,
and both rest on the same identification assumptions; they differ only in
which nuisance model carries the load —
an outcome model for `pud` in Chapter 7, a treatment model for
`cox2_init` here. That two estimators relying on different nuisance
models land in the same place is mild evidence that neither model is
badly wrong, although it is not evidence that the assumptions hold, since
both would be wrong together if `ns_covs` were missing an important
confounder.

Chapter 8's matched analysis, recomputed here, gave a risk difference of
`r sprintf("%+.1f", 100 * att_rd)` percentage points, further from the
null than either of these, which is not a contradiction, since
matching targeted the ATT, as the SMR weights above did, averaging over
the older and sicker Cox-2 initiators rather than over the whole cohort,
and did so using roughly a sixth of the data. Its confidence interval
overlaps these comfortably.

All three estimators agree on the shape of the answer. The crude signal
of harm is confounding, and once the arms are made comparable the
estimates hover near the null, with intervals wide enough to admit either
a modest benefit or a modest harm. With `r sum(ns$pud)` ulcer cases and
`r sum(ns$cox2_init)` Cox-2 initiators, that is as far as these data can
take us, and no alternative estimator would sharpen the conclusion, since
a better estimator cannot supply information the cohort does not contain.

## Exercises

1. Rerun `iptw_estimator()` with `stabilize = FALSE` and compare the
   point estimate, the standard error, and the weight distribution to the
   stabilized version. The estimate and its robust standard error will be
   unchanged; explain why stabilization cannot matter for a *saturated*
   marginal structural model, and describe a setting — an MSM with
   effect modifiers, or the time-varying weights of Chapter 11 — where
   it would.
2. Drop five covariates from `ns_covs` to misspecify the treatment model
   — take `arthritis`, `corticosteroid_use`, `aspirin_use`, `ppi_use`,
   and `h2_antagonist_use`, all plausible confounders — and rerun the
   estimator. Report the maximum stabilized weight and the risk
   difference under both covariate sets. Which moves more, the estimate
   or the weight distribution, and what does that tell you about using
   weight diagnostics to detect an inadequate adjustment set?
3. Formalize what the SMR section did by hand: write `iptw_att()`, a
   copy of `iptw_estimator()` in which treated patients receive weight 1
   and untreated patients receive `ps / (1 - ps)`. Confirm that it
   reproduces the SMR-weighted estimate above, then compare it with
   Chapter 8's matched estimate. Why is stabilization unnecessary for
   these weights, and which arm's weights now carry the long tail?
4. In `ns` the ATE and ATT estimates above sit close together, which is a
   fact about this cohort rather than a general result. Kurth and
   colleagues (2006, AJE) analyzed thrombolysis with t-PA and mortality
   after ischemic stroke, where the treatment effect varied sharply with
   baseline stroke severity, and found that IPTW and SMR weighting of the
   same cohort gave sharply different answers, far apart enough that a
   reader would draw different conclusions from the two. Using this
   chapter's vocabulary, explain which patients
   dominate the weighted comparison under each scheme, and why effect
   heterogeneity turns the choice between them from a technical preference
   into a substantive one. Then say which of the two you would specify in
   advance for a question about whether to start treatment in a patient
   who currently would not receive it, and which for a question about
   whether current prescribing practice is doing harm.
