# Inverse probability of treatment weighting

```{r setup-ch09, include=FALSE}
source("R/setup.R")
```

Chapter 8 built a propensity score and then spent it two ways, both of
which threw something away: stratification coarsened the score into five
bands, and matching discarded every nonselective initiator who failed to
pair with a Cox-2 patient. This chapter puts the same fitted score to a
third use that keeps every patient. Rather than grouping people who
resemble each other, we leave the cohort intact and give each patient a
*weight* — larger for those who received the treatment their covariates
made unlikely, smaller for those who received the expected one. In the
reweighted cohort treatment and covariates are unrelated by
construction, so a plain comparison of two weighted means estimates the
causal contrast. This is **inverse probability of treatment weighting**,
the last of Part II's three estimators and the one whose logic carries
furthest, since it alone extends cleanly to treatments that change over
time.

## Reweighting to a pseudo-population

The intuition is easiest to see in a cohort small enough to work by
hand. Suppose the only confounder is frailty, present in 400 of 1,000
patients. Frail patients are treated with probability 0.60 and
non-frail patients with probability 0.15, and the true risk of the
outcome is 20% in frail patients and 10% in non-frail patients
regardless of treatment — so the treatment does nothing at all.

```{r ch09-toy}
toy <- tibble(
  frail = c(  1,   1,   0,   0),
  arm   = c(  1,   0,   1,   0),
  n     = c(240, 160,  90, 510),
  cases = c( 48,  32,   9,  51)
) |>
  mutate(
    e        = if_else(frail == 1, 0.60, 0.15),
    wt       = if_else(arm == 1, 1 / e, 1 / (1 - e)),
    pseudo_n = n * wt,
    pseudo_y = cases * wt
  )

toy |> knitr::kable(digits = 3)
```

```{r ch09-toy-values, include=FALSE}
crude_toy <- with(toy, tapply(cases, arm, sum) / tapply(n, arm, sum))
wtd_toy   <- with(toy, tapply(pseudo_y, arm, sum) / tapply(pseudo_n, arm, sum))
```

Compared crudely, treatment looks harmful: risk
`r sprintf("%.1f", 100 * crude_toy[2])`% among the treated against
`r sprintf("%.1f", 100 * crude_toy[1])`% among the untreated, a risk
difference of
`r sprintf("%+.1f", 100 * (crude_toy[2] - crude_toy[1]))` percentage
points for a drug with no effect. The reason is entirely in the `n`
column: the treated arm is nearly three-quarters frail, the untreated
arm three-quarters not.

Now read the `wt` column. A frail treated patient had probability 0.60
of ending up where she did, so she stands in for $1/0.60 = 1.67$
patients like her; a frail *untreated* patient had probability 0.40, so
she stands in for two and a half. Look at `pseudo_n`: both arms now
contain 400 frail and 600 non-frail patients, 1,000 apiece — each arm
inflated into a copy of the whole cohort. In this
**pseudo-population** — Hernán and Robins's term for the reweighted
cohort — frailty is distributed identically across arms, its association
with treatment broken not by conditioning on frailty but by duplicating
people until it no longer matters, and the weighted risks come out at
`r sprintf("%.1f", 100 * wtd_toy[2])`% and
`r sprintf("%.1f", 100 * wtd_toy[1])`%, a risk difference of exactly
zero.

That is the whole idea. The general form of the weight for a patient
with covariates $L$ who received treatment $A$ is

$$W = \frac{1}{P(A \mid L)} = \frac{A}{e(L)} + \frac{1-A}{1 - e(L)},$$

the reciprocal of the probability of the treatment actually received —
which is where the estimator's name comes from. Each patient represents
herself plus everyone with her covariates who *would* have gone the
other way, and Chapter 3's identification conditions license that
substitution: consistency, conditional exchangeability given $L$, and
positivity. Positivity now does visible work. As $e(L)$ approaches zero
for a treated patient her weight grows without bound, and a handful of
patients come to speak for the entire pseudo-population.

## Building the weights

The weight above, $1/P(A \mid L)$, is the **unstabilized** weight. It
averages to about 2, because each arm is blown up to the size of the
full cohort and the pseudo-population is therefore roughly twice the
real one. Nothing is wrong with that, but it makes the weights larger
and more variable than they need to be, and variance in the weights is
variance in the estimate.

**Stabilized** weights fix this by multiplying by the marginal
probability of the treatment received,

$$W^{s} = \frac{P(A)}{P(A \mid L)} = A\,\frac{\Pr(A=1)}{e(L)} + (1-A)\,\frac{1 - \Pr(A=1)}{1 - e(L)}.$$

The numerator does not depend on covariates, so it cannot undo the
balancing the denominator achieves; it simply rescales each arm so that
the pseudo-population is about the size of the real cohort. The
diagnostic consequence is convenient: stabilized weights should average
close to 1, and a mean far from 1 is a sign that the treatment model is
misbehaving. Cole and Hernán (2008) are the practical reference for
constructing these weights and for reading their mean this way.

Our estimator packages the whole procedure — fit the treatment model,
form the weights, optionally truncate them, and fit a weighted
regression of the outcome on treatment alone.

```{r ch09-iptw-def, eval=FALSE}
iptw_estimator <- function(data, treat_model, outcome, treat,
                           family = gaussian(), stabilize = TRUE,
                           truncate_at = NULL) {
  data <- est_ps(data, treat_model)
  p_treat <- mean(data[[treat]])

  data <- data |>
    mutate(
      wt = if_else(.data[[treat]] == 1, 1 / ps, 1 / (1 - ps)),
      wt = if (stabilize) {
        if_else(.data[[treat]] == 1, p_treat * wt, (1 - p_treat) * wt)
      } else wt
    )

  if (!is.null(truncate_at)) {
    data <- data |> mutate(wt = pmin(wt, quantile(wt, truncate_at)))
  }

  fit <- geepack::geeglm(
    reformulate(treat, outcome),
    id = seq_len(nrow(data)), weights = data$wt,
    family = family, data = data
  )
  broom::tidy(fit, conf.int = TRUE)
}
```

Three details deserve comment. The function calls Chapter 8's `est_ps()`
rather than fitting its own logistic model, which keeps one definition of
"propensity score" in the book. The outcome regression is
`reformulate(treat, outcome)` — treatment and nothing else, because the
weights have already dealt with the covariates; a weighted regression
that also adjusted for `ns_covs` would be a different, doubly adjusted
estimator. And the fit goes through `geepack::geeglm()` with
`id = seq_len(nrow(data))`, making every patient her own cluster under an
independent working correlation. That is an elaborate way to reproduce a
`glm()` point estimate; what it buys is the sandwich standard error. A
weighted `glm()` treats the weights as counts of real observations and
reports intervals far too narrow; the robust variance from `geeglm()`
does not.

That code lives in `R/iptw_estimator.R`. Before using it, look at the
weights themselves.

```{r ch09-weights}
source("R/propensity.R")

ps_model <- reformulate(ns_covs, "cox2_init")
wts <- est_ps(ns, ps_model) |>
  mutate(
    p_treat  = mean(cox2_init),
    w_unstab = if_else(cox2_init == 1, 1 / ps, 1 / (1 - ps)),
    w_stab   = if_else(cox2_init == 1, p_treat * w_unstab,
                       (1 - p_treat) * w_unstab)
  )

summary(wts$w_unstab)
summary(wts$w_stab)
```

```{r ch09-wt-hist}
ggplot(wts, aes(x = w_stab)) +
  geom_histogram(bins = 40, fill = "grey35") +
  facet_wrap(
    ~ factor(cox2_init, labels = c("Nonselective NSAID", "Cox-2 selective")),
    scales = "free_y"
  ) +
  labs(x = "Stabilized weight", y = "Patients") +
  theme_minimal()
```

```{r ch09-wt-values, include=FALSE}
mean_stab <- mean(wts$w_stab)
max_un    <- max(wts$w_unstab)
max_st    <- max(wts$w_stab)
ess <- function(w) sum(w)^2 / sum(w^2)
ess_un <- ess(wts$w_unstab); ess_st <- ess(wts$w_stab)
n_big <- sum(wts$w_stab > 1.5)
big_trt <- sum(wts$w_stab > 1.5 & wts$cox2_init == 1)
ps_min_trt <- min(wts$ps[wts$cox2_init == 1])
```

The stabilized weights average `r sprintf("%.3f", mean_stab)`, close
enough to 1 to be reassuring, while the unstabilized ones average about
2 as advertised. The maxima are more informative:
`r sprintf("%.1f", max_un)` unstabilized against
`r sprintf("%.1f", max_st)` stabilized. That largest unstabilized weight
belongs to the Cox-2 initiator whose estimated propensity score is
`r sprintf("%.3f", ps_min_trt)` — a patient whose covariates made the
Cox-2 drug very unlikely, who received it anyway, and who is therefore
asked to represent thirty-odd patients like herself who did not. Kish's
effective sample size, $(\sum w)^2 / \sum w^2$, falls from
`r nrow(ns)` unweighted to `r sprintf("%.0f", ess_st)` under stabilized
weights and all the way to `r sprintf("%.0f", ess_un)` unstabilized: a
compact measure of how much information a few large weights consume.

The histogram shows where the asymmetry lives. Nonselective initiators
are the common case, so their weights cluster tightly around 1, while
the Cox-2 arm spreads out with a long right tail —
`r n_big` patients carry a stabilized weight above 1.5, of whom
`r big_trt` are Cox-2 initiators. Extreme weights are not a numerical
annoyance to be tidied away; they are the positivity assumption
complaining. That patient sits in the sparse left tail of Chapter 8's
density plot, where matching declined to find a partner: matching
reported the difficulty by leaving her out, weighting reports it by
making her enormously influential, which is easier to overlook — so the
weight distribution should be printed and plotted every time.

## Estimating the effect

With the weights in hand the effect estimate is a weighted regression of
`pud` on `cox2_init` and nothing else. The link function chooses the
scale: an identity link on a 0/1 outcome — `family = gaussian()` — gives
the treatment coefficient as a risk difference, and a log link with
`poisson()` gives it as a log risk ratio, which we exponentiate.

```{r iptw}
source("R/propensity.R"); source("R/iptw_estimator.R")
ps_model <- reformulate(ns_covs, "cox2_init")
iptw_estimator(ns, ps_model, outcome = "pud", treat = "cox2_init",
               family = gaussian()) |> knitr::kable(digits = 4)
iptw_estimator(ns, ps_model, outcome = "pud", treat = "cox2_init",
               family = poisson(link = "log")) |>
  mutate(across(c(estimate, conf.low, conf.high), exp)) |> knitr::kable(digits = 3)
```

```{r ch09-iptw-values, include=FALSE}
g <- function(fam, col) {
  r <- iptw_estimator(ns, ps_model, outcome = "pud", treat = "cox2_init",
                      family = fam)
  r[[col]][r$term == "cox2_init"]
}
ipw_rd <- g(gaussian(), "estimate")
ipw_lo <- g(gaussian(), "conf.low"); ipw_hi <- g(gaussian(), "conf.high")
ipw_rr    <- exp(g(poisson(link = "log"), "estimate"))
ipw_rr_lo <- exp(g(poisson(link = "log"), "conf.low"))
ipw_rr_hi <- exp(g(poisson(link = "log"), "conf.high"))
```

The weighted risk difference is
`r sprintf("%+.2f", 100 * ipw_rd)` percentage points, with a 95%
interval from `r sprintf("%.1f", 100 * ipw_lo)` to
`r sprintf("%+.1f", 100 * ipw_hi)` points, and the weighted risk ratio
is `r sprintf("%.2f", ipw_rr)`
(`r sprintf("%.2f", ipw_rr_lo)` to `r sprintf("%.2f", ipw_rr_hi)`).
Both sit essentially on the null. The Poisson fit deserves a note: the
outcome is binary, not a count, so the likelihood is deliberately wrong —
but the coefficient of a log-link regression on a 0/1 outcome still
estimates a log risk ratio, and the sandwich variance repairs the
standard error the wrong likelihood would give. This is the modified
Poisson approach to risk ratios, and another reason the `geeglm()`
machinery is there.

Notice how little there is to the outcome side: the regression has one
covariate. Everything we know about age, arthritis, steroids, and the
other seventeen variables entered through the weights, built without ever
looking at `pud` — Chapter 8's design-stage discipline carried through to
the estimate.

## Truncation and diagnostics

The obvious response to a weight of `r sprintf("%.1f", max_un)` is to
cap it. **Truncation** replaces every weight above a chosen percentile
with the value at that percentile; our function does this with
`pmin(wt, quantile(wt, truncate_at))` when `truncate_at` is supplied.

```{r ch09-truncate}
iptw_estimator(ns, ps_model, outcome = "pud", treat = "cox2_init",
               family = gaussian(), truncate_at = 0.99) |>
  knitr::kable(digits = 4)
```

```{r ch09-trunc-values, include=FALSE}
tr <- iptw_estimator(ns, ps_model, outcome = "pud", treat = "cox2_init",
                     family = gaussian(), truncate_at = 0.99)
tr_rd <- tr$estimate[tr$term == "cox2_init"]
tr_se <- tr$std.error[tr$term == "cox2_init"]
un_se <- g(gaussian(), "std.error")
n_cap <- sum(wts$w_stab > quantile(wts$w_stab, 0.99))
```

Capping at the 99th percentile touches `r n_cap` of the
`r nrow(ns)` patients and moves the risk difference from
`r sprintf("%+.2f", 100 * ipw_rd)` to
`r sprintf("%+.2f", 100 * tr_rd)` percentage points, with the standard
error changing from `r sprintf("%.4f", un_se)` to
`r sprintf("%.4f", tr_se)`. In this cohort truncation buys almost
nothing, which is the most useful thing it could have told us: the
estimate does not hang on a few influential patients.

The trade-off is worth stating plainly. Truncation is a
deliberate exchange of bias for variance: the untruncated estimator is
consistent for the ATE if the treatment model is right, and the
truncated one is not, because we have knowingly misrepresented how much
of the pseudo-population the extreme patients should represent. What we
get back is stability, and when a single weight dominates the sum a
little bias can be a good trade. The honest way to use truncation is as
a sensitivity analysis rather than a default: report the untruncated
estimate, report it capped at the 99th and 95th percentiles, and if they
disagree materially, say so — a result that depends on whether one
patient counts thirty times or ten is a result driven by extrapolation
into a region where the data are thin, and no cap repairs that.
Trimming, which drops patients outside a range of the propensity score
altogether, is the other common option; it changes the population and
therefore the estimand, and should be described as such. The distinction
is worth holding onto, since the two words are not always kept apart:
truncation caps a weight and keeps the patient, trimming discards the
patient — often those outside a percentile range of the propensity score
— and both the course and much of the literature use "trimming" loosely
for either operation, so it is worth saying which one you did.

## Standardized mortality ratio weights

The weights built so far blow both arms up into a copy of the whole
cohort, which is why they estimate the ATE. A small change to the weight
redirects the estimator at a different population. Leave the treated arm
alone — every treated patient keeps weight 1 — and give each untreated
patient $e(L) / (1 - e(L))$, the odds of treatment her covariates imply.
The untreated are then reweighted to resemble the treated rather than the
cohort, the treated covariate distribution becomes the standard, and the
estimand is the ATT: the effect of treatment among the treated. In the
course these are called **SMR (standardized mortality ratio) weights**,
after the indirect standardization that gives a standardized mortality
ratio its name.

```{r ch09-smr}
source("R/propensity.R")

smr <- est_ps(ns, ps_model) |>
  mutate(w_smr = if_else(cox2_init == 1, 1, ps / (1 - ps)))

smr_fit <- geepack::geeglm(
  pud ~ cox2_init,
  id = seq_len(nrow(smr)), weights = smr$w_smr,
  family = gaussian(), data = smr
)
broom::tidy(smr_fit, conf.int = TRUE) |> knitr::kable(digits = 4)
```

```{r ch09-smr-values, include=FALSE}
st <- broom::tidy(smr_fit, conf.int = TRUE)
gs <- function(col) st[[col]][st$term == "cox2_init"]
smr_rd <- gs("estimate"); smr_lo <- gs("conf.low"); smr_hi <- gs("conf.high")
w_unt     <- smr$w_smr[smr$cox2_init == 0]
smr_mean  <- mean(w_unt); smr_max <- max(w_unt)
smr_match <- ps_match(est_ps(ns, ps_model), treat = "cox2_init",
                      caliper = 0.05)
smr_mr    <- tapply(smr_match$pud, smr_match$cox2_init, mean)
smr_att   <- smr_mr[2] - smr_mr[1]
```

The SMR-weighted risk difference is
`r sprintf("%+.2f", 100 * smr_rd)` percentage points
(`r sprintf("%.1f", 100 * smr_lo)` to
`r sprintf("%+.1f", 100 * smr_hi)`). Set beside the ATE of
`r sprintf("%+.2f", 100 * ipw_rd)` points from the same score and
Chapter 8's matched estimate of `r sprintf("%+.2f", 100 * smr_att)`
points on the same cohort, it lands between them. That is a reasonable
place to be: SMR weighting targets the ATT, as matching did, but reaches
it by keeping the unmatched controls at reduced weight rather than
discarding them, so the two ATT estimates need not coincide, and neither
sits far from the null. The weights themselves show what that costs:
among the untreated they average only `r sprintf("%.2f", smr_mean)` and
none exceeds `r sprintf("%.2f", smr_max)`, because the common arm has to
be shrunk hard to resemble the rare one. The interval widens
accordingly — the price of asking about the treated rather than everyone.

## MSMs and where weighting shines

The one-covariate regression we fit in the pseudo-population has a name.
A **marginal structural model** is a model for the marginal mean of a
potential outcome as a function of treatment,

$$E[Y(a)] = \beta_0 + \beta_1 a,$$

*structural* because its arguments are counterfactual treatments rather
than observed ones, *marginal* because it averages over the covariate
distribution instead of conditioning on it. The class, and the name, are
Robins, Hernán, and Brumback's (2000). Under the identification
conditions, fitting it in the weighted data estimates its parameters, so
$\beta_1$ is the ATE of Chapter 3 — the same estimand standardization
targeted, reached by another route. Ours is *saturated*: with a binary
treatment it has one parameter per arm and so cannot be misspecified.
Larger MSMs can be, and often carry baseline effect modifiers on the
right-hand side to allow subgroup effects.

The deeper reason to learn weighting is that this framework survives
time-varying treatment, and conditioning does not. Suppose patients
start and stop a drug over months of follow-up while their covariates
evolve. A confounder measured at month six is affected by treatment at
month three — a mediator as well as a confounder. Adjust for it in an
outcome regression and you block part of the effect you are trying to
measure, and may induce collider bias besides; leave it out and the
later treatment stays confounded. No covariate set resolves that
conflict by conditioning. Weighting does: build a weight from a
treatment model at each time point, multiply the per-visit weights along
each patient's history, and fit a marginal structural model in the
resulting pseudo-population. The time-varying covariates appear in the
weight models, where they cause no trouble, and never in the outcome
model, so nothing is blocked. Chapter 11 applies the identical logic to
censoring rather than treatment, and Chapter 12 uses it to emulate
trials with sustained strategies. Part III is largely a series of
variations on this chapter.

## Three estimators, one estimand

Part II asked one question three times. Here are the answers, with the
crude comparison for reference. The standardization row re-runs
`standardize()` with Chapter 7's seed, so its bootstrap interval
reproduces that chapter's exactly.

```{r ch09-compare}
source("R/standardize.R")

crude <- ns |>
  group_by(cox2_init) |>
  summarise(r = mean(pud), .groups = "drop") |>
  pull(r)
std <- standardize(ns, reformulate(c("cox2_init", ns_covs), "pud"),
                   treat = "cox2_init", n_boot = 200, seed = 921)
ipw <- iptw_estimator(ns, ps_model, outcome = "pud", treat = "cox2_init")

pick <- function(d, what, col) d[[col]][d[[1]] == what]

comparison <- tibble(
  estimator = c("Crude", "Standardization (Ch. 7)", "IPTW, stabilized"),
  estimand  = c("none (confounded)", "ATE", "ATE"),
  rd_pp     = 100 * c(crude[2] - crude[1], pick(std, "rd", "estimate"),
                      pick(ipw, "cox2_init", "estimate")),
  lower     = 100 * c(NA, pick(std, "rd", "conf_low"),
                      pick(ipw, "cox2_init", "conf.low")),
  upper     = 100 * c(NA, pick(std, "rd", "conf_high"),
                      pick(ipw, "cox2_init", "conf.high"))
)

comparison |> knitr::kable(digits = 2)
```

```{r ch09-compare-values, include=FALSE}
crude_rd <- crude[2] - crude[1]
std_rd   <- pick(std, "rd", "estimate")
matched <- ps_match(est_ps(ns, ps_model), treat = "cox2_init", caliper = 0.05)
mr <- tapply(matched$pud, matched$cox2_init, mean)
att_rd <- mr[2] - mr[1]
```

The crude risk difference of
`r sprintf("%+.1f", 100 * crude_rd)` percentage points is the confounded
starting point, and both adjusted estimators pull it across the null:
standardization to `r sprintf("%+.2f", 100 * std_rd)` points and IPTW to
`r sprintf("%+.2f", 100 * ipw_rd)`. The agreement is no accident. Both
target the ATE, both use the same nineteen covariates, and both rest on
the same identification assumptions; they differ only in which nuisance
model carries the load — an outcome model for `pud` in Chapter 7, a
treatment model for `cox2_init` here. Two estimators leaning on
different models landing in the same place is mild evidence that neither
model is badly wrong. It is not evidence that the assumptions hold,
since both would be wrong together if `ns_covs` were missing an
important confounder.

Chapter 8's matched analysis, recomputed here, gave a risk difference of
`r sprintf("%+.1f", 100 * att_rd)` percentage points, further from the
null than either of these — not a contradiction, since matching targeted
the ATT, as the SMR weights above did, averaging over the older and
sicker Cox-2 initiators rather than the whole cohort, and did so with
roughly a sixth of the data. Its interval overlaps these comfortably.

What all three agree on is the shape of the answer: the crude signal of
harm is confounding, and once the arms are comparable the estimates
hover near the null with intervals wide enough to admit a modest benefit
or a modest harm. With `r sum(ns$pud)` ulcer cases and
`r sum(ns$cox2_init)` Cox-2 initiators, that is the honest summary. No
choice of estimator will sharpen it, because better estimators cannot
manufacture information the cohort does not contain.

## Exercises

1. Rerun `iptw_estimator()` with `stabilize = FALSE` and compare the
   point estimate, the standard error, and the weight distribution to the
   stabilized version. The estimate and its robust standard error will be
   unchanged; explain why stabilization cannot matter for a *saturated*
   marginal structural model, and describe a setting — an MSM with
   effect modifiers, or the time-varying weights of Chapter 11 — where
   it would.
2. Drop five covariates from `ns_covs` to misspecify the treatment model
   — take `arthritis`, `corticosteroid_use`, `aspirin_use`, `ppi_use`,
   and `h2_antagonist_use`, all plausible confounders — and rerun the
   estimator. Report the maximum stabilized weight and the risk
   difference under both covariate sets. Which moves more, the estimate
   or the weight distribution, and what does that tell you about using
   weight diagnostics to detect an inadequate adjustment set?
3. Formalize what the SMR section did by hand: write `iptw_att()`, a
   copy of `iptw_estimator()` in which treated patients receive weight 1
   and untreated patients receive `ps / (1 - ps)`. Confirm that it
   reproduces the SMR-weighted estimate above, then compare it with
   Chapter 8's matched estimate. Why is stabilization unnecessary for
   these weights, and which arm's weights now carry the long tail?
