# (PART) Time, Censoring, and Survival {-}

# Survival outcomes and censoring

```{r setup-ch10, include=FALSE}
source("R/setup.R")
```

Part II asked one question three ways and got one answer: whatever Cox-2
selectivity does to ulcer risk, the `ns` cohort cannot see it clearly,
and the crude signal of harm was confounding by indication. That question
had a convenient shape — one row, one treatment, and one binary outcome
per patient, so a risk was a proportion and an adjusted risk a weighted
proportion. Part III gives the convenience up. We move to `sta`, the
statin cohort, where the outcome is a cardiovascular event that may occur
at any point over a decade and where patients stop being observed at
wildly different times: because they die, because they disappear from the
data, or because the study ends. Confounding does not go away; a second
problem is layered on top of it, and this chapter is about learning to
see that second problem before adjusting for either.

## The statin question

The target trial discipline of Chapter 4 works exactly as before; only
some of the rows change.

| Protocol component | Target trial | Emulation using `sta` |
|---|---|---|
| Eligibility | Adults with an indication for lipid lowering and not already taking a statin | The patients assembled into `sta`, each entering at a qualifying visit (Chapter 6's new-user question) |
| Treatment strategies | (1) Initiate a statin; (2) do not initiate | `statin == 1` versus `statin == 0` |
| Assignment | Random allocation | Not randomized; conditioning on `sta_covs` under conditional exchangeability |
| Follow-up | Time zero at randomization, to the event, death, loss to follow-up, or `r admin_end` years | Time zero at the qualifying visit; `event_time` |
| Outcome | First cardiovascular event | `cv_event == 1` at `event_time` |
| Causal contrast | Intention-to-treat difference in risk at a stated horizon | Same, at 5 and `r admin_end` years |
| Analysis plan | Kaplan–Meier risk in each arm, contrasted at the horizon | Same, with adjustment standing in for randomization |

The two rows that changed, Follow-up and Causal contrast, changed
together. In Part II "risk" needed no qualifier because everyone was
followed for the same nominal period. Here it does: a risk is always a
risk *by some time*, and the trial must name the horizon before the
analysis can estimate anything. Ignore that and the Part II machinery
still runs, which is the danger — there is a `cv_event` column of zeros
and ones, and `mean(cv_event)` will happily return a number.

```{r ch10-naive}
sta |>
  group_by(statin) |>
  summarise(n = n(), events = sum(cv_event), proportion = mean(cv_event),
            median_followup = median(event_time), .groups = "drop")
```

```{r ch10-naive-values, include=FALSE}
naive <- tapply(sta$cv_event, sta$statin, mean)
med   <- tapply(sta$event_time, sta$statin, median)
p_full <- mean(sta$event_time >= admin_end)
p_short <- mean(sta$event_time < 2)
```

The two proportions are `r sprintf("%.1f", 100 * naive[1])`% and
`r sprintf("%.1f", 100 * naive[2])`%, and neither is a risk. They are
counts of events divided by counts of people who were watched for
different lengths of time: median follow-up is
`r sprintf("%.1f", med[1])` years in one arm and
`r sprintf("%.1f", med[2])` in the other, only
`r sprintf("%.0f", 100 * p_full)`% of the cohort is followed the full
`r admin_end` years, and `r sprintf("%.0f", 100 * p_short)`% leave within
two. A logistic regression of `cv_event` on `statin` and `sta_covs` would
produce a tidy odds ratio for a quantity with no definition, since the
denominator of each "risk" mixes people observed for one year with people
observed for ten. Nor can the problem be cleaned up by restricting to
patients with complete follow-up: staying under observation for ten years
is not a baseline characteristic but something that happens after time
zero, and Chapter 5 named the cost of selecting on such a variable. We
need estimators built for unequal follow-up.

## Anatomy of time-to-event data

Time-to-event data have exactly two columns at their core: a time and an
indicator of what happened at that time. `R/setup.R` builds both, and it
is worth reading that code as the substantive statement it is rather than
as boilerplate.

```{r ch10-build, eval=FALSE}
admin_end <- 10  # administrative end of follow-up (years)

sta <- sta |>
  mutate(
    statin     = as.integer(statin_use == "Statin"),
    event_time = pmin(cv_time, death_time, loss_fu_time, admin_end, na.rm = TRUE),
    cv_event   = as.integer(!is.na(cv_time) & cv_time <= event_time)
  )
```

The raw cohort supplies three latent times. `cv_time` is when a
cardiovascular event occurred, and is `NA` for the
`r sprintf("%.0f", 100 * mean(is.na(sta$cv_time)))`% of patients in whom
one never did; `death_time` and `loss_fu_time` are the analogous times of
death and of dropping out of observation, each `NA` for patients to whom
that never happened. No patient's data continue past the earliest of
these, so `pmin(..., na.rm = TRUE)` — with `admin_end` as a fourth
competitor that every patient always faces — gives `event_time`, the
moment observation stops. Then `cv_event` asks a single question about
that moment: was the thing that stopped observation the outcome itself?
When it was not, the patient is **censored** — we know only that no
cardiovascular event had occurred by `event_time`, not that none ever
will. Tabulating which of the four possibilities came first shows what
the cohort is made of.

```{r ch10-reasons}
ended <- sta |>
  mutate(ended_by = case_when(
    cv_event == 1                                     ~ "CV event",
    !is.na(death_time)   & death_time   <= event_time ~ "Death",
    !is.na(loss_fu_time) & loss_fu_time <= event_time ~ "Lost to follow-up",
    TRUE                                              ~ "Administrative end"
  ))

ended |>
  count(statin, ended_by) |>
  pivot_wider(names_from = statin, values_from = n, names_prefix = "statin_")
```

```{r ch10-reason-values, include=FALSE}
n_ev  <- sum(sta$cv_event)
n_dth <- sum(ended$ended_by == "Death")
d_by_arm <- tapply(as.integer(ended$ended_by == "Death"), ended$statin, mean)
```

`r n_ev` patients of `r nrow(sta)` have an observed cardiovascular event;
everyone else is censored, most of them by loss to follow-up or by the
administrative end. Censoring at the administrative end is the benign
kind. It happens at a calendar boundary that has nothing to do with any
patient's prognosis, and every estimator in this chapter handles it
without complaint.

Death is the uncomfortable case. `r n_dth` patients died before having a
cardiovascular event, and the code above treats them exactly like the
patients who moved away: censored, contributing follow-up up to the
moment of death and nothing after. That is a substantive assumption, and
a strong one. A patient who dies has not merely become unobservable — she
can never have the outcome, so treating her as though she were still at
risk in some unseen continuation of the study makes the estimated risk a
risk in a hypothetical world where death from other causes has been
removed. Worse, death is not evenly distributed here: it ends follow-up for
`r sprintf("%.1f", 100 * d_by_arm[2])`% of statin initiators against
`r sprintf("%.1f", 100 * d_by_arm[1])`% of non-initiators, so whatever
distortion the assumption introduces is not the same in the two arms.
The honest treatments of a **competing risk** are to report the
cumulative incidence function, which gives the fraction who actually
experience a cardiovascular event first and leaves the deaths where they
are, or to define a composite outcome of cardiovascular event or death.
We take the conventional route in this book because it keeps the
censoring machinery of Chapter 11 in one piece, and we flag it as a
limitation rather than a solution — though not before returning to it
below, once the Kaplan–Meier estimator has given us something to compare
against.

## Kaplan–Meier

The Kaplan–Meier estimator solves the unequal-follow-up problem by
refusing to compute a single proportion. It walks forward through the
observed event times, asking at each one what fraction of the patients
*still under observation just before that instant* survived it, and
multiplies those conditional fractions together. A patient censored at
2.3 years contributes to every factor up to 2.3 years and then quietly
leaves the denominator, so her partial information is used without her
absence being mistaken for good news. The product is the survival curve
$S(t)$, and one minus it is the cumulative incidence — the risk by time
$t$ — which is what we want to contrast.

```{r km}
library(survival)
km <- survfit(Surv(event_time, cv_event) ~ statin, data = sta)
km_tidy <- broom::tidy(km) |>
  mutate(statin = if_else(str_detect(strata, "=1"), "Statin", "No statin"),
         risk = 1 - estimate)
ggplot(km_tidy, aes(time, risk, color = statin)) +
  geom_step() + labs(x = "Years", y = "Cumulative incidence of CV event")
```

`survfit()` fits one curve per level of the right-hand side, and
`broom::tidy()` returns it as a data frame with one row per event time —
`time`, `estimate` for $S(t)$, pointwise confidence limits, and a
`strata` label from which the arm is recovered with `str_detect()`. Since
this picture recurs throughout survival analysis, it is worth wrapping
once, with confidence bands added and an argument for weights that a
weighted analysis can supply.

```{r ch10-plot-km-def, eval=FALSE}
plot_km <- function(data, treat, time = "event_time", event = "cv_event",
                    wt = NULL, labels = NULL) {
  form <- reformulate(treat, sprintf("Surv(%s, %s)", time, event))
  fit  <- survfit(form, data = data,
                  weights = if (!is.null(wt)) data[[wt]])
  curves <- broom::tidy(fit) |>
    mutate(arm  = str_remove(strata, ".*="),
           arm  = if (is.null(labels)) arm else unname(labels[arm]),
           risk = 1 - estimate)
  ggplot(curves, aes(time, risk, colour = arm, fill = arm)) +
    geom_ribbon(aes(ymin = 1 - conf.high, ymax = 1 - conf.low),
                alpha = 0.15, colour = NA) +
    geom_step() +
    scale_y_continuous(labels = scales::percent) +
    labs(x = "Years since time zero", y = "Cumulative incidence of CV event",
         colour = NULL, fill = NULL) +
    theme_minimal()
}
```

That code lives in `R/km.R`, so later analyses can source it rather than
redefine it. The `wt` argument names a column of weights and is inert
unless one is supplied; it is how a weighted analysis would pass
censoring weights.

```{r ch10-plot-km}
source("R/km.R")

plot_km(sta, "statin", labels = c("0" = "No statin", "1" = "Statin"))
```

```{r ch10-km-values, include=FALSE}
h <- summary(km, times = c(5, admin_end))
km_risk <- matrix(1 - h$surv, nrow = 2, byrow = TRUE,
                  dimnames = list(c("no", "statin"), c("y5", "y10")))
```

Read the curves at a horizon and the contrast becomes a statement someone
could act on. At five years the estimated risk is
`r sprintf("%.1f", 100 * km_risk["no", "y5"])`% without a statin against
`r sprintf("%.1f", 100 * km_risk["statin", "y5"])`% with one; at
`r admin_end` years, `r sprintf("%.1f", 100 * km_risk["no", "y10"])`%
against `r sprintf("%.1f", 100 * km_risk["statin", "y10"])`%. Both
`r admin_end`-year risks exceed the naive proportions we started with,
for the obvious reason: the naive calculation charged each arm only with
the events its censored patients stayed under observation long enough to
have. These are crude curves, mind — nothing has been adjusted, the arms
differ at baseline, and the outcomes in `sta` are simulated. Treat the
numbers as an illustration of what the estimator computes, not as
evidence about statins.

## Death as a competing risk

Now return to the deaths. Every risk above is a **net** risk: a risk in a
hypothetical world where the cardiovascular event is the only thing that
can happen to a patient, death from other causes having been abolished.
The alternative leaves the deaths where they fall and estimates the
**cumulative incidence function** — the fraction of patients who actually
experience a cardiovascular event first, in a world that still contains
death. That is a **crude** risk. Neither quantity is the wrong one; they
answer different questions, and which to report depends on whether the
reader wants to know what treatment would do absent competing mortality
or what will happen to patients like these.

Coding it needs a three-valued status rather than a yes-or-no one. Death
competes only when it arrives first: a patient who has a cardiovascular
event and dies a year later is a cardiovascular event and nothing else.

```{r ch10-competing-type}
sta_cr <- sta |>
  mutate(event_type = case_when(
    cv_event == 1                                 ~ 1L,  # CV event first
    !is.na(death_time) & death_time <= event_time ~ 2L,  # death first
    TRUE                                          ~ 0L   # still censored
  ))

table(sta_cr$event_type, sta_cr$statin)
```

The `1` row recovers the `r n_ev` cardiovascular events exactly, and the
`2` row isolates the `r n_dth` patients whose follow-up ended in death
with no cardiovascular event before it — the patients the Kaplan–Meier
analysis quietly treated as though they had merely moved away. Handing
`survfit()` a *factor* status instead of an indicator switches it into
multi-state mode: the first level is censoring, and each remaining level
gets a curve of its own. The cardiovascular curve is the
**Aalen–Johansen** estimate, which for a single competing event is the
cumulative incidence function we are after.

```{r ch10-competing-cif}
cr   <- survfit(Surv(event_time, factor(event_type)) ~ statin, data = sta_cr)
cr_h <- summary(cr, times = c(5, admin_end))
cif  <- matrix(cr_h$pstate[, "1"], nrow = 2, byrow = TRUE,
               dimnames = list(c("no", "statin"), c("y5", "y10")))
round(100 * cif, 2)
```

At five years the crude risk is
`r sprintf("%.2f", 100 * cif["no", "y5"])`% without a statin against the
net `r sprintf("%.2f", 100 * km_risk["no", "y5"])`%, and
`r sprintf("%.2f", 100 * cif["statin", "y5"])`% with a statin against the
net `r sprintf("%.2f", 100 * km_risk["statin", "y5"])`%. The crude risk is
lower in both arms, and it must be: some patients die before they can have
the event, and the cumulative incidence function counts them among those
who did not have it while the Kaplan–Meier estimator removes them from the
denominator and lets the remaining patients speak for them. Here the two
differ by less than a tenth of a percentage point, because death before a
cardiovascular event is uncommon in the first five years. By
`r admin_end` years the gap has grown, to
`r sprintf("%.2f", 100 * (km_risk["no", "y10"] - cif["no", "y10"]))` points
in the untreated arm and
`r sprintf("%.2f", 100 * (km_risk["statin", "y10"] - cif["statin", "y10"]))`
in the treated — still small, but moving as the mechanism predicts once
more of the cohort has died. The size of the discrepancy is governed by
how much competing death accumulates over the horizon, so the same code
applied to an older cohort followed for twenty years would separate the
two curves visibly.

The regression analog of this estimator is **Fine–Gray** subdistribution
hazards regression, which models the cumulative incidence function
directly the way Cox regression models the cause-specific hazard. We do
not pursue it here; the point of this section is to make the assumption
behind the rest of the chapter visible and quantified rather than merely
disclosed.

## Risks versus hazards

The Kaplan–Meier curve is built from conditional pieces, and those pieces
have a name of their own. The **hazard** at time $t$ is the instantaneous
rate of the event among those still event-free and still under
observation at $t$; the **cumulative risk** by $t$ is the probability of
having had the event by then, starting from time zero. Cox proportional
hazards regression, the default survival model in most of applied
epidemiology, estimates a ratio of hazards.

```{r ch10-hazard}
coxph(Surv(event_time, cv_event) ~ statin, data = sta) |>
  broom::tidy(exponentiate = TRUE, conf.int = TRUE)
```

```{r ch10-hazard-split}
early <- sta |> mutate(t = pmin(event_time, 2),
                       e = cv_event * (event_time <= 2))
late  <- sta |> filter(event_time > 2)

bind_rows(
  "0-2 years"  = broom::tidy(coxph(Surv(t, e) ~ statin, data = early),
                             exponentiate = TRUE, conf.int = TRUE),
  "after 2 years" = broom::tidy(coxph(Surv(event_time, cv_event) ~ statin,
                                      data = late),
                                exponentiate = TRUE, conf.int = TRUE),
  .id = "period"
)
```

```{r ch10-hazard-values, include=FALSE}
hr_all <- broom::tidy(coxph(Surv(event_time, cv_event) ~ statin, data = sta),
                      exponentiate = TRUE)$estimate
hr_e <- broom::tidy(coxph(Surv(t, e) ~ statin, data = early),
                    exponentiate = TRUE)$estimate
hr_l <- broom::tidy(coxph(Surv(event_time, cv_event) ~ statin, data = late),
                    exponentiate = TRUE)$estimate
```

One hazard ratio for the whole decade comes out at
`r sprintf("%.2f", hr_all)`. Split the decade in two and the first
interval gives `r sprintf("%.2f", hr_e)` and the remainder
`r sprintf("%.2f", hr_l)`. The single number was an average over periods
that do not agree, which is the mildest of the interpretive problems.

The serious one is what the second interval conditions on. Its risk set
holds only patients who reached two years event-free and still observed,
and surviving two years is not a baseline characteristic — it is an
outcome, influenced by treatment and by every unmeasured prognostic
factor a patient carries. Comparing arms within that risk set conditions
on a post-treatment common effect: the collider stratification of Chapter
5, arriving uninvited. If a statin genuinely prevents early events, the
statin patients still at risk at two years include people the other arm
has already lost, and the arms are no longer exchangeable however
carefully we adjusted at time zero. This is what Hernán (2010) called
the hazard ratio's **built-in selection bias**, and the reasoning above
follows his "The hazards of hazard ratios" editorial. Every hazard
beyond the first instant is conditional on survival, so a hazard ratio
is not a contrast of two potential outcomes
in a common population, and our two period-specific estimates cannot
distinguish a truly time-varying effect from selection. Nor is
proportional hazards a technical detail a test can rescue; the assumption
is that this awkward ratio happens to be constant.

Risks do not have this problem. $\Pr(Y(1) \le t)$ and
$\Pr(Y(0) \le t)$ are both defined over everyone who started, so their
difference or ratio is a causal contrast in the sense of Chapter 3. This
book therefore reports risks at named horizons — five years and
`r admin_end` years for `sta` — and treats hazard ratios, when they
appear at all, as a modeling device on the way to a risk curve rather
than as the estimand.

## When censoring lies

Kaplan–Meier removes a censored patient from the denominator and carries
on with the rest, which is a fair thing to do only if the patients who
remain represent the patients who left. Formally, censoring must be
**non-informative**: at each time, those still under observation must
have the same hazard as those censored just before, conditional on
whatever we have adjusted for. Administrative censoring satisfies this by
construction. Loss to follow-up need not, and in `sta` it is the largest
single reason follow-up ends.

Since censoring is not random by design, the first thing to do is look at
who leaves. We compare patients censored by loss to follow-up in the
first two years against everyone else on the baseline covariates we have.

```{r ch10-loss}
loss <- sta |>
  mutate(early_loss = as.integer(!is.na(loss_fu_time) &
                                 loss_fu_time <= 2 &
                                 event_time == loss_fu_time))

loss |>
  group_by(early_loss) |>
  summarise(n = n(), age = mean(age), statin = mean(statin),
            hypertension = mean(hypertension == "Yes"),
            heart_failure = mean(heart_failure == "Yes"),
            kidney_disease = mean(chronic_kidney_disease == "Yes"),
            tobacco_use = mean(tobacco_use == "Yes"), .groups = "drop") |>
  knitr::kable(digits = 3)
```

```{r ch10-loss-values, include=FALSE}
n_el <- sum(loss$early_loss)
smd <- function(v) {
  x <- if (is.numeric(loss[[v]])) loss[[v]] else as.integer(loss[[v]] == "Yes")
  m <- tapply(x, loss$early_loss, mean); s <- tapply(x, loss$early_loss, sd)
  (m[2] - m[1]) / sqrt((s[1]^2 + s[2]^2) / 2)
}
smds <- vapply(c("age", "hypertension", "heart_failure",
                 "chronic_kidney_disease", "tobacco_use"), smd, numeric(1))
max_smd <- max(abs(smds))
```

The `r n_el` patients lost early are slightly younger and slightly less
hypertensive than the rest, but every standardized difference in the
table is below `r sprintf("%.2f", ceiling(100 * max_smd) / 100)` in
absolute value, and loss is about equally common in the two arms. On the
measured variables, then, this cohort's censoring looks close to
non-informative — a genuinely reassuring finding and a strictly limited
one. The comparison speaks only to covariates recorded at baseline, while
the assumption we need concerns whatever drives dropping out at each
moment of follow-up: a worsening symptom, a change of insurer, a decision
to stop attending. None of that is in the table.

The structure of the worry is one we have already drawn. Suppose an
unmeasured element of prognosis affects both the cardiovascular hazard
and the chance of staying under observation, and suppose treatment also
influences whether a patient stays.

```{r ch10-censoring-dag, message=FALSE}
library(ggdag)
dag_censor <- dagify(
  cv ~ statin + u,
  c  ~ statin + u,
  exposure = "statin", outcome = "cv",
  labels = c(statin = "Statin", cv = "CV event",
             u = "Unmeasured prognosis", c = "Lost to follow-up")
)
ggdag(dag_censor, text = FALSE, use_labels = "label") + theme_dag()
```

Loss to follow-up is a collider: treatment points into it and unmeasured
prognosis points into it. Every risk set beyond time zero consists of
patients who have not yet been lost, so the analysis conditions on that
collider at every step, whether or not anyone wrote a `filter()` call —
the same argument Chapter 5 made about selection bias, now indexed by
time. The consequence is an association between `statin` and `cv` running
through `u` that no amount of baseline adjustment removes, because the
bias is created after time zero by who is left to observe.

The fix follows the logic of Chapter 9 rather than of Chapter 7. If we
cannot condition our way out of a collider, we can reweight: model the
probability of remaining uncensored, given treatment and the covariates
that predict dropping out, and give each patient still under observation
a weight inversely proportional to that probability, so that those who
remain stand in for the similar patients who left. Those are inverse
probability of censoring weights, they multiply cleanly against the
treatment weights we already know how to build, and Chapter 11 constructs
them for exactly the curves we have just drawn.

## Exercises

1. Using `km_tidy` from the `km` chunk, extract the estimated cumulative
   incidence at five years in each arm — take the last row in each arm
   with `time <= 5` — and report the risk difference and risk ratio.
   Confirm your answer against `summary(km, times = 5)`. Why is taking
   the last row at or before the horizon the right way to read a step
   function, rather than interpolating between rows?
2. Build a survival outcome for all-cause death. Do not reuse
   `event_time`, which is already truncated at cardiovascular events:
   build a fresh follow-up time from `death_time`, `loss_fu_time`, and
   `admin_end` only, and set `death_event` to 1 when `death_time` is the
   earliest of those three. Pass the new columns to `plot_km()` through
   its `time` and `event` arguments to draw cumulative mortality by
   `statin`. Which arm has higher mortality, and how does that picture
   change your reading of the cardiovascular curves in this chapter?
3. Run `plot_km()` with `treat = "statinpotency_use"`, which has three
   levels rather than two. Report the `r admin_end`-year risk in each of
   the three groups. What would have to be true for the contrast between
   the low- and high-potency groups to be interpretable as a causal
   effect of potency, and which row of this chapter's target trial table
   is doing the most work in that argument?
