# Inverse probability of censoring weights

```{r setup-ch11, include=FALSE}
source("R/setup.R")
```

Chapter 10 closed with a diagram that placed loss to follow-up between
treatment and an unmeasured element of prognosis, making every risk set
past time zero a selected sample and every Kaplan–Meier factor a
comparison among survivors of that selection. It also promised that we
already have the tool for the problem. Chapter 9 encountered a problem of
the same shape, namely a variable associated with treatment that we could
not condition on without introducing bias elsewhere, and addressed it by
reweighting rather than conditioning. This chapter builds the
corresponding weights for censoring, which requires reshaping the cohort
into a form we will use for the rest of Part III.

## Weighting returns

The two estimators apply the same idea to different columns. An
inverse probability of treatment weight is the reciprocal of the
probability of the treatment a patient actually received, $1/P(A \mid
L)$, and it builds a pseudo-population in which treatment is unrelated to
$L$. An **inverse probability of censoring weight** is the reciprocal of
the probability of the *observation status* a patient actually had: among
patients still under observation at time $t$, each is weighted by one
over her probability of having remained uncensored that long. She then
stands in for herself and for the patients like her who dropped out, and in
the reweighted data the people we can still see represent the people we
started with.

One structural difference matters. Treatment here happens once, at time
zero, so its weight is a single number per patient. Censoring accumulates:
a patient must survive being censored in the first year, and then the
second, and so on, and the chance of getting through all of it is a
product of the chances of getting through each piece. Splitting follow-up
into intervals $k = 1, \dots, K$ and writing $C_k = 1$ for censoring
during interval $k$, the weight in effect at interval $t$ is

$$W^{C}(t) = \prod_{k \le t} \frac{1}{\Pr(C_k = 0 \mid C_{k-1} = 0, A, L)},$$

and its stabilized version replaces the 1 in each numerator with
$\Pr(C_k = 0 \mid C_{k-1} = 0)$, a probability that depends on time but
not on covariates. As in Chapter 9 the numerator cannot undo the
balancing the denominator achieves; it rescales, so that a stabilized
weight averages near 1 and a mean far from 1 signals a misbehaving
censoring model.

What the weights buy is a weaker but still unverifiable assumption.
Kaplan–Meier requires censoring to be non-informative outright, whereas
the weighted analysis requires it to be non-informative *given the
covariates in the censoring model*, which is the survival-data version of
conditional exchangeability. Two things are needed to fit it: a data
structure with one row per patient per interval, and a model for
censoring in that structure.

## Person-period data

The step that makes everything else straightforward is to stop treating
each patient as one row. Cut follow-up into yearly intervals and give a
patient one row for each interval she survives into, carrying her
baseline covariates along; the row records the interval, whether the
outcome happened in it, and whether censoring happened in it. A patient
followed 3.1 years contributes four rows, the last a stub from year 3 to
year 3.1. `survival::survSplit()` does the cutting, and wrapping it fixes
the column names we will rely on later.

```{r ch11-pp-def, eval=FALSE}
make_person_period <- function(data, time, event, cuts) {
  data <- data |> mutate(.row_id = dplyr::row_number())
  survival::survSplit(
    as.formula(paste0("Surv(", time, ", ", event, ") ~ .")),
    data = data, cut = cuts, start = "tstart", end = "tstop",
    event = event
  ) |> as_tibble()
}
```

The `. ~` on the right of the formula keeps every other column, so
covariates are repeated down a patient's rows unchanged. `.row_id` is
added before splitting and is what glues the rows back into people —
needed for the running product of weights and for the cluster-robust
variance at the end. The event we split on is censoring, defined as
being lost to follow-up before having a cardiovascular event; the next
section takes up why that definition and not another. Because
`survSplit()` cannot make an interval out of nothing, we first set aside
the handful of patients whose follow-up ends at time zero.

```{r ch11-pp, message=FALSE, warning=FALSE}
library(survival)
source("R/ipcw.R")

sta_cens <- sta |>
  filter(event_time > 0) |>
  mutate(
    censored  = as.integer(!is.na(loss_fu_time) & loss_fu_time <= event_time &
                             cv_event == 0),
    cens_time = event_time
  )

pp <- make_person_period(sta_cens, time = "cens_time", event = "censored",
                         cuts = 1:9) |>
  mutate(event = as.integer(cv_event == 1 &
                            event_time > tstart & event_time <= tstop))
```

Two details of bookkeeping deserve naming, because getting either wrong
silently changes the analysis. First, `cens_time` is a copy of
`event_time`: `survSplit()` consumes the time variable it splits on and
replaces it with `tstart` and `tstop`, so splitting on the copy leaves
the original `event_time` intact as an ordinary carried-along column.
Second, the outcome indicator has to be rebuilt by hand. The `censored`
column that comes back from `survSplit()` is already interval-specific —
1 only in the interval where censoring happened — but `cv_event` is
still the patient-level value, repeated down every row, so a patient with
an event would appear to have one in each of her intervals. The
`event_time > tstart & event_time <= tstop` comparison puts it in the one
interval
that contains it. Look at two patients, one lost to follow-up and one
with an event, before and after expansion.

```{r ch11-one-person}
mid <- sta_cens$event_time > 2 & sta_cens$event_time < 3
ids <- c(first(which(sta_cens$censored == 1 & mid)),
         first(which(sta_cens$cv_event == 1 & mid)))

sta_cens |> mutate(.row_id = row_number()) |> slice(ids) |>
  select(.row_id, statin, event_time, cv_event, censored)

pp |> filter(.row_id %in% ids) |>
  select(.row_id, statin, tstart, tstop, event, censored)
```

The rows read as a small clinical history: at risk through each interval,
then one final interval in which either the outcome occurs or the patient
leaves observation. Because the two indicators are now built from three
different columns, they should be checked against the patient-level
cohort rather than taken on trust.

```{r ch11-checks}
tibble(
  quantity        = c("patients", "person-period rows", "outcome events",
                      "patients censored"),
  person_period   = c(n_distinct(pp$.row_id), nrow(pp), sum(pp$event),
                      sum(pp$censored)),
  patient_level   = c(nrow(sta_cens), NA, sum(sta_cens$cv_event),
                      sum(sta_cens$censored))
) |> knitr::kable()
```

```{r ch11-pp-values, include=FALSE}
n_drop  <- nrow(sta) - nrow(sta_cens)
n_ppl   <- n_distinct(pp$.row_id)
n_rows  <- nrow(pp)
n_ev    <- sum(pp$event)
n_cens  <- sum(pp$censored)
ok      <- n_ev == sum(sta_cens$cv_event) && n_cens == sum(sta_cens$censored) &&
           n_ev == sum(sta$cv_event)
```

Dropping the `r n_drop` patients with no follow-up leaves
`r format(n_ppl, big.mark = ",")` patients and
`r format(n_rows, big.mark = ",")` person-period rows, about
`r sprintf("%.1f", n_rows / n_ppl)` intervals each. The expansion
preserves `r n_ev` events — the same count as `sum(sta$cv_event)`, since
none of the dropped patients had one — and `r format(n_cens, big.mark = ",")`
censorings, matching the patient-level indicator exactly — both
identities hold, which is `r ok`. Checks as mundane as these are worth
printing every time, because an event indicator that is correct at the
patient level and wrong at the interval level produces plausible output
and a wrong answer.

The reason to pay this price in rows is that a logistic regression fitted
to person-period data estimates a **discrete-time hazard**. Each row asks
whether the event happened in that interval among patients who reached
it, which is exactly the conditional quantity Kaplan–Meier computes; a
logistic model of that indicator, pooled over intervals and with time on
the right-hand side, is a **pooled logistic** model and approximates a
proportional hazards model when the interval hazards are small. Ours
carries time as `splines::ns(tstop, 3)` rather than `factor(tstop)`,
spending three degrees of freedom instead of nine on a hazard we expect
to be smooth. Furthermore, because the hazards can be multiplied back up
into a survival curve, the model returns risks at the horizons the target
trial named, not only ratios.

## Building censoring weights

The censoring indicator above makes a substantive choice worth defending.
Of the ways follow-up ends in `sta`, only loss to follow-up is treated as
the censoring we weight for. Administrative censoring at
`r admin_end` years is excluded because it happens to everyone still
present at a calendar boundary that no covariate predicts — the benign
case of Chapter 10. Death is excluded for a less comfortable reason: it
is a competing risk that Chapter 10 chose to treat as censoring, and
putting it in the censoring model would weight the analysis toward a
world in which nobody dies of anything else, an estimand harder to defend
than the one we have. Both exclusions are substantive assumptions rather
than conveniences, and the second is the weakest point in this chapter's
analysis.

With the indicator fixed, the censoring model is a pooled logistic
regression of `censored` on time and the baseline covariates, fitted
across all person-period rows at once. In lecture, and in much of the
literature, censoring models are instead fitted with Cox proportional
hazards regression; the pooled logistic on person-period data is its
discrete-time analog, and with intervals this short the two give
essentially the same weights. Our estimator wraps the whole
sequence: fit the censoring model, fit the numerator model, take the
running product down each patient's rows, truncate, and fit the weighted
outcome model.

```{r ch11-ipcw-def, eval=FALSE}
ipcw_estimator <- function(pp_data, censor_model, outcome_formula,
                           id = ".row_id", numerator_model = NULL,
                           truncate_at = 0.99) {
  cens_fit <- glm(censor_model, family = binomial(), data = pp_data)
  pp_data <- pp_data |> mutate(p_uncens = 1 - predict(cens_fit, type = "response"))

  if (!is.null(numerator_model)) {
    num_fit <- glm(numerator_model, family = binomial(), data = pp_data)
    pp_data <- pp_data |> mutate(p_uncens_num = 1 - predict(num_fit, type = "response"))
  } else {
    pp_data <- pp_data |> mutate(p_uncens_num = 1)
  }

  pp_data <- pp_data |>
    group_by(.data[[id]]) |>
    arrange(tstop, .by_group = TRUE) |>
    mutate(w_cens = cumprod(p_uncens_num) / cumprod(p_uncens)) |>
    ungroup() |>
    mutate(w_cens = pmin(w_cens, quantile(w_cens, truncate_at)))

  # geeglm() looks up `id` and `weights` in the environment of the formula it
  # is handed; ours was built by the caller, so re-home it here.
  environment(outcome_formula) <- environment()

  fit <- geepack::geeglm(
    outcome_formula, id = pp_data[[id]], weights = pp_data$w_cens,
    family = binomial(), data = pp_data
  )
  list(fit = broom::tidy(fit, conf.int = TRUE), data = pp_data)
}
```

The `cumprod()` line carries the time-varying idea in a single
expression. Grouped by patient and ordered by interval, the weight at interval $t$ is
the running product of stabilized numerators over the running product of
uncensored probabilities, exactly the $W^{C}(t)$ written above. When
`numerator_model` is `NULL` the numerator is 1 and the weights are
unstabilized. Truncation reuses Chapter 9's `pmin(w, quantile(w, p))`,
here applied across all rows rather than all patients. The fit goes
through `geepack::geeglm()` with `id` set to the patient identifier, which
now does real work: a patient contributes several correlated rows, and
treating them as independent observations would understate the standard
error on two counts, once for the weights and once for the repeated rows.
Clustering on the patient addresses both. The function returns the tidied
fit together with the weighted data, so that the weights themselves can
be inspected.

```{r ipcw, warning=FALSE}
res <- ipcw_estimator(
  pp,
  censor_model    = reformulate(c("splines::ns(tstop, 3)", sta_covs), "censored"),
  outcome_formula = reformulate(c("statin", "splines::ns(tstop, 3)"), "event"),
  numerator_model = reformulate("splines::ns(tstop, 3)", "censored")
)

res$fit |> knitr::kable(digits = 3)
```

The outcome model is deliberately thin: `statin`, a spline in time, and
nothing else. Confounding of the treatment contrast is Chapter 9's job
and is not addressed here, so the `statin` coefficient in that table is a
censoring-corrected *crude* comparison, not an effect estimate. (A weighted binomial fit also
provokes a warning about non-integer successes, since the weights are
fractional; it is expected, and the sandwich variance is what we are
relying on rather than the likelihood.)

## Weight diagnostics

A weight is a claim about how many unseen patients each observed patient
represents, and such a claim should be examined before it is relied upon.

```{r ch11-wt-interval}
res$data |>
  group_by(interval = ceiling(tstop)) |>
  summarise(rows = n(), mean = mean(w_cens), sd = sd(w_cens),
            max = max(w_cens), .groups = "drop") |>
  knitr::kable(digits = 3)
```

```{r ch11-wt-values, include=FALSE}
w   <- res$data$w_cens
res_un <- ipcw_estimator(
  pp,
  censor_model    = reformulate(c("splines::ns(tstop, 3)", sta_covs), "censored"),
  outcome_formula = reformulate(c("statin", "splines::ns(tstop, 3)"), "event"),
  numerator_model = reformulate("splines::ns(tstop, 3)", "censored"),
  truncate_at = 1
)
w_raw  <- res_un$data$w_cens
n_cap  <- sum(w_raw > quantile(w_raw, 0.99))
ess    <- sum(w)^2 / sum(w^2)
```

Two features of the table deserve comment, of which the second is the
more important. The stabilized weights average
`r sprintf("%.3f", mean(w))` with a standard
deviation of `r sprintf("%.3f", sd(w))`, which is the reassuring
diagnostic, indicating that the censoring model is not producing a
pseudo-population of the wrong size. However, the whole distribution is
narrow. The largest weight
in the data is `r sprintf("%.2f", max(w_raw))` before truncation, and the
weights drift away from 1 only slowly as the intervals accumulate,
because a running product of numbers very close to 1 stays close to 1.
Kish's effective sample size, `r format(round(ess), big.mark = ",")` of
`r format(n_rows, big.mark = ",")` rows, records the same fact: these
weights cost almost no information.

That narrowness is a finding about this cohort rather than about the
method. The weights spread out only to the extent that the covariates in
the censoring model predict who drops out, and Chapter 10 already showed
that they do so weakly, since every standardized difference between the
early-loss patients and everyone else was under 0.10. A censoring model built from variables
that do not predict censoring returns probabilities near the overall
average, and those cancel almost exactly against the stabilizing
numerator.

Truncation, on this evidence, is a formality. Capping at the 99th
percentile touches `r format(n_cap, big.mark = ",")` of the
`r format(n_rows, big.mark = ",")` rows and changes the largest weight
from `r sprintf("%.2f", max(w_raw))` to `r sprintf("%.2f", max(w))`, so
nothing here hangs on it. The habit is worth keeping in any case, since a
running product is the kind of weight most liable to misbehave: a single
interval with a low estimated probability of remaining uncensored
multiplies into every later weight for that patient, and with ten
intervals and a richer censoring model, weights in the hundreds are
routine. The rule from Chapter 9 carries over, which is to report the
untruncated estimate, report it capped, and if they disagree materially,
say so rather than choosing between them.

## Did it matter?

The honest test is to fit the same pooled logistic model twice, once with
the censoring weights and once with weights of 1, and put the answers
beside each other. Since the model is a hazard model, we also multiply
the fitted hazards back into risks at the horizons Chapter 10 named.

```{r ch11-compare, warning=FALSE}
d      <- res$data
w_one  <- rep(1, nrow(d))
outcome_formula <- reformulate(c("statin", "splines::ns(tstop, 3)"), "event")

wfit <- geepack::geeglm(outcome_formula, id = d$.row_id, weights = d$w_cens,
                        family = binomial(), data = d)
ufit <- geepack::geeglm(outcome_formula, id = d$.row_id, weights = w_one,
                        family = binomial(), data = d)

bind_rows("IPCW weighted" = broom::tidy(wfit, conf.int = TRUE),
          "Unweighted"    = broom::tidy(ufit, conf.int = TRUE),
          .id = "analysis") |>
  filter(term == "statin") |>
  mutate(across(c(estimate, conf.low, conf.high), exp)) |>
  select(analysis, odds_ratio = estimate, conf.low, conf.high) |>
  knitr::kable(digits = 3)
```

```{r ch11-risk}
grid <- expand_grid(statin = 0:1, tstop = 1:admin_end)

risk_at <- function(fit) {
  grid |>
    mutate(h = predict(fit, newdata = grid, type = "response")) |>
    group_by(statin) |>
    mutate(risk = 1 - cumprod(1 - h)) |>
    ungroup() |>
    filter(tstop %in% c(5, admin_end))
}

bind_rows("IPCW weighted" = risk_at(wfit), "Unweighted" = risk_at(ufit),
          .id = "analysis") |>
  select(analysis, statin, years = tstop, risk) |>
  knitr::kable(digits = 4)
```

```{r ch11-compare-values, include=FALSE}
pull_or <- function(fit) {
  t <- broom::tidy(fit); exp(t$estimate[t$term == "statin"])
}
or_w <- pull_or(wfit); or_u <- pull_or(ufit)
rk <- function(fit, a, y) {
  r <- risk_at(fit); r$risk[r$statin == a & r$tstop == y]
}
rd_w <- rk(wfit, 1, 5) - rk(wfit, 0, 5)
rd_u <- rk(ufit, 1, 5) - rk(ufit, 0, 5)
km5 <- 1 - summary(survfit(Surv(event_time, cv_event) ~ statin, data = sta),
                   times = 5)$surv
```

The weighted odds ratio for `statin` is `r sprintf("%.3f", or_w)` against
`r sprintf("%.3f", or_u)` unweighted, and the five-year risk difference
moves from `r sprintf("%+.2f", 100 * rd_u)` percentage points to
`r sprintf("%+.2f", 100 * rd_w)`. Neither quantity moved materially, and
the confidence intervals are effectively identical. The weighted
five-year risks,
`r sprintf("%.1f", 100 * rk(wfit, 0, 5))`% without a statin and
`r sprintf("%.1f", 100 * rk(wfit, 1, 5))`% with one, also land near
Chapter 10's Kaplan–Meier estimates of
`r sprintf("%.1f", 100 * km5[1])`% and
`r sprintf("%.1f", 100 * km5[2])`% — reassurance that the pooled logistic
is estimating the same curve, with the remaining gap the price of
describing a hazard with three degrees of freedom.

This is the expected result, and it is worth examining rather than
apologizing for. Weights near 1 cannot move an estimate, and the weights
are near 1 because the variables we measured do not predict who left. The
correction was available, we applied it, and the data indicate that the
correction was not needed *for the censoring that baseline covariates can
see*. That qualification carries most of the weight of the sentence. If
dropping out is driven by something we did not record, such as a symptom
that worsened, a decision to stop attending, or the unmeasured prognosis
in Chapter 10's diagram, then the collider is still being conditioned on,
the estimate is still biased, and this analysis would look exactly as it
does now. A well-behaved weight distribution is evidence that the
censoring model found little to correct, and not evidence that there was
little to correct. The parallel with unmeasured confounding in Chapter 9
is uncomfortable but exact, and it is the reason for the sensitivity
analyses of Chapter 14.

One further element is deliberately missing from the model above. A complete
analysis of this question would weight for both problems at once: build
Chapter 9's stabilized treatment weight from a model for `statin` given
`sta_covs`, build this chapter's censoring weight for each interval,
multiply them per row, and fit the same pooled logistic marginal
structural model with the product as its weight. The two weights address
different arrows and compose cleanly, and their product is the standard
construction for time-to-event marginal structural models. Chapter 12
assembles it, in the service of a harder question than a point treatment.

## Exercises

1. Refit the censoring model without the comorbidity variables — keep
   only `age`, `sex`, `race`, `region`, and `year` alongside the spline in
   time. Report the mean, standard deviation, and maximum of the
   resulting weights and the `statin` odds ratio. Which of the two
   changes more, and what does that tell you about how much of the
   censoring signal the comorbidities carried?
2. Rerun `ipcw_estimator()` with `numerator_model = NULL` to get
   unstabilized weights. Report their mean and maximum and explain why
   the mean is no longer near 1. Then explain why the `statin` estimate
   barely changes even though the weights are much larger, and describe a
   situation in which stabilization would matter more.
3. Set `truncate_at = 0.95` and compare the weight distribution and the
   estimate against the 99th-percentile version. Given the largest
   untruncated weight in this cohort, was truncation ever going to matter
   here? Describe the features of a censoring model — number of
   intervals, strength of the covariate-censoring association — that
   would make the answer different.
