# Clone-censor-weight

```{r setup-ch12, include=FALSE}
source("R/setup.R")
```

Chapter 11 corrected a comparison for one thing that happens after time
zero, namely patients disappearing from view, and left another
untouched. Patients also stop the drug they started, and patients who
began untreated start one. Chapter 10's Kaplan–Meier curves ignore that
entirely: a patient assigned to the statin arm by her `statin` value at
time zero stays in that arm for ten years whatever she swallows
afterward. Although that is a respectable estimand, it is not the only
one worth estimating, and this chapter turns to the other one, which asks
what would happen if patients actually followed the strategy their arm
describes, and to the machinery that estimand requires.

## From point treatments to treatment strategies

The contrast in Chapter 10 was between two *point interventions*:
initiate a statin at time zero, or do not. Nothing was said about the
years after, so the estimate answers a question about assignment. In a
trial the analogue is the intention-to-treat effect, which a trial can
deliver without further assumptions, although it is rarely the question a
patient asks. The question underneath is usually the **per-protocol**
one: what happens if you initiate a statin *and stay on it* for the next
decade, compared with never initiating one? These are two strategies,
each sustained over time, rather than two decisions taken once.

Sustained strategies carry their own price. The intention-to-treat
contrast is identified by baseline exchangeability alone, because
assignment is complete at time zero and nothing later can confound it.
Once the strategy extends over time, adherence becomes part of the
treatment, and patients who adhere differ from those who do not, in that
they tend to be healthier and better attended, in ways that also predict
cardiovascular events. Deviating from protocol is a post-baseline
variable with its own causes, and it needs its own adjustment.

The obvious move is to keep the patients who followed protocol and
analyze them as though they had been randomized, and it is the wrong
one, wrong in a way Chapter 5 already named. Because deviation happens
*after* time zero, selecting on it conditions on the future. A patient
who has an event in year two never had the chance to stop her statin in
year six, so she stays in
the analysis; a patient who survived to year six and then stopped is
deleted from it retrospectively, along with her early event-free time.
The comparison is between people conditioned on their own subsequent
histories, and the resulting numbers cannot be read as risks under either
strategy.

## The clone-censor-weight recipe

The fix keeps the per-protocol estimand and repairs the selection with
weights. It has three steps. The strategy, and the formulation of the
censoring rule in terms of when a person's data cease to be compatible
with the protocol she was assigned, come from Hernán and Robins's
target-trial work (Hernán and Robins 2016; Hernán and Robins 2017).

**Clone.** Every eligible person is copied into each strategy arm whose
protocol her data are compatible with at baseline. A clone is the same
person entered into two different bookkeeping systems, each asking a
different question of her follow-up, rather than a duplicate observation
to be counted twice by accident.

**Censor.** Each clone is followed until her observed data first
contradict the strategy of the arm she sits in, at which point that clone
is *artificially censored* — removed from that arm's risk set, though she
may continue in the other arm's. Artificial censoring differs from the
censoring of Chapter 11 only in provenance: it is imposed by the analysis
rather than by the data, which means we know exactly what caused it and
can model it.

**Weight.** Because we created the censoring, we are obliged to correct
for it, with exactly Chapter 11's inverse probability of censoring
weight: each surviving clone at interval $t$ is weighted by the
reciprocal of her probability of having remained uncensored through $t$,
given time, arm, and covariates. Patients who stayed on protocol stand in
for the similar patients who did not.

Here the first step is nearly vacuous, and it is worth being explicit
about why. Our two strategies are distinguished by what happens at time
zero, that is, by whether a statin is initiated, and `statin` records
exactly that, with no grace period in which either strategy is still
possible. Each person's baseline data are therefore compatible with one
arm and one arm only, cloning assigns each person to her observed arm,
and the cohort we analyze has the same number of rows as the cohort we
started with, so that censoring and weighting do all of the work.

Literal cloning earns its name when the protocol allows a window. Suppose
the strategies were "initiate a statin within six months of the
qualifying visit" and "do not initiate within six months". A patient who
starts at month four is compatible with *both* protocols at time zero and
for the first four months: she is following the initiate-within-six-months
strategy the whole time, and she is also following the do-not-initiate
strategy right up until she starts. She is cloned into both arms,
contributes person-time to both, and her non-initiator clone is
artificially censored at month four while her initiator clone continues.
The grace period removes the immortal-time problem that classifying her
by a later event would create, at the cost of a patient who is
genuinely two rows-worth of follow-up. Our cohort has no
grace period, so we get the simple case; the recipe is identical either
way.

## Deviation in the statin cohort

The `sta` cohort supplies the deviation time directly. `nonadherence` is
the number of years until a patient departs from the strategy she began
with, and crucially it is defined in both arms, meaning different things
in each: statin initiators stopping, and untreated patients starting.

```{r ch12-nonadherence}
sta |>
  group_by(statin) |>
  summarise(n = n(), with_deviation = sum(!is.na(nonadherence)),
            median_years = median(nonadherence, na.rm = TRUE),
            .groups = "drop") |>
  knitr::kable(digits = 2)
```

Follow-up in the per-protocol analysis now ends at the earliest of five
things: the cardiovascular event, death, loss to follow-up, the
administrative end at `r admin_end` years — all inherited from
`event_time` — and deviation. The last is the new one, and the only one
we will weight for.

```{r ccw-data, message=FALSE, warning=FALSE}
library(survival)
source("R/ipcw.R")
source("R/ccw_estimator.R")

sta_ccw <- sta |>
  filter(event_time > 0) |>
  mutate(
    dev_time   = if_else(is.na(nonadherence), Inf, nonadherence),
    fu_end     = pmin(event_time, dev_time),
    deviated   = as.integer(dev_time <= event_time &
                              !(cv_event == 1 & event_time <= dev_time)),
    event_c    = as.integer(cv_event == 1 & event_time <= dev_time),
    split_time = fu_end
  )

pp <- make_person_period(sta_ccw, time = "split_time", event = "deviated",
                         cuts = 1:9) |>
  mutate(event = as.integer(event_c == 1 & fu_end > tstart & fu_end <= tstop))
```

Four points of bookkeeping deserve comment, since each changes the
analysis if it goes wrong. Patients who never deviate get
`dev_time = Inf`, so `pmin()` leaves their follow-up at `event_time`.
The `deviated` indicator is 1 when
deviation came first, with the `!(cv_event == 1 & ...)` clause settling
ties in the outcome's favour: a patient whose event and deviation are
recorded at the same instant had the event, and `event_c` — the outcome
indicator for this analysis — says so. `split_time` is a copy of `fu_end`
for the same reason Chapter 11 copied `event_time`: `survSplit()` consumes
the variable it splits on, and we need the original left standing to place
the outcome in its interval. And as in Chapter 11 the handful of patients
whose follow-up ends at time zero are dropped, since no interval can be
made out of nothing.

```{r ch12-checks}
tibble(
  quantity      = c("patients", "person-period rows", "outcome events",
                    "deviations", "rows flagged as both"),
  person_period = c(n_distinct(pp$.row_id), nrow(pp), sum(pp$event),
                    sum(pp$deviated), sum(pp$event == 1 & pp$deviated == 1)),
  patient_level = c(nrow(sta_ccw), NA, sum(sta_ccw$event_c),
                    sum(sta_ccw$deviated), 0)
) |> knitr::kable()
```

```{r ch12-check-values, include=FALSE}
# The "44,485 rows" quoted in the prose below is a literal taken from
# Chapter 11's render (its person-period expansion); update it there and here
# together if that expansion ever changes.
n_drop <- nrow(sta) - nrow(sta_ccw)
n_dev  <- sum(sta_ccw$deviated)
n_ev   <- sum(pp$event)
n_lost <- sum(sta_ccw$cv_event) - n_ev
by_arm <- sta_ccw |> group_by(statin) |>
  summarise(dev = sum(deviated), ev_all = sum(cv_event), ev_pp = sum(event_c))
ok <- n_ev == sum(sta_ccw$event_c) && n_dev == sum(pp$deviated) &&
      sum(pp$event == 1 & pp$deviated == 1) == 0
```

The two indicators agree with their patient-level counts and never
co-occur in a row, which is `r ok`. What the table also shows is the cost
of the estimand. Artificial censoring at deviation ends follow-up early
for `r format(n_dev, big.mark = ",")` patients, cutting the person-period
data from Chapter 11's 44,485 rows to
`r format(nrow(pp), big.mark = ",")`, and it discards `r n_lost` of the
`r sum(sta_ccw$cv_event)` cardiovascular events in the cohort because they
happened after their patient had already left her protocol. The statin arm
is hit hardest: `r by_arm$ev_pp[2]` of its `r by_arm$ev_all[2]` events
survive the censoring, against `r by_arm$ev_pp[1]` of
`r by_arm$ev_all[1]` among non-initiators. Per-protocol questions are
therefore always answered with less information than intention-to-treat
questions.

## Estimating per-protocol risk curves

The estimator is Chapter 11's weight construction pointed at the new
censoring indicator, with the outcome model replaced by direct weighted
hazards so that risk curves come back rather than coefficients.

```{r ch12-ccw-def, eval=FALSE}
ccw_estimator <- function(pp_data, censor_model, treat, id = ".row_id",
                          numerator_model = NULL, truncate_at = 0.99) {
  dev_fit <- glm(censor_model, family = binomial(), data = pp_data)
  pp_data <- pp_data |> mutate(p_stay = 1 - predict(dev_fit, type = "response"))

  if (!is.null(numerator_model)) {
    num_fit <- glm(numerator_model, family = binomial(), data = pp_data)
    pp_data <- pp_data |> mutate(p_stay_num = 1 - predict(num_fit, type = "response"))
  } else {
    pp_data <- pp_data |> mutate(p_stay_num = 1)
  }

  pp_data <- pp_data |>
    group_by(.data[[id]]) |>
    arrange(tstop, .by_group = TRUE) |>
    mutate(w = cumprod(p_stay_num) / cumprod(p_stay)) |>
    ungroup() |>
    mutate(w = pmin(w, quantile(w, truncate_at)))

  # A patient's last row is a stub ending at her exact follow-up time, so raw
  # `tstop` is continuous; hazards are pooled over the interval it falls in.
  pp_data |>
    mutate(interval = ceiling(tstop)) |>
    group_by(.data[[treat]], interval) |>
    summarise(hazard = weighted.mean(event, w), .groups = "drop_last") |>
    arrange(interval, .by_group = TRUE) |>
    mutate(surv = cumprod(1 - hazard), risk = 1 - surv) |>
    ungroup() |>
    rename(tstop = interval)
}
```

`p_stay` is the probability of *not* deviating in an interval, and the
`cumprod()` line is again the running product that turns per-interval
probabilities into a weight at time $t$. The last block is new: within an
arm and interval, `weighted.mean(event, w)` is the weighted discrete-time
hazard, and `cumprod(1 - hazard)` multiplies the hazards back into a
survival curve. This is a weighted Kaplan–Meier computed in
person-period form, and it needs no outcome model at all.

The deviation model does need thought, because deviation is not the same
event in the two arms. We let the arms differ in their time trend and in
two covariates by interacting `statin` with the spline and with `age` and
`diabetes`, rather than fitting two fully arm-specific models. The
interaction model is the more parsimonious of the two here and the fit
turns out to be well behaved, so we prefer it; had it been unstable, two
`glm()` calls stratified by arm would give the same weights with more
freedom and less pooling.

```{r ccw, warning=FALSE}
dev_form <- reformulate(c("splines::ns(tstop, 3) * statin", sta_covs,
                          paste0("statin:", c("age", "diabetes"))), "deviated")
num_form <- reformulate("splines::ns(tstop, 3) * statin", "deviated")

curves  <- ccw_estimator(pp, censor_model = dev_form, treat = "statin",
                         numerator_model = num_form)
crude   <- ccw_estimator(pp, censor_model = deviated ~ 1, treat = "statin",
                         numerator_model = deviated ~ 1)
```

The second call is a device worth knowing: identical numerator and
denominator models give every row a weight of exactly 1, so `crude`
is the same analysis with the artificial censoring left uncorrected.

```{r ch12-dev-check, include=FALSE}
dev_fit <- glm(dev_form, family = binomial(), data = pp)
b       <- coef(dev_fit)
p_dev   <- predict(dev_fit, type = "response")

dev_fit_c <- glm(update(dev_form, . ~ . - year + I(year - 2007)),
                 family = binomial(), data = pp)
p_dev_c   <- predict(dev_fit_c, type = "response")
```

Well behaved, here, means `glm()` converges with no aliased coefficient,
the largest coefficient other than the intercept is
`r sprintf("%.2f", max(abs(b[-1])))`, and the fitted probability of
deviating within an interval runs from
`r sprintf("%.4f", min(p_dev))` to `r sprintf("%.2f", max(p_dev))`.
The intercept is the exception at `r sprintf("%.0f", b[1])`, and it is a
matter of scale rather than instability: `sta_covs` carries `year` as an
uncentered calendar year, so the intercept is the log-odds extrapolated
back to year 0. Refitting with `I(year - 2007)` moves it to
`r sprintf("%.1f", coef(dev_fit_c)[1])` and changes no fitted probability
— the largest discrepancy is
`r sprintf("%.0e", max(abs(p_dev - p_dev_c)))` — so we leave `sta_covs`
as the rest of the book uses it.

```{r ch12-curves}
bind_rows("Weighted" = curves, "Unweighted" = crude, .id = "analysis") |>
  mutate(arm = if_else(statin == 1, "Statin, sustained", "No statin, sustained")) |>
  ggplot(aes(tstop, risk, colour = arm, linetype = analysis)) +
  geom_step() +
  scale_y_continuous(labels = scales::percent) +
  scale_x_continuous(breaks = 1:admin_end) +
  labs(x = "Years since time zero", y = "Cumulative incidence of CV event",
       colour = NULL, linetype = NULL) +
  theme_minimal()
```

Unlike `ipcw_estimator()`, this function returns only the curves, so the
weights must be rebuilt to be inspected.

```{r ch12-weights}
wts <- pp |>
  mutate(p_stay     = 1 - predict(glm(dev_form, family = binomial(), data = pp),
                                  type = "response"),
         p_stay_num = 1 - predict(glm(num_form, family = binomial(), data = pp),
                                  type = "response")) |>
  group_by(.row_id) |> arrange(tstop, .by_group = TRUE) |>
  mutate(w = cumprod(p_stay_num) / cumprod(p_stay)) |> ungroup()

wts |>
  group_by(interval = ceiling(tstop)) |>
  summarise(rows = n(), mean = mean(w), sd = sd(w), max = max(w),
            .groups = "drop") |>
  knitr::kable(digits = 2)
```

```{r ch12-wt-values, include=FALSE}
w <- wts$w
ess <- sum(w)^2 / sum(w^2)
```

These weights behave nothing like Chapter 11's. They still average near
`r sprintf("%.2f", mean(w))`, as stabilized weights should, but their
standard deviation is `r sprintf("%.2f", sd(w))` and the largest is
`r sprintf("%.0f", max(w))`, with the spread widening as the running
product accumulates and peaking in the seventh interval. Kish's effective sample size is
`r format(round(ess), big.mark = ",")` of
`r format(nrow(pp), big.mark = ",")` rows. These are the untruncated
weights, since the rebuilt version above stops short of the estimator's
final `pmin()`; inside `ccw_estimator()` the 99th percentile,
`r sprintf("%.2f", quantile(w, 0.99))`, becomes the ceiling. The reason is the contrast we
drew in Chapter 11: covariates that barely predicted loss to follow-up
predict adherence quite well, so the estimated probabilities of staying on
protocol vary across patients, and ten intervals of variation compound.

Extreme weights are the positivity assumption complaining, as in
Chapter 9. A weight of `r sprintf("%.0f", max(w))` says that somewhere in
covariate-and-time space staying on protocol was close to impossible,
since `r sum(p_dev > 0.5)` of the `r format(nrow(pp), big.mark = ",")`
rows carry a fitted deviation probability above one half, so that the few
patients there who did stay must speak for everyone who did not.
Positivity is a heavier demand for a sustained strategy than for a point
treatment, because it has to hold in *every* interval. Truncation at the
99th percentile, a formality in the last chapter, is doing visible work
here, and would be the first thing to vary in a sensitivity analysis.

A single number for the contrast needs an interval, and the weights make
the model-based one untrustworthy: the curve is a function of two fitted
models and a running product, so we resample patients instead. Each of
100 replicates draws `r format(nrow(sta_ccw), big.mark = ",")` patient
identifiers with replacement, reassembles their person-period rows under
fresh identifiers, and reruns the whole pipeline — fewer than Chapter 7's
200, because each replicate here refits two models and a running product.

```{r ch12-boot, warning=FALSE}
set.seed(20260822)
ids <- unique(pp$.row_id)

rd_at <- function(cv, years = 5) {
  r <- cv$risk[cv$tstop == years]
  r[2] - r[1]
}

rd_boot <- replicate(100, {
  draw <- tibble(.row_id = sample(ids, length(ids), replace = TRUE)) |>
    mutate(.boot = row_number())
  bs <- draw |>
    left_join(pp, by = ".row_id", relationship = "many-to-many") |>
    mutate(.row_id = .boot)
  rd_at(ccw_estimator(bs, censor_model = dev_form, treat = "statin",
                      numerator_model = num_form))
})

tibble(risk_difference = rd_at(curves),
       conf.low  = quantile(rd_boot, 0.025),
       conf.high = quantile(rd_boot, 0.975)) |>
  knitr::kable(digits = 4)
```

Reassigning identifiers with `.boot` is the detail that makes this
correct: a patient drawn twice must appear as two different people, or the
`group_by(.data[[id]])` inside the estimator would splice her duplicated
rows into one impossible history.

## Interpreting the per-protocol effect

```{r ch12-compare}
# Chapter 14 quotes the two clone-censor-weight five-year risks from this table
# (1.61% under sustained statin use, 5.78% under none) as literals; if anything
# here changes those numbers, update that chapter's `ch14-evalue-values` chunk.
km5 <- 1 - summary(survfit(Surv(event_time, cv_event) ~ statin, data = sta),
                   times = 5)$surv
drop5 <- 1 - summary(survfit(Surv(event_time, cv_event) ~ statin,
                             data = filter(sta_ccw, deviated == 0)),
                     times = 5)$surv
at5 <- function(cv, a) cv$risk[cv$statin == a & cv$tstop == 5]

tibble(
  analysis  = c("ITT-analog (Chapter 10)", "Naive per-protocol (drop deviators)",
                "Censor at deviation, unweighted", "Clone-censor-weight"),
  no_statin = c(km5[1], drop5[1], at5(crude, 0), at5(curves, 0)),
  statin    = c(km5[2], drop5[2], at5(crude, 1), at5(curves, 1))
) |>
  mutate(difference = statin - no_statin) |>
  knitr::kable(digits = 4)
```

```{r ch12-compare-values, include=FALSE}
rd_pp  <- rd_at(curves)
rd_itt <- km5[2] - km5[1]
ci     <- quantile(rd_boot, c(0.025, 0.975))
```

Consider the difference column. The intention-to-treat analogue puts the
five-year risk difference at `r sprintf("%+.1f", 100 * rd_itt)` percentage
points; the clone-censor-weight estimate is
`r sprintf("%+.1f", 100 * rd_pp)`, with a bootstrap interval of
`r sprintf("%.1f", 100 * ci[1])` to `r sprintf("%.1f", 100 * ci[2])`. The
per-protocol contrast is larger, which is the expected direction, since a
strategy followed for five years should do more than a strategy assigned
and then abandoned, and in these simulated outcomes the gap is
substantial. It rests on thin evidence, however, because the
sustained-statin curve is built from `r by_arm$ev_pp[2]` events, which is
why it sits flat for whole intervals and should not be read closely. The
naive analysis that deletes deviators lands between the two, close enough
to be mistaken for a reasonable answer, although it was arrived at by
conditioning on the future.

The weights also matter within the per-protocol column: uncorrected,
censoring at deviation understates risk in both arms, because the patients
who stay on protocol are the ones who stay well, and weighting them back
up to represent the deviators they resemble raises both curves. That the
correction moves the arms in the same direction is a reminder that
artificial censoring is not a bias with a sign we can guess.

What did the extra estimand cost in assumptions? Chapter 10's contrast
needed baseline exchangeability. This one needs that *and* no unmeasured
common causes of deviation and outcome, at every interval: whatever makes
a patient stop her statin in year four must be captured, at that time, by
the variables in `dev_form`. Ours contains baseline covariates and time
and nothing else, which is a strong claim about a cohort in which the
obvious reasons for stopping a statin — a new symptom, a rising liver
enzyme, a change in perceived prognosis — are time-varying and
unrecorded. If they matter, the weights are built from the wrong model
and this analysis looks exactly as it does now, which is Chapter 11's
uncomfortable symmetry once more. It needs positivity at every interval
besides: in each stratum the model uses, some patients like her must
actually have stayed on protocol. Unlike exchangeability, that assumption
leaves a trace in the weights we have just examined.

Two further limits belong on the record. The comparison here is still
crude with respect to baseline confounding: the arms differ at time zero
and nothing in this chapter fixes that. The complete recipe, promised at
the end of Chapter 11, multiplies three weights per row — Chapter 9's
stabilized treatment weight, Chapter 11's loss-to-follow-up weight, and
this chapter's deviation weight — and the product goes into exactly the
calculation above. We carry one factor at a time so the mechanism stays
visible. Furthermore, "stay on it" is a crude version of a sustained
strategy, since real protocols specify doses, gaps, and restarts, and
each such refinement changes both the estimand and the deviation model.

## Exercises

1. Refit the deviation model with only `age`, `sex`, and a spline in
   `tstop`, dropping the comorbidities and the arm interactions. Compare
   the weight distribution and the five-year risk difference against the
   full model. Which moves more, and what does that tell you about where
   the adherence signal lives?
2. Report the risk difference at three years instead of five, with a
   bootstrap interval. Compare its width with the five-year interval and
   account for the difference in terms of the events and the weights that
   each horizon uses.
3. Suppose the protocol allowed a six-month grace period for initiation.
   Describe what would change in `sta_ccw`: who would be cloned into both
   arms, when each clone's artificial censoring would occur, and why the
   `.row_id` column would no longer identify a patient. Would you expect
   the per-protocol estimate to move toward the intention-to-treat one, or
   away from it?
