# (APPENDIX) Appendices {-}

# R and RStudio setup

This appendix collects the one-time setup we assumed you had already
done, back in Chapter 2. If everything already works, meaning that you
can knit a chapter and see `ns` and `sta` load without complaint, you can
skip past it.

## Installing R and RStudio

Install R first, then RStudio. Both are free.

- **R**: download and install the current release from
  [CRAN](https://cran.r-project.org/), the Comprehensive R Archive
  Network. Pick the installer for your operating system and accept the
  defaults.
- **RStudio**: download and install RStudio Desktop from
  [posit.co/download/rstudio-desktop](https://posit.co/download/rstudio-desktop/).
  RStudio is an editor for R, not a replacement for it, so R must already
  be installed before RStudio is of any use.

Open RStudio once to confirm that it starts and finds your R
installation (it will say so in the Console pane on startup).

## Installing the packages this book uses

Every chapter in this book leans on the `tidyverse` for data manipulation
and plotting, and most chapters bring in one or two special-purpose
packages on top of that. Install all of them at once by pasting the
following into the R console and running it:

```{r app-install-packages, eval = FALSE}
install.packages(c(
  "tidyverse",  # dplyr, ggplot2, purrr, and friends
  "bookdown",   # builds this book; only needed if you render it yourself
  "dagitty",    # causal diagrams, Chapter 5
  "ggdag",      # plotting causal diagrams, Chapter 5 and later chapters
  "survival",   # Surv(), survfit(), survSplit(), Chapters 10-12
  "cobalt",     # covariate-balance tables and plots, Chapter 8
  "geepack",    # weighted GEE, the engine behind iptw_estimator() and
                # ipcw_estimator(), Chapters 9 and 11
  "broom",      # tidy() for turning model objects into tibbles
  "remotes"     # installs packages from GitHub, including NAMCS below
))
```

One package used in Chapters 11 and 12, `splines`, ships with base R
and never needs installing.

## Installing the NAMCS teaching package

The two cohorts this book uses throughout, `ns` and `sta`, are not on
CRAN. They live in a companion package, `NAMCS`, distributed from
GitHub. With `remotes` installed from the previous step, run:

```{r app-install-namcs, eval = FALSE}
remotes::install_github("alanbrookhart/NAMCS")
```

If that command fails, it is almost always one of two things: `remotes`
is not installed (revisit the previous step), or GitHub is unreachable
from your network (try again from a different network, or ask your
instructor for the data files directly).

## Loading everything a chapter needs

Every chapter and appendix in this book opens with a chunk that runs

```{r app-setup-source, eval = FALSE}
source("R/setup.R")
```

`R/setup.R` loads the tidyverse and the `NAMCS` package, loads `ns` and
`sta`, and derives the handful of analysis columns and covariate lists
(`ns_covs`, `sta_covs`, and so on) that later chapters take for granted.
Sourcing it is the whole of what you need to do to follow along with any
chapter's code on its own, without having run the earlier chapters
first.

```{r setup-app, include = FALSE}
source("R/setup.R")
```

# NAMCS codebook

Both teaching cohorts carry many more columns than the ones we actually
use in adjustment models (`ns_covs` and `sta_covs` in `R/setup.R` are
short lists curated from a much wider survey). This appendix is the
reference for the rest: a full variable-by-variable codebook, generated
directly from the data rather than typed out by hand, so it will always
match whatever is actually in the installed `NAMCS` package.

## The NSAID cohort (`ns`)

```{r app-nsaid-codebook}
tibble(
  variable = names(ns),
  type = map_chr(ns, ~ class(.x)[1]),
  example_values = map_chr(ns, ~ paste(head(unique(.x), 3), collapse = ", "))
) |>
  knitr::kable(caption = "Variables in the `ns` (NSAID) cohort.")
```

## The statin cohort (`sta`)

```{r app-statin-codebook}
tibble(
  variable = names(sta),
  type = map_chr(sta, ~ class(.x)[1]),
  example_values = map_chr(sta, ~ paste(head(unique(.x), 3), collapse = ", "))
) |>
  knitr::kable(caption = "Variables in the `sta` (statin) cohort.")
```

## A note on the outcomes

As Chapter 2 mentioned, both cohorts are built from real NAMCS office
visits, but their outcomes are simulated: `incident_pud` in `ns`, and the
event-time variables (`cv_time`, `death_time`, `loss_fu_time`) that
`R/setup.R` combines into `event_time` and `cv_event` in `sta`. Nothing
about the covariates or the treatment assignment is invented; only the
outcome is, generated from a data-generating process known exactly to
whoever wrote the simulation. That is a deliberate choice for a teaching
book, since it means every chapter's estimate can in principle be checked
against a right answer, which is a luxury no real analysis has. We
analyze the cohorts here the way we would analyze real data, arguing for
each assumption rather than looking the truth up, although the truth is
there behind the file, which is what makes these two cohorts safe to
learn on.

# The estimator toolkit

Across the book we wrote one small function per chapter, each solving
the identification or estimation problem that chapter introduced, and
each built to be reused rather than thrown away. This table collects all
of them in one place, giving the name of each, the chapter that
introduced it, the arguments it takes, and what it returns. Every one of them lives
in its own file under `R/`, and is loaded with `source("R/<file>")` after
the `source("R/setup.R")` that opens every chapter — exactly as the
chapters do it. One dependency is worth naming: `iptw_estimator()` calls
`est_ps()`, so source `R/propensity.R` alongside
`R/iptw_estimator.R`.

```{r app-toolkit-table, echo = FALSE}
tribble(
  ~ name, ~ file, ~ chapter, ~ signature, ~ returns,
  "table1_continuous", "table1.R", "2",
    "table1_continuous(data, vars, by)",
    "Tibble of mean (SD) for each continuous variable, one column per level of `by`.",
  "table1_categorical", "table1.R", "2",
    "table1_categorical(data, vars, by)",
    "Tibble of n (%) for each level of each categorical variable, one column per level of `by`.",
  "standardize", "standardize.R", "7",
    "standardize(data, outcome_model, treat, family = binomial(), n_boot = 200, conf_level = 0.95, seed = NULL)",
    "Tibble with one row per estimand (risk0, risk1, rd, rr) giving the point estimate and a bootstrap confidence interval, from parametric g-formula standardization.",
  "est_ps", "propensity.R", "8",
    "est_ps(data, treat_model)",
    "The input data with a `ps` column added, holding fitted propensity scores from a logistic regression.",
  "ps_match", "propensity.R", "8",
    "ps_match(data, treat, ps = \"ps\", caliper = 0.05)",
    "Matched-pairs data frame from greedy 1:1 nearest-neighbor matching on `ps`, without replacement, with a `pair_id` column.",
  "iptw_estimator", "iptw_estimator.R", "9",
    "iptw_estimator(data, treat_model, outcome, treat, family = gaussian(), stabilize = TRUE, truncate_at = NULL)",
    "Tidy tibble (via `broom::tidy()`) of a weighted-GEE treatment-effect estimate with a robust confidence interval, from stabilized or unstabilized IPTW.",
  "plot_km", "km.R", "10",
    "plot_km(data, treat, time = \"event_time\", event = \"cv_event\", wt = NULL, labels = NULL)",
    "A ggplot object: cumulative incidence curves by treatment arm, with confidence bands, optionally weighted.",
  "make_person_period", "ipcw.R", "11",
    "make_person_period(data, time, event, cuts)",
    "Person-period expansion of one-row-per-patient survival data, via `survival::survSplit()`.",
  "ipcw_estimator", "ipcw.R", "11",
    "ipcw_estimator(pp_data, censor_model, outcome_formula, id = \".row_id\", numerator_model = NULL, truncate_at = 0.99)",
    "A list with `fit` (tidy weighted-GEE coefficients) and `data` (the person-period data with censoring weights attached), from IPCW.",
  "ccw_estimator", "ccw_estimator.R", "12",
    "ccw_estimator(pp_data, censor_model, treat, id = \".row_id\", numerator_model = NULL, truncate_at = 0.99)",
    "Tibble of hazard, survival, and cumulative risk by treatment arm and interval, from the clone-censor-weight algorithm.",
  "iv_estimator", "iv_estimator.R", "13",
    "iv_estimator(data, instrument, treat, outcome, covs = NULL, n_boot = 200, conf_level = 0.95, seed = NULL)",
    "Tibble with one row giving the two-stage-least-squares estimate and a bootstrap confidence interval.",
  "evalue_rr", "sensitivity.R", "14",
    "evalue_rr(rr, conf_low = NULL, conf_high = NULL)",
    "Tibble giving the E-value for the point estimate and, if a confidence interval is supplied, for the bound closest to the null.",
  "adjust_rr_unmeasured", "sensitivity.R", "14",
    "adjust_rr_unmeasured(rr_obs, p_u1, p_u0, rr_ud)",
    "A single numeric: the risk ratio that would remain after removing the bias contributed by a specified unmeasured binary confounder."
) |>
  knitr::kable(
    col.names = c("Function", "File", "Ch.", "Signature", "Returns"),
    caption = "Estimator and helper functions defined in R/, in the order they are introduced."
  )
```

# Generalized linear models for causal contrasts

Chapters 7 through 9 fit a regression and then read either a single
coefficient or a pair of predicted means as a causal contrast. Which
contrast the coefficient carries is settled by the link function of the
model, and those chapters chose links without stopping to explain the
family of models they come from. This appendix is that explanation.

A **generalized linear model** writes the conditional mean of the outcome
as a transformation of a linear predictor,

$$g\bigl(E[Y \mid X]\bigr) = \beta_0 + \beta_1 X_1 + \cdots + \beta_p X_p,$$

where $g$ is the **link function** and the **family** names the assumed
relationship between the mean and the variance. McCullagh and Nelder
assembled the class in this form. What matters for us is that the same
fitting machinery returns a different causal contrast for each link, with
the data and the covariates held fixed.

Three combinations do nearly all of the work in this book. The identity
link with the gaussian family regresses $Y$ on treatment directly, so its
coefficient is a difference of means, which for a binary outcome is a
**risk difference**. The log link with the Poisson family gives a
coefficient that exponentiates to a ratio of means, a **risk ratio** when
$Y$ is binary; the likelihood is then deliberately wrong, since a 0/1
outcome is not a count, although the mean model is correct and the
coefficient still estimates a log risk ratio. That is the **modified
Poisson** device Chapter 9 uses. The logit link with the binomial family
is ordinary logistic regression, and its coefficient exponentiates to an
**odds ratio**. All three fit with the same call to `glm()`, changed only
in `family =`.

```{r app-glm-links}
links <- list(
  "identity / gaussian" = gaussian(link = "identity"),
  "log / Poisson"       = poisson(link = "log"),
  "logit / binomial"    = binomial(link = "logit")
)

glm_tab <- imap_dfr(links, function(fam, label) {
  fit <- glm(pud ~ cox2_init, family = fam, data = ns)
  b   <- broom::tidy(fit) |> filter(term == "cox2_init")
  tibble(
    link        = label,
    contrast    = c("risk difference", "risk ratio", "odds ratio")[
                    match(label, names(links))],
    coefficient = b$estimate,
    value       = if (fam$link == "identity") b$estimate else exp(b$estimate)
  )
})

glm_tab |>
  knitr::kable(digits = 3,
               caption = "Three links, three crude contrasts, one cohort.")
```

```{r app-glm-values, include=FALSE}
gv <- function(lab, col) glm_tab[[col]][glm_tab$link == lab]
crude_rd <- gv("identity / gaussian", "value")
crude_rr <- gv("log / Poisson", "value")
crude_or <- gv("logit / binomial", "value")
```

Reading down the value column, Cox-2 initiation is associated with a
crude risk difference of `r sprintf("%+.1f", 100 * crude_rd)` percentage
points, a crude risk ratio of `r sprintf("%.2f", crude_rr)`, and a crude
odds ratio of `r sprintf("%.2f", crude_or)`. These are crude
associations: nothing has been adjusted, the two arms differ at baseline
in the ways Chapter 2 tabulated, and the outcome in `ns` is simulated.
Adjustment is the business of the body chapters, and each of them reaches
one of these three contrasts by a route that gives it a causal reading.
The odds ratio sits further from the null than the risk ratio, which is
the arithmetic of an uncommon outcome rather than a finding, and the
non-collapsibility Chapter 7 discusses is a further reason to prefer the
first two rows when the question is about risk.

Two of the three fits report a standard error we do not want. For the
gaussian fit on a binary outcome the variance depends on the mean rather
than being constant, and for the Poisson fit the assumed variance exceeds
the Bernoulli variance, so in both cases the model-based standard error
comes from a variance model we know to be false. A **robust**, or
sandwich, standard error estimates the variance of the estimating
equation from the data instead, which repairs the interval and leaves the
point estimate alone. Chapters 9 and 11 obtain it from
`geepack::geeglm()`, which also takes the weights those chapters need.

```{r app-glm-robust}
imap_dfr(links[1:2], function(fam, label) {
  m <- glm(pud ~ cox2_init, family = fam, data = ns)
  g <- geepack::geeglm(pud ~ cox2_init, id = seq_len(nrow(ns)),
                       family = fam, data = ns)
  tibble(link = label,
         se_model  = broom::tidy(m)$std.error[2],
         se_robust = broom::tidy(g)$std.error[2])
}) |>
  knitr::kable(digits = 4,
               caption = "Model-based and robust standard errors for the same coefficient.")
```

The two disagreements run in opposite directions, the robust error being
the larger one for the identity link and the smaller one for the log
link. Neither direction is guaranteed in general, which is the reason to
compute the robust error rather than to reason about which way a
misspecified variance will err.

In the course, session 8 takes up generalized linear models before the
propensity-score sessions that follow, and every body chapter from
Chapter 7 onward assumes that material. This appendix is where a reader
who wants it written down once can find it.

# A bridge to Cox regression

This book reports risks at named horizons and treats a hazard ratio,
where one appears at all, as a modeling device on the way to a risk
curve. Much of the literature a student will read reports the hazard
ratio itself, from a Cox model, as the result. Being able to read those
papers is a separate skill from being able to defend an estimand, and
this appendix supplies it.

The **Cox proportional hazards model** writes the hazard at time $t$ for
a patient with covariates $X$ as

$$h(t \mid X) = h_0(t)\exp(\beta_1 X_1 + \cdots + \beta_p X_p),$$

leaving the baseline hazard $h_0(t)$ entirely unspecified and estimating
$\beta$ from a partial likelihood that compares, at each observed event
time, the patient who had the event against the risk set she was drawn
from. Because $h_0(t)$ cancels out of that comparison, the model needs no
assumption about the shape of the hazard over time, which is most of its
appeal. What it does assume is that the ratio $\exp(\beta)$ — the
**hazard ratio** — is the same at every follow-up time. Fitting one on
`sta` takes a single call.

```{r app-cox-crude}
library(survival)

cox_crude <- coxph(Surv(event_time, cv_event) ~ statin, data = sta)
broom::tidy(cox_crude, exponentiate = TRUE, conf.int = TRUE) |>
  knitr::kable(digits = 3)
```

That contrast is crude. To adjust it we reuse the machinery of Chapters 8
and 9 without change: fit the treatment model, build stabilized weights
from the propensity score, and pass them to `coxph()` through its
`weights` argument, with `robust = TRUE` so the variance accounts for the
weighted rows no longer being independent draws.

```{r app-cox-iptw}
source("R/propensity.R")

sta_w <- est_ps(sta, reformulate(sta_covs, response = "statin")) |>
  mutate(p_treat = mean(statin),
         wt = if_else(statin == 1, p_treat / ps, (1 - p_treat) / (1 - ps)))

cox_ipw <- coxph(Surv(event_time, cv_event) ~ statin, data = sta_w,
                 weights = wt, robust = TRUE)
broom::tidy(cox_ipw, exponentiate = TRUE, conf.int = TRUE) |>
  knitr::kable(digits = 3)
```

```{r app-cox-values, include=FALSE}
cx <- function(fit) broom::tidy(fit, exponentiate = TRUE, conf.int = TRUE)
hr_crude <- cx(cox_crude)$estimate[1]
hr_w     <- cx(cox_ipw)$estimate[1]
hr_w_lo  <- cx(cox_ipw)$conf.low[1]
hr_w_hi  <- cx(cox_ipw)$conf.high[1]
```

The crude hazard ratio is `r sprintf("%.2f", hr_crude)`; weighted, it
becomes `r sprintf("%.2f", hr_w)` (`r sprintf("%.2f", hr_w_lo)` to
`r sprintf("%.2f", hr_w_hi)`). Adjustment moves the estimate further from
the null, which is a reminder that the arms in `sta` differ at baseline in
ways that bear on prognosis. The outcomes here are simulated, so neither
number is evidence about statins.

Two cautions carry over from the body of the book. The first is Chapter
10's argument, following Hernán (2010), that a hazard ratio beyond the
first instant of follow-up is conditional on having survived, and so
carries a built-in selection bias that no baseline adjustment removes;
we do not restate it here beyond noting that it is the reason this book
reports risks at five and `r admin_end` years instead. The second is that
proportional hazards is a substantive claim about the cohort rather than a
technical detail. Chapter 10 fit the same model in two periods and got
two different answers, and when the assumption fails, the single reported
ratio is a weighted average over periods that disagree, with weights
determined by when the events happened to fall.

Cox regression appears in the course in a second role as well. Lecture 15
uses it, as much of the literature does, to model censoring as well as the
outcome, estimating each patient's hazard of dropping out and converting
it into a survival probability. Chapter 11 does the same work in discrete
time: the pooled logistic model it fits on person-period data approximates
a proportional hazards model when the interval hazards are small, and it
has the practical advantage that the weights and the outcome model are
fit by the same machinery.

# Course map: PHS 921

This book accompanies PHS 921, and the two are organized on their own
terms: the course moves in the order that works for a semester of
lectures, the book in the order that works for a reader building one
idea on the last. This table lines the two up, session by session, so
that a student can find the chapter behind any lecture and a reader can
see which lecture a chapter was written for. The lab meets separately
and works through the code, so the R material appears here only where a
session is built around it.

| Session | Topic | Where in the book |
|---|---|---|
| 1 | Introduction | Chapter 1 |
| 2–3 | Intro to R workshop | Chapter 2, Appendix A |
| 4 | RCTs and experiments | Chapter 4 |
| 5 | Counterfactuals | Chapter 3 |
| 6 | Study design and immortal time bias | Chapter 6 |
| 7 | Real-world data sources | Chapter 2 (the two cohorts), Appendix B |
| 8 | Generalized linear models | Appendix D; used throughout Chapters 7–9 |
| 9–11 | Propensity scores and IPTW | Chapters 8–9 |
| 12 | G-computation | Chapter 7 |
| 13–14 | Variable selection and DAGs (+ workshop) | Chapter 5; Chapter 8 (choosing the propensity-score model) |
| 15 | Censored data and Kaplan–Meier | Chapter 10 |
| 17 | Cox regression, IPCW, competing risks | Chapter 11; Chapter 10 (death as a competing risk); Appendix E (Cox regression) |
| 18 | IPCW workshop | Chapter 11 |
| 19–20 | Artificial censoring; clone-censor-weight | Chapter 12 |
| 22 | Natural experiments and instrumental variables | Chapter 13 |
| 23 | Negative control outcomes | Chapter 14 (negative control outcomes) |
| 24 | External control arms | Chapter 4 (target trials, and the external control arm paragraph that closes that section) |
| 25–26 | Protocol presentations | Chapter 4's protocol table with Chapter 14's reporting checklist |

Two places in the table run backwards relative to the page order. The
book introduces causal diagrams (Chapter 5) before study design
(Chapter 6), and g-computation (Chapter 7) before propensity scores
(Chapters 8–9), because each earlier chapter supplies vocabulary the
later one uses. The course takes both pairs in the other order. Where
the two disagree, follow the map rather than the page numbers.

Two sessions have no counterpart here. Session 16 is the midterm, and
session 21 takes up missing data, which this book does not treat: the two
cohorts are complete on the variables we use, and the assumptions and
imputation machinery that incomplete data require are a subject of their
own rather than a section of a causal-inference chapter. Students should
take that session from the lecture notes and the reading assigned with
them.

# Further reading

This book is deliberately narrow, with two running examples, one
estimator per chapter, and just enough theory to justify each one. The
sources below go further, in a number of directions.

**On the framework as a whole.** Miguel Hernán and James Robins, *Causal
Inference: What If* (Chapman & Hall/CRC) covers the same potential-outcomes framework at greater depth and rigor, with a fuller treatment of time-varying treatments, g-methods, and the assumptions each identification strategy needs. It is available free online and is the natural next stop after this course.

**On where the framework came from.** Two short papers repay careful reading.

- Holland (1986, JASA) — statistics and causal inference: the paper that
  set the two side by side and fixed much of the potential-outcomes
  vocabulary Chapter 3 uses.
- Hernán (2004, JECH) — a definition of a causal effect, written for
  epidemiologists, insisting that we say what an effect is a statement
  about before choosing anything to estimate it with.

**On causal diagrams.** Chapter 5 uses DAGs as a working tool; these two
papers are where the tool and its most useful application were set out
for epidemiology.

- Greenland, Pearl, and Robins (1999, Epidemiology) — causal diagrams for
  epidemiologic research: the paper that introduced DAGs to the field,
  with the backdoor reasoning Chapter 5 runs on.
- Hernán, Hernández-Díaz, and Robins (2004, Epidemiology) — a structural
  approach to selection bias, which recasts a long list of named biases
  as conditioning on a collider. The source for Chapter 5's selection
  bias section, and the reason Chapters 10 and 11 can treat censoring and
  survivorship as the same problem.

**On epidemiologic method more broadly.** Kenneth Rothman, Sander
Greenland, and Timothy Lash, *Modern Epidemiology* (Lippincott Williams
& Wilkins), is the standard reference for study design, bias, and
measurement issues that sit upstream of any causal estimator — how a
cohort or case-control study is assembled, what confounding and
selection bias look like in practice, and how the field has thought
about validity before "causal inference" was its own subfield.

**On the target trial.** The framework Chapter 4 adopts and Chapters 6
and 12 keep returning to.

- Hernán and Robins (2016, AJE) — using big data to emulate a target
  trial when a randomized trial is not available: the paper that named
  the framework and the source of Chapter 4's protocol table.
- Hernán et al. (2016, Journal of Clinical Epidemiology) — specifying a
  target trial prevents immortal time bias, which is the argument behind
  Chapter 6's insistence that eligibility, assignment, and start of
  follow-up share a date.

**On design in healthcare databases.** Both papers bear directly on
Chapter 6.

- Ray (2003, AJE) — the case for new-user designs, and what a cohort of
  prevalent users quietly does to a comparison.
- Brookhart (2015, AJE) — the treatment decision design: the general
  form of the new-user idea, anchoring comparisons at any clinical
  decision point, and useful background for target trial emulation.
- Schneeweiss et al. (2019, Annals of Internal Medicine) — a graphical
  convention for depicting longitudinal study designs, which makes an
  argument about eligibility and time zero reviewable at a glance.

**On standardization and the g-formula.** The machinery of Chapter 7,
and why a regression coefficient is not the estimand.

- Robins (1986, Mathematical Modelling) — the paper that introduced the
  g-formula, of which Chapter 7's predict-then-average recipe is the
  simplest case.
- Greenland, Robins, and Pearl (1999, Statistical Science) — confounding
  and collapsibility in causal inference: the reference for why a
  conditional and a marginal odds ratio differ even with no confounding
  at all.

**On confounding control and propensity scores.** These take up the
questions Chapter 8 raises and then sets aside.

- Rubin (2008, Annals of Applied Statistics) — for objective causal
  inference, design trumps analysis: the argument behind Chapter 8's
  claim that building the score without looking at the outcome is most
  of what the score is for.
- Brookhart et al. (2006, AJE) — variable selection for propensity score
  models: which covariates belong in a treatment model, and what
  including or excluding one does to bias and variance.
- Brookhart et al. (2010, Medical Care) — a survey of the options for
  confounding control in healthcare database studies, of which this book
  works through a few in detail.
- Schneeweiss et al. (2009, Epidemiology) — high-dimensional propensity
  score adjustment: the automated proxy-selection algorithm Chapter 8
  mentions but does not implement.
- Lipsitch, Tchetgen Tchetgen, and Cohen (2010, Epidemiology) — negative
  controls, and what a non-null association with an outcome the treatment
  cannot cause does and does not establish. The source for Chapter 14's
  section on them.

**On weighting and marginal structural models.** Chapter 9's estimator,
in the papers that established it.

- Robins, Hernán, and Brumback (2000, Epidemiology) — marginal structural
  models and causal inference in epidemiology: the source of the model
  class and of the name Chapter 9 explains.
- Cole and Hernán (2008, AJE) — constructing inverse probability weights
  for marginal structural models, which is the practical companion to
  Chapter 9's stabilization and truncation.
- Kurth et al. (2006, AJE) — results of multivariable logistic regression,
  propensity matching, propensity adjustment, and propensity-based
  weighting under conditions of nonuniform effect, in which IPTW and SMR
  weighting of the same t-PA cohort gave sharply different answers. The
  source for Chapter 9's fourth exercise, and the clearest published
  illustration of why the choice between the ATE and the ATT matters when
  the effect is heterogeneous.

**On what to report.** All three argue for a particular quantity rather
than a particular estimator.

- Cole et al. (2015, AJE) — risk: a case for reporting risks and risk
  contrasts rather than ratios of hazards, which is why Chapters 10
  through 12 keep returning to cumulative incidence.
- Hernán (2010, Epidemiology) — the hazards of hazard ratios: the short
  editorial behind Chapter 10's built-in selection argument, and the
  reason this book treats a hazard ratio as a modeling device rather than
  an estimand.
- Hernán and Robins (2017, NEJM) — per-protocol analyses of pragmatic
  trials, the argument behind Chapter 4's per-protocol discussion and
  Chapter 12's censoring and weighting.

**On instrumental variables.** Chapter 13 leans on all three.

- Angrist, Imbens, and Rubin (1996, JASA) — identification of causal
  effects using instrumental variables, and the paper that made the
  complier-average interpretation standard.
- Hernán and Robins (2006, Epidemiology) — instruments for causal
  inference: an epidemiologist's dream? The source of Chapter 13's
  statement of the three conditions, and of its skepticism about them.
- Brookhart et al. (2006, Epidemiology) — preference-based instrumental
  variables: what makes a prescriber's preference a defensible
  instrument, and where that assumption tends to break.

**On sensitivity analysis.** The two devices Chapter 14 builds, each in
the paper that introduced it.

- Schneeweiss (2006, Pharmacoepidemiology and Drug Safety) — sensitivity
  analysis and external adjustment for unmeasured confounders, the source
  of Chapter 14's term and of the bias-factor parameterization it uses.
- VanderWeele and Ding (2017, Annals of Internal Medicine) — sensitivity
  analysis in observational research: introducing the E-value, and the
  derivation Chapter 14's second device rests on.

**On the software.** Yihui Xie, J.J. Allaire, and Garrett Grolemund,
*R Markdown: The Definitive Guide* and the `bookdown` documentation at
[bookdown.org](https://bookdown.org/), cover the publishing system this
book itself is written in. Hadley Wickham, Mine Çetinkaya-Rundel, and
Garrett Grolemund, *R for Data Science* ("R4DS", also free online),
is the best general reference for the tidyverse functions — `dplyr`
verbs, `ggplot2`, `purrr` — that Chapter 2 introduces only as much of as
this book needs.
