# (APPENDIX) Appendices {-}

# R and RStudio setup

This appendix collects the one-time setup we assumed you had already
done, back in Chapter 2. If everything already works, meaning that you
can knit a chapter and see `ns` and `sta` load without complaint, you can
skip past it.

## Installing R and RStudio

Install R first, then RStudio. Both are free.

- **R**: download and install the current release from
  [CRAN](https://cran.r-project.org/), the Comprehensive R Archive
  Network. Pick the installer for your operating system and accept the
  defaults.
- **RStudio**: download and install RStudio Desktop from
  [posit.co/download/rstudio-desktop](https://posit.co/download/rstudio-desktop/).
  RStudio is an editor for R, not a replacement for it, so R must already
  be installed before RStudio is any use.

Open RStudio once, just to confirm it starts and finds your R
installation (it will say so in the Console pane on startup).

## Installing the packages this book uses

Every chapter in this book leans on the `tidyverse` for data manipulation
and plotting, and most chapters bring in one or two special-purpose
packages on top of that. Install all of them at once by pasting the
following into the R console and running it:

```{r app-install-packages, eval = FALSE}
install.packages(c(
  "tidyverse",  # dplyr, ggplot2, purrr, and friends
  "bookdown",   # builds this book; only needed if you render it yourself
  "dagitty",    # causal diagrams, Chapter 5
  "ggdag",      # plotting causal diagrams, Chapter 5 and later chapters
  "survival",   # Surv(), survfit(), survSplit(), Chapters 10-12
  "cobalt",     # covariate-balance tables and plots, Chapter 8
  "geepack",    # weighted GEE, the engine behind iptw_estimator() and
                # ipcw_estimator(), Chapters 9 and 11
  "broom",      # tidy() for turning model objects into tibbles
  "remotes"     # installs packages from GitHub, including NAMCS below
))
```

One package used in Chapters 11 and 12, `splines`, ships with base R
and never needs installing.

## Installing the NAMCS teaching package

The two cohorts this book uses throughout, `ns` and `sta`, are not on
CRAN. They live in a companion package, `NAMCS`, distributed from
GitHub. With `remotes` installed from the previous step, run:

```{r app-install-namcs, eval = FALSE}
remotes::install_github("alanbrookhart/NAMCS")
```

If that command fails, it is almost always one of two things: `remotes`
is not installed (revisit the previous step), or GitHub is unreachable
from your network (try again from a different network, or ask your
instructor for the data files directly).

## Loading everything a chapter needs

Every chapter and appendix in this book opens with a chunk that runs

```{r app-setup-source, eval = FALSE}
source("R/setup.R")
```

`R/setup.R` loads the tidyverse and the `NAMCS` package, loads `ns` and
`sta`, and derives the handful of analysis columns and covariate lists
(`ns_covs`, `sta_covs`, and so on) that later chapters take for granted.
Sourcing it is the whole of what you need to do to follow along with any
chapter's code on its own, without having run the earlier chapters
first.

```{r setup-app, include = FALSE}
source("R/setup.R")
```

# NAMCS codebook

Both teaching cohorts carry many more columns than the ones we actually
use in adjustment models (`ns_covs` and `sta_covs` in `R/setup.R` are
short lists curated from a much wider survey). This appendix is the
reference for the rest: a full variable-by-variable codebook, generated
directly from the data rather than typed out by hand, so it will always
match whatever is actually in the installed `NAMCS` package.

## The NSAID cohort (`ns`)

```{r app-nsaid-codebook}
tibble(
  variable = names(ns),
  type = map_chr(ns, ~ class(.x)[1]),
  example_values = map_chr(ns, ~ paste(head(unique(.x), 3), collapse = ", "))
) |>
  knitr::kable(caption = "Variables in the `ns` (NSAID) cohort.")
```

## The statin cohort (`sta`)

```{r app-statin-codebook}
tibble(
  variable = names(sta),
  type = map_chr(sta, ~ class(.x)[1]),
  example_values = map_chr(sta, ~ paste(head(unique(.x), 3), collapse = ", "))
) |>
  knitr::kable(caption = "Variables in the `sta` (statin) cohort.")
```

## A note on the outcomes

As Chapter 2 mentioned, both cohorts are built from real NAMCS office
visits, but their outcomes are simulated: `incident_pud` in `ns`, and the
event-time variables (`cv_time`, `death_time`, `loss_fu_time`) that
`R/setup.R` combines into `event_time` and `cv_event` in `sta`. Nothing
about the covariates or the treatment assignment is invented; only the
outcome is, generated from a data-generating process known exactly to
whoever wrote the simulation. That is a deliberate choice for a teaching
book, since it means every chapter's estimate can in principle be checked
against a right answer, which is a luxury no real analysis has. We
analyze the cohorts here the way we would analyze real data, arguing for
each assumption rather than looking the truth up, although the truth is
there behind the file, which is what makes these two cohorts safe to
learn on.

# The estimator toolkit

Across the book we wrote one small function per chapter, each solving
the identification or estimation problem that chapter introduced, and
each built to be reused rather than thrown away. This table collects all
of them in one place, giving the name of each, the chapter that
introduced it, what you hand it, and what it returns. Every one of them lives
in its own file under `R/`, and is loaded with `source("R/<file>")` after
the `source("R/setup.R")` that opens every chapter — exactly as the
chapters do it. One dependency is worth naming: `iptw_estimator()` calls
`est_ps()`, so source `R/propensity.R` alongside
`R/iptw_estimator.R`.

```{r app-toolkit-table, echo = FALSE}
tribble(
  ~ name, ~ file, ~ chapter, ~ signature, ~ returns,
  "table1_continuous", "table1.R", "2",
    "table1_continuous(data, vars, by)",
    "Tibble of mean (SD) for each continuous variable, one column per level of `by`.",
  "table1_categorical", "table1.R", "2",
    "table1_categorical(data, vars, by)",
    "Tibble of n (%) for each level of each categorical variable, one column per level of `by`.",
  "standardize", "standardize.R", "7",
    "standardize(data, outcome_model, treat, family = binomial(), n_boot = 200, conf_level = 0.95, seed = NULL)",
    "Tibble with one row per estimand (risk0, risk1, rd, rr) giving the point estimate and a bootstrap confidence interval, from parametric g-formula standardization.",
  "est_ps", "propensity.R", "8",
    "est_ps(data, treat_model)",
    "The input data with a `ps` column added, holding fitted propensity scores from a logistic regression.",
  "ps_match", "propensity.R", "8",
    "ps_match(data, treat, ps = \"ps\", caliper = 0.05)",
    "Matched-pairs data frame from greedy 1:1 nearest-neighbor matching on `ps`, without replacement, with a `pair_id` column.",
  "iptw_estimator", "iptw_estimator.R", "9",
    "iptw_estimator(data, treat_model, outcome, treat, family = gaussian(), stabilize = TRUE, truncate_at = NULL)",
    "Tidy tibble (via `broom::tidy()`) of a weighted-GEE treatment-effect estimate with a robust confidence interval, from stabilized or unstabilized IPTW.",
  "plot_km", "km.R", "10",
    "plot_km(data, treat, time = \"event_time\", event = \"cv_event\", wt = NULL, labels = NULL)",
    "A ggplot object: cumulative incidence curves by treatment arm, with confidence bands, optionally weighted.",
  "make_person_period", "ipcw.R", "11",
    "make_person_period(data, time, event, cuts)",
    "Person-period expansion of one-row-per-patient survival data, via `survival::survSplit()`.",
  "ipcw_estimator", "ipcw.R", "11",
    "ipcw_estimator(pp_data, censor_model, outcome_formula, id = \".row_id\", numerator_model = NULL, truncate_at = 0.99)",
    "A list with `fit` (tidy weighted-GEE coefficients) and `data` (the person-period data with censoring weights attached), from IPCW.",
  "ccw_estimator", "ccw_estimator.R", "12",
    "ccw_estimator(pp_data, censor_model, treat, id = \".row_id\", numerator_model = NULL, truncate_at = 0.99)",
    "Tibble of hazard, survival, and cumulative risk by treatment arm and interval, from the clone-censor-weight algorithm.",
  "iv_estimator", "iv_estimator.R", "13",
    "iv_estimator(data, instrument, treat, outcome, covs = NULL, n_boot = 200, conf_level = 0.95, seed = NULL)",
    "Tibble with one row giving the two-stage-least-squares estimate and a bootstrap confidence interval.",
  "evalue_rr", "sensitivity.R", "14",
    "evalue_rr(rr, conf_low = NULL, conf_high = NULL)",
    "Tibble giving the E-value for the point estimate and, if a confidence interval is supplied, for the bound closest to the null.",
  "adjust_rr_unmeasured", "sensitivity.R", "14",
    "adjust_rr_unmeasured(rr_obs, p_u1, p_u0, rr_ud)",
    "A single numeric: the risk ratio that would remain after removing the bias contributed by a specified unmeasured binary confounder."
) |>
  knitr::kable(
    col.names = c("Function", "File", "Ch.", "Signature", "Returns"),
    caption = "Estimator and helper functions defined in R/, in the order they are introduced."
  )
```

# Course map: PHS 921

This book accompanies PHS 921, and the two are organized on their own
terms: the course moves in the order that works for a semester of
lectures, the book in the order that works for a reader building one
idea on the last. This table lines the two up, session by session, so
that a student can find the chapter behind any lecture and a reader can
see which lecture a chapter was written for. The lab meets separately
and works through the code, so the R material appears here only where a
session is built around it.

| Session | Topic | Where in the book |
|---|---|---|
| 1 | Introduction | Chapter 1 |
| 2–3 | Intro to R workshop | Chapter 2, Appendix A |
| 4 | RCTs and experiments | Chapter 4 |
| 5 | Counterfactuals | Chapter 3 |
| 6 | Study design and immortal time bias | Chapter 6 |
| 7 | Real-world data sources | Chapter 2 (the two cohorts), Appendix B |
| 8 | Generalized linear models | used throughout Chapters 7–9 |
| 9–11 | Propensity scores and IPTW | Chapters 8–9 |
| 12 | G-computation | Chapter 7 |
| 13–14 | Variable selection and DAGs (+ workshop) | Chapter 5; Chapter 8 (choosing the propensity-score model) |
| 15 | Censored data and Kaplan–Meier | Chapter 10 |
| 17 | Cox regression, IPCW, competing risks | Chapter 11; Chapter 10 (death as a competing risk) |
| 18 | IPCW workshop | Chapter 11 |
| 19–20 | Artificial censoring; clone-censor-weight | Chapter 12 |
| 22 | Natural experiments and instrumental variables | Chapter 13 |
| 23 | Negative control outcomes | Chapter 14 (negative control outcomes) |
| 24 | External control arms | Chapter 4 (target trials, for context) |
| 25–26 | Protocol presentations | Chapter 4's protocol table with Chapter 14's reporting checklist |

Two places in the table run backwards relative to the page order. The
book introduces causal diagrams (Chapter 5) before study design
(Chapter 6), and g-computation (Chapter 7) before propensity scores
(Chapters 8–9), because each earlier chapter supplies vocabulary the
later one uses. The course takes both pairs in the other order. Where
the two disagree, follow the map rather than the page numbers.

# Further reading

This book is deliberately narrow, with two running examples, one
estimator per chapter, and just enough theory to justify each one. The
sources below go further, in a number of directions.

**On the framework as a whole.** Miguel Hernán and James Robins, *Causal
Inference: What If* (Chapman & Hall/CRC) covers the same potential-outcomes framework at greater depth and rigor, with a fuller treatment of time-varying treatments, g-methods, and the assumptions each identification strategy needs. It is available free online and is the natural next stop after this course.

**On where the framework came from.** Two short papers are well worth a careful read.

- Holland (1986, JASA) — statistics and causal inference: the paper that
  set the two side by side and fixed much of the potential-outcomes
  vocabulary Chapter 3 uses.
- Hernán (2004, JECH) — a definition of a causal effect, written for
  epidemiologists, insisting that we say what an effect is a statement
  about before choosing anything to estimate it with.

**On causal diagrams.** Chapter 5 uses DAGs as a working tool; these two
papers are where the tool and its most useful application were set out
for epidemiology.

- Greenland, Pearl, and Robins (1999, Epidemiology) — causal diagrams for
  epidemiologic research: the paper that introduced DAGs to the field,
  with the backdoor reasoning Chapter 5 runs on.
- Hernán, Hernández-Díaz, and Robins (2004, Epidemiology) — a structural
  approach to selection bias, which recasts a long list of named biases
  as conditioning on a collider. The source for Chapter 5's selection
  bias section, and the reason Chapters 10 and 11 can treat censoring and
  survivorship as the same problem.

**On epidemiologic method more broadly.** Kenneth Rothman, Sander
Greenland, and Timothy Lash, *Modern Epidemiology* (Lippincott Williams
& Wilkins), is the standard reference for study design, bias, and
measurement issues that sit upstream of any causal estimator — how a
cohort or case-control study is assembled, what confounding and
selection bias look like in practice, and how the field has thought
about validity before "causal inference" was its own subfield.

**On the target trial.** The framework Chapter 4 adopts and Chapters 6
and 12 keep returning to.

- Hernán and Robins (2016, AJE) — using big data to emulate a target
  trial when a randomized trial is not available: the paper that named
  the framework and the source of Chapter 4's protocol table.
- Hernán et al. (2016, Journal of Clinical Epidemiology) — specifying a
  target trial prevents immortal time bias, which is the argument behind
  Chapter 6's insistence that eligibility, assignment, and start of
  follow-up share a date.

**On design in healthcare databases.** Both papers bear directly on
Chapter 6.

- Ray (2003, AJE) — the case for new-user designs, and what a cohort of
  prevalent users quietly does to a comparison.
- Brookhart (2015, AJE) — the treatment decision design: the general
  form of the new-user idea, anchoring comparisons at any clinical
  decision point, and useful background for target trial emulation.
- Schneeweiss et al. (2019, Annals of Internal Medicine) — a graphical
  convention for depicting longitudinal study designs, which makes an
  argument about eligibility and time zero reviewable at a glance.

**On standardization and the g-formula.** The machinery of Chapter 7,
and why a regression coefficient is not the estimand.

- Robins (1986, Mathematical Modelling) — the paper that introduced the
  g-formula, of which Chapter 7's predict-then-average recipe is the
  simplest case.
- Greenland, Robins, and Pearl (1999, Statistical Science) — confounding
  and collapsibility in causal inference: the reference for why a
  conditional and a marginal odds ratio differ even with no confounding
  at all.

**On confounding control and propensity scores.** These take up the
questions Chapter 8 raises and then sets aside.

- Rubin (2008, Annals of Applied Statistics) — for objective causal
  inference, design trumps analysis: the argument behind Chapter 8's
  claim that building the score without looking at the outcome is most
  of what the score is for.
- Brookhart et al. (2006, AJE) — variable selection for propensity score
  models: which covariates belong in a treatment model, and what
  including or excluding one does to bias and variance.
- Brookhart et al. (2010, Medical Care) — a survey of the options for
  confounding control in healthcare database studies, of which this book
  works through a few in detail.
- Schneeweiss et al. (2009, Epidemiology) — high-dimensional propensity
  score adjustment: the automated proxy-selection algorithm Chapter 8
  mentions but does not implement.
- Lipsitch, Tchetgen Tchetgen, and Cohen (2010, Epidemiology) — negative
  controls, and what a non-null association with an outcome the treatment
  cannot cause does and does not establish. The source for Chapter 14's
  section on them.

**On weighting and marginal structural models.** Chapter 9's estimator,
in the papers that established it.

- Robins, Hernán, and Brumback (2000, Epidemiology) — marginal structural
  models and causal inference in epidemiology: the source of the model
  class and of the name Chapter 9 explains.
- Cole and Hernán (2008, AJE) — constructing inverse probability weights
  for marginal structural models, which is the practical companion to
  Chapter 9's stabilization and truncation.

**On what to report.** All three argue for a particular quantity rather
than a particular estimator.

- Cole et al. (2015, AJE) — risk: a case for reporting risks and risk
  contrasts rather than ratios of hazards, which is why Chapters 10
  through 12 keep returning to cumulative incidence.
- Hernán (2010, Epidemiology) — the hazards of hazard ratios: the short
  editorial behind Chapter 10's built-in selection argument, and the
  reason this book treats a hazard ratio as a modeling device rather than
  an estimand.
- Hernán and Robins (2017, NEJM) — per-protocol analyses of pragmatic
  trials, the argument behind Chapter 4's per-protocol discussion and
  Chapter 12's censoring and weighting.

**On instrumental variables.** Chapter 13 leans on all three.

- Angrist, Imbens, and Rubin (1996, JASA) — identification of causal
  effects using instrumental variables, and the paper that made the
  complier-average interpretation standard.
- Hernán and Robins (2006, Epidemiology) — instruments for causal
  inference: an epidemiologist's dream? The source of Chapter 13's
  statement of the three conditions, and of its skepticism about them.
- Brookhart et al. (2006, Epidemiology) — preference-based instrumental
  variables: what makes a prescriber's preference a defensible
  instrument, and where that assumption tends to break.

**On sensitivity analysis.** The two devices Chapter 14 builds, each in
the paper that introduced it.

- Schneeweiss (2006, Pharmacoepidemiology and Drug Safety) — sensitivity
  analysis and external adjustment for unmeasured confounders, the source
  of Chapter 14's term and of the bias-factor parameterization it uses.
- VanderWeele and Ding (2017, Annals of Internal Medicine) — sensitivity
  analysis in observational research: introducing the E-value, and the
  derivation Chapter 14's second device rests on.

**On the software.** Yihui Xie, J.J. Allaire, and Garrett Grolemund,
*R Markdown: The Definitive Guide* and the `bookdown` documentation at
[bookdown.org](https://bookdown.org/), cover the publishing system this
book itself is written in. Hadley Wickham, Mine Çetinkaya-Rundel, and
Garrett Grolemund, *R for Data Science* ("R4DS", also free online),
is the best general reference for the tidyverse functions — `dplyr`
verbs, `ggplot2`, `purrr` — that Chapter 2 introduces only as much of as
this book needs.
